---
title: "Importing and exporting data"
author: "Lars Relund Nielsen"
output:
  xaringan::moon_reader:
    css: "./libs/slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, child = "setup.Rmd", echo=FALSE}
```

```{r include=FALSE}
# use xaringan::inf_mr(cast_from = ".") to preview
```

layout: true

<div class="my-footer">
<span>
<a href="https://bss-osca.github.io/tfa/mod-r-io" target="_blank">Book</a> | <a href="https://bss-osca.github.io/tfa/slides/07-02_r-io-slides.html" target="_blank">Slides</a> | <a href="https://github.com/bss-osca/tfa/blob/master/slides/07-02_r-io-slides.Rmd" target="_blank">Source</a> | <a href="https://rstudio.cloud/spaces/176810/project/3026754" target="_blank">R project</a>
</span>
</div> 

---

## Input and output

  * For doing data driven analytics you first must import some data. That is, take data from a database, file, web API etc. and transform it into a data frame/table. 
  * Two main types of data files. Information is either binary encoded or stored as text files. 
  * Text files can be read by humans and any programming language. 
  * Binary file formats cannot be read by humans but allows space-efficient data compression. Moreover they can be used to save non tabular data.
  * CSV files contain delimiter separated values (e.g. `,`, `;` or `_`) in plain text and are often named using the file suffix `.csv`. 
  * The [readxl][tidyverse-readxl] package can be used to read Excel files. Write to an Excel file  using the [openxlsx][pkg-openxlsx] package. 
  * You can import and export to Google sheets using the [googlesheets4][tidyverse-googlesheets4] package. 
  * Json is an open standard text file format, and data interchange format. It can be used to store non tabular data in text format. 

---

## Examples and code 

Let us have a look at an example in the [RStudio Cloud project][r-cloud-mod16]. 

We will use the file `lecture/slides-examples.Rmd` throughout this lecture. 



```{r copy, include=FALSE}
file.copy(list.files(pattern = ".html"), "../docs/slides/", overwrite = T)
file.copy("./slides.css", "./libs/", overwrite = T)
file.copy("libs", "../docs/slides/", recursive = T)
file.copy("img", "../docs/slides/", recursive = T)
system2("Rscript", args = "-e 'rmarkdown::render(\"index.Rmd\", quiet = TRUE)'")
```

```{r links, child="links.md", include=FALSE}
```