---
title: "Fitting probability distributions"
author: "Lars Relund Nielsen"
output:
  xaringan::moon_reader:
    css: "./libs/slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, child = "setup.Rmd", echo=FALSE}
```

```{r include=FALSE, eval=FALSE}
# to preview
xaringan::inf_mr(cast_from = ".", quiet = T)
```

layout: true

```{r, echo=FALSE}
module_name <- "r-dist-fit"
module_number_prefix <- "15"
module_number <- as.numeric(module_number_prefix)
add_footer(module_number_prefix, module_name, r_cloud_id = 6689209)
knitr::opts_chunk$set(fig.path=str_c("img/", module_name, "-"))
```  

---

## Fitting distributions

- Fitting distributions to data is a very common task in statistics.
- Consists in choosing a probability distribution modelling the random variable, as well as finding parameter estimates for that distribution. 
- Requires judgment and expertise and generally needs an iterative process of distribution choice, parameter estimation, and quality of fit assessment. 
- You often have to fit distributions when you write your master thesis.
- We consider univariate data, i.e. univariate distributions such as normal, uniform, Poisson and exponential distributions. 
- Use  the [`fitdistrplus`](https://lbbe.univ-lyon1.fr/fr/fitdistrplus) package for fitting distributions.


---

## Steps

  * Fitting a univariate distribution requires judgment and expertise and generally needs an iterative process of distribution choice, parameter estimation, and quality of fit assessment. Steps may be:
    - Examine data a decide on discrete vs continuous distribution.
    - Find a set of candidate distributions.
    - Fit the distributions using statistical methods and consider various plots.
    - Decide on a distribution

---

## Continuous data

- The empirical plots of the density and the CDF may give you a hit about the distribution.
- The the skewness and kurtosis, linked to the third and fourth moments, are useful to identify candidates. 
- A non-zero skewness reveals a lack of symmetry of the empirical distribution.
- The kurtosis value quantifies the weight of tails in comparison to the normal distribution for which the kurtosis equals 3.

---

## Discrete data

- Try to fit different discrete distributions.
- Plot the CDF and the density.

---

# Examples

Let us consider the R markdown document at RStudio Cloud.




```{r copy, include=FALSE}
file.copy(list.files(pattern = ".html"), "../docs/slides/", overwrite = T)
file.copy("./slides.css", "./libs/", overwrite = T)
file.copy("libs", "../docs/slides/", recursive = T)
file.copy("img", "../docs/slides/", recursive = T)
system2("Rscript", args = "-e 'rmarkdown::render(\"index.Rmd\", quiet = TRUE)'")
```

```{r links, child="links.md", include=FALSE}
```
