---
title: "Procedures"
author: "Lars Relund Nielsen"
output:
  xaringan::moon_reader:
    css: "./libs/slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, child = "setup.Rmd", echo=FALSE}
```

```{r include=FALSE}
# use xaringan::inf_mr(cast_from = ".") to preview
```

layout: true

<div class="my-footer">
<span>
<a href="https://bss-osca.github.io/tfa/mod-procedures.html" target="_blank">Book</a> | <a href="https://bss-osca.github.io/tfa/slides/02-02_vba-procedures-slides.html" target="_blank">Slides</a> | <a href="https://github.com/bss-osca/tfa/blob/master/slides/02-02_vba-procedures-slides.Rmd" target="_blank">Source</a> | <a href="https://github.com/bss-osca/tfa/blob/master/vba/vba-procedures-template.xlsm" target="_blank">Excel</a>
</span>
</div> 

---

## Procedures

- A procedure is a piece of code stored in a module. 
- A procedure contains a series of computational steps that will be carried out when the procedure is called. 

VBA has two kinds of procedures:
.pull-left[
Subs:
  - Can make changes to the worksheet.
  - Can modify its surroundings.
  - Can be executed by a button (a macro - if no arguments).
  - Cannot return anything.
]
.pull-right[
Functions:
  - Can return something.
  - Can be used in Excel.
  - Cannot be used as a macro.
]

---

## Declaring a procedure

Sub:

  ```
  Sub MySub(arg1 As {datatype}, arg2 As {datatype})
    ' Some code
  End Sub
  ```

Function:

  ```
  Function MyFunction(arg1 As {datatype}, arg2 As {datatype}) As {return datatype}
    ' Some code
    MyFunciton = value  ' assign a return value to the function
  End Sub
  ```

Hint: You can use `Exit Sub`/`Exit Function` to exit the sub/function early in the code.

---

## Public and Private Procedures

Procedures can be either public or private:

- Public (default): Can be used from other modules, from other files and from Excel.
  ```
  Public Sub MySub()
    ...
  End Sub
  ```
- Private: Can only be used from within its own module.
  ```
  Private Sub MySub()
    ...
  End Sub
  ```

It is good coding practice to use `Private` as much as possible.

---

## Calling a prodedure

Use the `Call` keyword to call a sub:

  ```
  Call MySub(arg1, arg2)
  ```

Call a function by assigning its return value to a variable

  ```
  result = MyFunction(arg1, arg2)
  ```

Let us try some examples in the modules `A_SimpleEx` and `B_InputArgEx` using this [Excel file](https://github.com/bss-osca/tfa/blob/master/vba/vba-procedures-template.xlsm).

---

## Input `ByRef` or `ByVal`

.midi[
There are two ways of passing arguments to procedures:
  ```
  Sub MySub(ByRef arg1 As {datatype}, ByVal arg2 As {datatype})
    ' Some code
  End Sub
  ```

Argument `arg1` is passed **by reference** (default). That is, no new memory is allocated when the procedure is called and hence the procedure can have changed the value of `arg1` when the procedure stops. Note since the default, the result is the same if we omitted the keyword `ByRef`. Argument `arg2` is passed **by value**. That is, a copy of the variable is created in memory with local scope. Hence the procedure cannot change the value of `arg2` when the procedure stops (the local variables is deleted). 

Using `ByRef` is faster and saves memory since we do not have to allocate new memory. We may use `ByRef` to return updated values of the input arguments e.g calling a sub `ReadArrays` that takes an unallocated array as input and afterwards the arrays has been allocated and values from a sheet stored in the array. 

Let us try to guess the output by running the procedure in module `C_ByRefValEx` in the [Excel file](https://github.com/bss-osca/tfa/blob/master/vba/vba-procedures-template.xlsm).
]

---

## Default input arguments

A procedure may often be called with the same input argument value. To avoid always specifying the value, you may set a default value using the `Optional` keyword:
  ```
  Sub MySub(arg1 As {datatype}, Optional arg2 As {datatype} = defaultValue)  
  ```
You can now call the procedure using:
  ```
  Call MySub(arg1)  ' assume that arg2 = defaultValue
  ```
Note: Every parameter following an optional parameter in the procedure definition must also be optional.

Let us try to format some cells by running the procedures in module `D_DefaultArgs` in the [Excel file](https://github.com/bss-osca/tfa/blob/master/vba/vba-procedures-template.xlsm).

---

## Using built-in functions

- VBA has a set of [built-in functions](https://bettersolutions.com/vba/functions/complete-list.htm) such as `Abs`, `Log` and `Date`. You call them by just writing their name:
  ```
  dtm as Date
  dtm = Date()
  ```
- You can also use the [worksheet functions](https://docs.microsoft.com/en-us/office/vba/excel/concepts/events-worksheetfunctions-shapes/list-of-worksheet-functions-available-to-visual-basic) in Excel. You call them using the `WorksheetFunction` object:
  ```
  sum = WorksheetFunction.Sum(Range("A1:D5"))
  ```


```{r copy to docs, include=FALSE}
file.copy(list.files(pattern = ".html"), "../docs/slides/", overwrite = T)
file.copy("./slides.css", "./libs/", overwrite = T)
file.copy("libs", "../docs/slides/", recursive = T)
file.copy("img", "../docs/slides/", recursive = T)
system2("Rscript", args = "-e 'rmarkdown::render(\"index.Rmd\", quiet = TRUE)'")
```