---
title: "Functions"
author: "Lars Relund Nielsen"
output:
  xaringan::moon_reader:
    css: "./libs/slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, child = "setup.Rmd", echo=FALSE}
```

```{r include=FALSE}
# use xaringan::inf_mr(cast_from = ".") to preview
```

layout: true

<div class="my-footer">
<span>
<a href="https://bss-osca.github.io/tfa/mod-r-functions" target="_blank">Book</a> | <a href="https://bss-osca.github.io/tfa/slides/06-02_r-functions-slides.html" target="_blank">Slides</a> | <a href="https://github.com/bss-osca/tfa/blob/master/slides/06-02_r-functions-slides.Rmd" target="_blank">Source</a> | <a href="https://rstudio.cloud/spaces/176810/project/3020322" target="_blank">R project</a>
</span>
</div> 

---

class: middle

> To understand computations in R, two slogans are helpful:
>
>    Everything that exists is an object.
>
>    Everything that happens is a function call.
>
>
>    .right[\- _John Chambers_]

---

## Functions

- Functions are often used to encapsulate a sequence of expressions that need to be executed numerous times, perhaps under slightly different conditions. 
<!-- - In programming, functional programming is a programming paradigm, a style how code is written. Rather than repeating code, functions and control structures allow one to build code in blocks.  -->
- Functions are (often) verbs, followed by what they will be applied to in parentheses.
- Functions can be defined using the `function()` directive:
```{r}
library(tidyverse)
loop_output <- function(dat, rows = 1:2, cols = 2:3, ...) {
  ite <- expand_grid(r = rows, c = cols)
  for (i in 1:nrow(ite)) {
    r <- ite$r[i]
    c <- ite$c[i]
    cat("Row ", r, " column ", c, " = ", dat[r,c], ".\n", ...)
  } 
  return("Finished!")
}
```
- Functions always return the last expression evaluated in the function body or when you use the `return` flow control statement (good coding practice).

---

## Functions - Arguments

```{r}
loop_output(mtcars)  # default values
loop_output(mtcars, rows = c(3,1), cols = 6)  # specify values
```

---

## Functions - The `...` argument

```{r}
loop_output(mtcars, rows = c(3,1), cols = 6)  # specify values
loop_output(mtcars, rows = c(3,1), cols = 6, sep = "")  # use the ... argument
```

---

## Functions - Document your functions

Use **Code > Insert Roxygen skeleton** (Ctrl+Alt+Shift+R).

```{r}
#' Loop through some values in a data frame
#'
#' @param dat A data frame.
#' @param rows Rows to loop through.
#' @param cols Columns to loop through.
#' @param ... Arguments passed to `cat`.
#'
#' @return A string.
loop_output <- function(dat, rows = 1:2, cols = 2:3, ...) {
  ite <- expand_grid(r = rows, c = cols)
  for (i in 1:nrow(ite)) {
    r <- ite$r[i]
    c <- ite$c[i]
    cat("Row ", r, " column ", c, " = ", dat[r,c], ".\n", ...)
  } 
  return("Finished!")
}
```


---

## Your Turn

.your-turn[
- Create a function that given some dice numbers and some cards from a 52-card deck (e.g. 1-spade), write out all the possible combinations as strings in the form `Dice = [value] and card = [value]`. 
- Moreover, if the dice numbers contains the number 2, function must return true; otherwise, false. Hint:

```r
get_combinations <- function(dice, card) {
  ite <- expand_grid(___)
  found <- FALSE
  for (i in 1:nrow(ite)) {
    ___
  } 
  return(found)
}
get_combinations(dice = c(2,3), card = c("2-spade", "ace-dimond"))
get_combinations(dice = c(3,1), card = c("10-heart", "king-dimond"))
```
]

```{r echo=FALSE}
countdown(minutes = 5, top = "0")
```

???

```{r, include=TRUE}
get_combinations <- function(dice, card) {
  ite <- expand_grid(d = dice, c = card)
  found <- FALSE
  for (i in 1:nrow(ite)) {
    d <- ite$d[i]
    c <- ite$c[i]
    cat("Dice ", d, " and card  = ", c, ".\n", sep = "")
    if (d == 2) found = TRUE
  } 
  return(found)
}
get_combinations(dice = c(2,3), card = c("2-spade", "ace-dimond"))
get_combinations(dice = c(3,1), card = c("10-heart", "king-dimond"))
```

---

## Your Turn

.your-turn[
- Create a function that calculates the present value (PV) of a future value. The equation is: 
$$PV = FV/(1+r)^n$$ where $FV$ is future value, $r$ is the interest rate, and $n$ is the number of periods. 
- Input arguments must be $FV$, $r$, $n$ and a boolean `round` equal true if the output should be rounded to two decimals. 
- The default interest rate is 0.1 (10%) and rounding is false by default.
]

```{r echo=FALSE}
countdown(minutes = 5)
```

???

```{r, include=TRUE}
get_PV <- function(FV, r = 0.1, n, round = FALSE) {
  PV <- FV/(1+r)^n
  if (round) return(round(PV, 2))
  return(PV)
}
get_PV(100, n = 7)
get_PV(100, n = 7, round = TRUE)
get_PV(100, n = 7, r = 0.5)
```





```{r copy, include=FALSE}
file.copy(list.files(pattern = ".html"), "../docs/slides/", overwrite = T)
file.copy("./slides.css", "./libs/", overwrite = T)
file.copy("libs", "../docs/slides/", recursive = T)
file.copy("img", "../docs/slides/", recursive = T)
system2("Rscript", args = "-e 'rmarkdown::render(\"index.Rmd\", quiet = TRUE)'")
```

```{r links, child="links.md", include=FALSE}
```
