---
title: "TFA exam"
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90, scipen = 10, digits = 2)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```















# R - `r ctrA()` (50%)
 
Consider the dataset in the file `orders.csv` with purchase orders for a group of ships. 

The dataset contains a row for each item used in an order. The columns/variables are:

  - `ship`: The ship considered.
  - `order_id`: Order id. An order is a group of items purchased in one batch from a single supplier.
  - `item_id`: Item id.
  - `item_desc`: Item description.
  - `quantity`: Number of items ordered.
  - `price`: Price per unit.
  - `order_date`: Order date.
  - `delivery_date`: Expected delivery date when order is made.
  - `delivery_place`: Delivery place.
  - `recieved_date`: Actual date the order is recieved.
  - `supplier`: Supplier for the order.
  - `delivery_cost`: Delivery cost.
  - `order_year`: Year the order was placed. 

The dataset can be read using:

```{r, warning=FALSE}
library(tidyverse)
dat <- read_csv("orders.csv") 
```

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (tibbles/data frames) and answer/complete the following questions/tasks:


## `r ctrQ()`

How many ships, suppliers, and different items are considered?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Do all orders use a single supplier and delivery place? 

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

For each order, calculate/find the number of different items, number of items, total price, total delivery cost, and order year (missing values are assumed zero). 

Which order has the highest delivery cost?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

For each order, find the supplier and delivery place. Hint: The `first` function may be used to select the first item within a group. 

Which supplier and delivery place is used most?

```{r}
# your code here
```

[Your comment here]





## `r ctrQ()`

Add a column to the dataset equal the *value* of the items in a row calculated as the price times the quantity. Next, create a summary table named `res1` that for each item id calculates the aggregated value and arrange them in descending order. 

Which item is the most costly one with respect to the calculated value in `res1`?

```{r}
# your code here
```

[Your comment here]

Given the dataset `res1`, add two new columns:

   - The relative value for each item as the ratio between the item's value and the total value of all items.
   - The cumulative relative value. Hint: You can use the `cumsum` function here.

```{r}
# your code here
```

[Your comment here]

Consider the most costly item with respect to the calculated value in `res1`. At which ships is the item used and how many times?

```{r}
# your code here
```

[Your comment here]





# R - `r ctrA()` (50%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you may need *dplyr* for preparing the datasets you want to plot). We work with the dataset from Assignment 1, which can be read using:

```{r}
library(tidyverse)
dat <- read_csv("orders.csv")
```




## `r ctrQ()`

Create a visualization showing the number of purchase orders for each year with the following features:

* Bars are used for each year.
* Fill colors are used to identify the ship. 
* Legends are put at the bottom of the plot.
* Informative figure title and axis titles are given.

Which year had the most orders?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Create a visualization showing the total number of items ordered for each ship with the following features:

* The numbers are shown using columns.
* Reorder the columns so they increase along the x-axis.
* Informative figure title and axis titles are given.

Which ship uses most items? 

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

For each ship, create a visualization showing the number of items used as a function of done date with the following features:

* The numbers are shown using columns with a fixed line width of 1 and a blue color. 
* The plot is divided using `ship` (facet). Hint: You may use `scales = "free_y"`. 
* Informative figure title and axis titles are given.

Which ship has most items used at a specific date?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Consider items with id:

```{r}
items <- c("601.003.004" ,"601.004.006", "601.026.128", "601.026.052")
```

For each row in the dataset, calculate an index variable as price divided with the maximum item id price. Hint: Group by item id and use `mutate` to add the column. Create a visualization showing the index as a function of order year with the following features:

* The numbers are shown using points.
* The plot is divided using item id (facet). 
* A trend line is added.
* Informative figure title and axis titles are given.

Take a closer look at the item with most price fluctuations. Are the prices reasonable for a singe product? 

```{r}
# your code here
```

[Your comment here]








## `r ctrQ()`

Make a summary table in the following way:

* Add a column `val` with the value of an item in a row equaling the price times quantity.
* Group each item id and set the total value `total_value` to the sum of `val` and calculate the total quantity purchased. 
* Arrange the total value in descending order.
* Add columns:
   - Relative total value equals `total_val/sum(total_val)`.
   - Relative item number equals `row_number()/n()`.
   - Cumulative relative total value (`cum_pct_val`). Hint: You can use the `cumsum` function here.
   - Class equals `class = case_when(cum_pct_val <= 0.8 ~ "A", cum_pct_val <= 0.95  ~ "B", TRUE ~ "C")` classifying items in A (high value items), B (middle), and C (low value items). 

Create a visualization showing the cumulative relative total value as a function of relative item number with the following features:

* Data points are given in the plot.
* Different colors are used for each class.
* Different sizes are used for each quantity.
* A line is added with black color and fixed size 0.3. 
* Informative figure title and axis titles are given.

How many items contribute to 80% of the total value?  

```{r}
# your code here
```

[Your comment here]





