---
title: "TFA exam - Guiding solution"
author: "Lars Relund Nielsen"
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90, scipen = 10, digits = 2)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```















# R - `r ctrA()` (50%)
 
Consider the dataset in the file `orders.csv` with purchase orders for a group of ships. 

The dataset contains a row for each item used in an order. The columns/variables are:

  - `ship`: The ship considered.
  - `order_id`: Order id. An order is a group of items purchased in one batch from a single supplier.
  - `item_id`: Item id.
  - `item_desc`: Item description.
  - `quantity`: Number of items ordered.
  - `price`: Price per unit.
  - `order_date`: Order date.
  - `delivery_date`: Expected delivery date when order is made.
  - `delivery_place`: Delivery place.
  - `recieved_date`: Actual date the order is recieved.
  - `supplier`: Supplier for the order.
  - `delivery_cost`: Delivery cost.
  - `order_year`: Year the order was placed. 

The dataset can be read using:

```{r, warning=FALSE}
library(tidyverse)
dat <- read_csv("orders.csv") 
```

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (tibbles/data frames) and answer/complete the following questions/tasks:


## `r ctrQ()`

How many ships, suppliers, and different items are considered?

```{r, solution=TRUE, warning=FALSE}
# load packages
library(tidyverse)
library(skimr)
skim(dat)
```

<div class="solution">
Orders are for 9 ships and from 31 suppliers. A total of 550 different items are considered.
</div>




## `r ctrQ()`

Do all orders use a single supplier and delivery place? 

```{r, solution=TRUE}
dat %>% 
   count(order_id, supplier) %>% 
   count(order_id) %>% 
   filter(n > 1)

dat %>% 
   count(order_id, delivery_place) %>% 
   count(order_id) %>% 
   filter(n > 1)
```

<div class="solution">
I count the number of suppliers per order id. Since there is only a single supplier per order id, orders are sent from a single supplier as expected. Similar for delivery place.
</div>




## `r ctrQ()`

For each order, calculate/find the number of different items, number of items, total price, total delivery cost, and order year (missing values are assumed zero). 

Which order has the highest delivery cost?

```{r, solution=TRUE}
res <- dat %>% 
   group_by(order_id) %>% 
   summarise(
      items = n(), 
      quantity = sum(quantity, na.rm = T),
      price = sum(price * quantity, na.rm = T), 
      delivery_cost = sum(delivery_cost, na.rm = T),
      year = first(order_year)) %>% 
   arrange(desc(delivery_cost)) %>% 
   print()
```

<div class="solution">
The highest delivery cost is `r res$delivery_cost[1]` for order `r res$order_id[1]`.
</div>




## `r ctrQ()`

For each order, find the supplier and delivery place. Hint: The `first` function may be used to select the first item within a group. 

Which supplier and delivery place is used most?

```{r, solution=TRUE}
res1 <- dat %>% 
   group_by(order_id) %>% 
   summarise(supplier = first(supplier), delivery_place = first(delivery_place)) 
res2 <- res1 %>% 
   count(supplier, sort = T) %>% 
   print()
res3 <- res1 %>% 
   count(delivery_place, sort = T) %>% 
   print()
```

<div class="solution">
The supplier used most is `r res2$supplier[1]`. The delivery place used most is `r res3$delivery_place[1]`.
</div>





## `r ctrQ()`

Add a column to the dataset equal the *value* of the items in a row calculated as the price times the quantity. Next, create a summary table named `res1` that for each item id calculates the aggregated value and arrange them in descending order. 

Which item is the most costly one with respect to the calculated value in `res1`?

```{r, solution=TRUE}
res1 <- dat %>% 
   mutate(val = price * quantity) %>% 
   group_by(item_id) %>% 
   summarise(val = sum(val), item_desc = first(item_desc)) %>% 
   arrange(desc(val)) %>% 
   print()
```

<div class="solution">
The most costly item is `r res1$item_id[1]` a `r res1$item_desc[1]`. 
</div>

Given the dataset `res1`, add two new columns:

   - The relative value for each item as the ratio between the item's value and the total value of all items.
   - The cumulative relative value. Hint: You can use the `cumsum` function here.

```{r, solution=TRUE}
res1 <- res1 %>% 
   ungroup() %>%
   mutate(pct = val/sum(val), cum_pct = cumsum(pct)) %>%
   print() 
```

<div class="solution">
The two new columns have been added. 
</div>

Consider the most costly item with respect to the calculated value in `res1`. At which ships is the item used and how many times?

```{r, solution=TRUE}
item <- res1$item_id[1]
res2 <- dat %>% 
   filter(item_id == item) %>% 
   group_by(ship) %>% 
   summarise(n = sum(quantity)) %>% 
   arrange(desc(n)) %>% 
   print() 
```

<div class="solution">
The item is used at `r nrow(res2)` ships (most on ship `r res2$ship[1]`).
</div>





# R - `r ctrA()` (50%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you may need *dplyr* for preparing the datasets you want to plot). We work with the dataset from Assignment 1, which can be read using:

```{r}
library(tidyverse)
dat <- read_csv("orders.csv")
```




## `r ctrQ()`

Create a visualization showing the number of purchase orders for each year with the following features:

* Bars are used for each year.
* Fill colors are used to identify the ship. 
* Legends are put at the bottom of the plot.
* Informative figure title and axis titles are given.

Which year had the most orders?

```{r, solution=TRUE}
dat %>% 
   group_by(order_year, order_id) %>% 
   summarise(ship = first(ship)) %>% 
   ggplot(aes(x = order_year, fill = ship)) +
   geom_bar() +
   labs(title = "Number of orders",
    x = "Year",
    y = "Orders",
    col = "Ship") +
   theme(legend.position = "bottom")
# or
# dat %>%
#    count(order_year, order_id, ship) %>%
#    ggplot(aes(x = order_year, fill = ship)) +
#    geom_bar()
```

<div class="solution">
Most orders are in 2018.
</div>




## `r ctrQ()`

Create a visualization showing the total number of items ordered for each ship with the following features:

* The numbers are shown using columns.
* Reorder the columns so they increase along the x-axis.
* Informative figure title and axis titles are given.

Which ship uses most items? 

```{r, solution=TRUE}
res <- dat %>% 
   group_by(ship) %>% 
   summarize(n = sum(quantity)) %>% 
   arrange(desc(n)) 
res %>% 
   ggplot(aes(x = reorder(ship, n), y = n)) +
   geom_col() +
   labs(title = "Total number of items used",
    x = "Ship",
    y = "Items") +
   theme(legend.position = "bottom")
```

<div class="solution">
Most items are used for `r res$ship[1]`.
</div>




## `r ctrQ()`

For each ship, create a visualization showing the number of items used as a function of done date with the following features:

* The numbers are shown using columns with a fixed line width of 1 and a blue color. 
* The plot is divided using `ship` (facet). Hint: You may use `scales = "free_y"`. 
* Informative figure title and axis titles are given.

Which ship has most items used at a specific date?

```{r, solution=TRUE, warning=FALSE}
res <- dat %>% 
   group_by(ship, order_date) %>% 
   summarize(n = sum(quantity)) 
res %>% 
   ggplot(aes(x = order_date, y = n)) +
   geom_col(linewidth = 1, color = "blue") +
   facet_wrap(~ ship, scales = "free_y") +
   labs(title = "Number of items used",
    x = "Date",
    y = "Items used") 
```

<div class="solution">
On a specific date must items are used for Ship-13.
</div>




## `r ctrQ()`

Consider items with id:

```{r}
items <- c("601.003.004" ,"601.004.006", "601.026.128", "601.026.052")
```

For each row in the dataset, calculate an index variable as price divided with the maximum item id price. Hint: Group by item id and use `mutate` to add the column. Create a visualization showing the index as a function of order year with the following features:

* The numbers are shown using points.
* The plot is divided using item id (facet). 
* A trend line is added.
* Informative figure title and axis titles are given.

Take a closer look at the item with most price fluctuations. Are the prices reasonable for a singe product? 

```{r, solution=TRUE}
res <- dat %>% 
   filter(item_id %in% items) %>% 
   group_by(item_id) %>% 
   mutate(idx = price/max(price))
res %>% 
   ggplot(aes(y = idx, x = order_year)) + 
   geom_point() +
   geom_smooth(se = F) +
   facet_wrap(~ item_id) +
   labs(title = "Price index (index 1 = max price)",
    x = "Order year",
    y = "Index") 
dat %>% 
   filter(item_id %in% "601.026.128")  
```

<div class="solution">
If consider item 601.026.128 it can be seen that prices are very different. By considering item description it can be seen that the item is in fact not a single product, i.e we have a misclassification of item id!
</div>








## `r ctrQ()`

Make a summary table in the following way:

* Add a column `val` with the value of an item in a row equaling the price times quantity.
* Group each item id and set the total value `total_value` to the sum of `val` and calculate the total quantity purchased. 
* Arrange the total value in descending order.
* Add columns:
   - Relative total value equals `total_val/sum(total_val)`.
   - Relative item number equals `row_number()/n()`.
   - Cumulative relative total value (`cum_pct_val`). Hint: You can use the `cumsum` function here.
   - Class equals `class = case_when(cum_pct_val <= 0.8 ~ "A", cum_pct_val <= 0.95  ~ "B", TRUE ~ "C")` classifying items in A (high value items), B (middle), and C (low value items). 

Create a visualization showing the cumulative relative total value as a function of relative item number with the following features:

* Data points are given in the plot.
* Different colors are used for each class.
* Different sizes are used for each quantity.
* A line is added with black color and fixed size 0.3. 
* Informative figure title and axis titles are given.

How many items contribute to 80% of the total value?  

```{r, solution=TRUE, warning=FALSE}
res1 <- dat %>% 
   mutate(val = price * quantity) %>% 
   group_by(item_id) %>% 
   summarise(total_val = sum(val), item_desc = first(item_desc), quantity = sum(quantity)) %>% 
   arrange(desc(total_val)) %>% 
   mutate(pct_val = total_val/sum(total_val), 
          pct_item = row_number()/n(),
          cum_pct_val = cumsum(pct_val), 
          class = case_when(
            cum_pct_val <= 0.8 ~ "A",
            cum_pct_val <= 0.95  ~ "B",
            TRUE ~ "C"
         )) %>% 
   # tail() %>% 
   print() 

res1 %>% 
   ggplot(aes(x = pct_item, y = cum_pct_val, col = class, size = quantity)) +
   geom_point() +
   geom_line(col = "black", size = 0.3) +
   labs(title = "Usage of items (one point per item)",
    x = "Item percentage",
    y = "Cumulative value",
    col = "Class",
    size = "Purchased") 
```

<div class="solution">
Items in the A class (80% of value) are approx. 7% of the total items. 
</div>





