---
title: "Mutating and filtering joins"
author: "Lars Relund Nielsen"
date: "`r Sys.Date()`"
repository_url: https://github.com/bss-osca/tfa/blob/master/slides/05-joins/05-joins_examples.Rmd
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
if (interactive()) setwd(dirname(rstudioapi::getActiveDocumentContext()$path)) # set working dir to current file location
knitr::opts_chunk$set(echo = TRUE, 
                      cache = TRUE, autodep = TRUE)
options(width = 100)  # set output width
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center",
                      message = FALSE)
```

## Setup

```{r}
library(tidyverse)
band_members
band_instruments
band_instruments2
```


## Mutating joins

Mutating joins allow you to combine variables from multiple tables. There are different types of joins. We will consider:

- `inner_join()`: all rows from x where there are matching values in y, return 
all combination of multiple matches in the case of multiple matches
- `left_join()`: all rows from x
- `right_join()`: all rows from y
- `full_join()`: all rows from both x and y

While mutating joins are primarily used to add new variables, they can also generate new observations. If a match is not unique, a join will add all possible combinations (the Cartesian product) of the matching observations.


### `inner_join()`

```{r}
band_members %>% 
  inner_join(band_instruments)
```


### `left_join()`


```{r}
band_members %>% 
  left_join(band_instruments)
```

### `right_join()`


```{r}
band_members %>%
  right_join(band_instruments)
```


### `full_join()`


```{r}
band_members %>% 
  full_join(band_instruments)
```


## Filtering joins

Filtering joins match observations in the same way as mutating joins, but affect the observations, not the variables. There are two types:

- `semi_join()`: all rows from x where there are matching values in y, keeping just columns from x.
- `anti_join()`: return all rows from x where there are not matching values in y, never duplicate rows of x
 
These are most useful for diagnosing join mismatches. If you're worried about what observations your joins will match, start with a `semi_join()` or `anti_join()`. `semi_join()` and `anti_join()` never duplicate; they only remove observations. 


### `semi_join()`


```{r}
band_members %>% 
  semi_join(band_instruments)
```


### `anti_join()`


```{r}
band_members %>% 
  anti_join(band_instruments)
```


## What if column names are not the same?

Use the `by`argument:

```{r}
band_members
band_instruments2
band_members %>% 
  full_join(
    band_instruments2, 
    by = c("name" = "artist"))
```
]