---
title: "`r params$title`"
author: "`r params$author`"
date: "`r params$date`"
output: html_document
params:
  doc_output: solution
  title: "Main file TFA reexam (R part)"
  author: "Lars Relund Nielsen"
  date: ""
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90)

a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}
```


```{r main, include=FALSE}
doc_output <- params$doc_output

knitr::knit_engines$set(
  templatetext = function(options) {
    if (doc_output == "template") return(options$code)
    return("")
  },
  solutiontext = function(options) {
    if (doc_output == "solution") 
      return(paste0('\n<div class="solution">\n', options$code, '</div>'))
    return("")
  },
  privatetext = function(options) {
    return("")
  })

if (doc_output == "solution") {
  knitr::opts_hooks$set(solution = function(options) {
    if (isTRUE(options$solution)) {
      options$eval = TRUE
    }
    options
  })
  knitr::knit_hooks$set(solution = function(before, options, envir) {
    if (before) {
      paste0('\n<div class="solution">\nGuiding solution:\n\n')
    } else {
      paste0('\n</div>\n')
    }
  })
}

if (doc_output == "template") {
  knitr::opts_hooks$set(output = function(options) {
    if (!is.na(options$output)) {
      if (options$output == "solution") {
        options$code = "# Your code here"
      }
    }
    options
  })
}

if (doc_output == "exam") {
  knitr::opts_hooks$set(solution = function(options) {
    if (isTRUE(options$solution)) {
      options$eval = FALSE
      options$include = FALSE
    }
    options
  })
}
```

In the following, use R to answer the questions/assignments. Your solutions must be written in a R Markdown (Rmd) file. This file must include your code and a (brief) comment for each question. For example, "The three companies with smallest profit are ..." or "The plot shows that ...".

An R markdown template file is given in the appendix that you may use as a starting point. 

You may load and use the following packages:

```{r}
library(tidyverse)
library(skimr)
```


# R - `r ctrA()` (10%)

Consider the vector:

```{r}
v <- c(9, 19, 2, 8, NA, 12, 9, 23, NA, 34)
v
```

Use R to answer the following questions.

## `r ctrQ()`

Is any of the entries in `v` below or equal to 2?

```{r, solution=TRUE}
any(v <= 2)
```


<div class="solution">
Yes since the vector contains 2.</div>




## `r ctrQ()` 

Is all of the entries in `v` above or equal to 2?

```{r, solution=TRUE}
all(v >= 2)
```


<div class="solution">
We don't know since we have missing values.</div>


## `r ctrQ()`

Does `v` have missing values?

```{r, solution=TRUE}
any(is.na(v))
```


<div class="solution">
Yes, since `v` contains `NA` values.</div>


## `r ctrQ()` 

Which entries in `v` are above 10? You must return the indices, e.g. the index of `v[3]` is 3.

```{r, solution=TRUE}
which(v > 10)
```


<div class="solution">
The index are `r which(v > 10)`.</div>


## `r ctrQ()` 

Create a vector `res` where `res[i]` is equal to `v[i]` if `v[i]` is less than 10 and otherwise zero (also if `v[i]` is `NA`).  

```{r, solution=TRUE}
res <- if_else(v < 10, v, 0, 0)
res
```


<div class="solution">
We use `if_else` to set values.</div>




# R - `r ctrA()` (20%)

The dataset `titanic`, given in the appendix, lists approx. 1300 passengers on Titanic. The column/variables are:

* `pclass`: Passenger class (1 = 1st; 2 = 2nd; 3 = 3rd).
* `survived`: Survival (0 = No; 1 = Yes).
* `name`: Name.
* `sex`: Sex.
* `age`: Age.
* `fare`: Passenger Fare.
* `cabin`: Cabin number.
* `embarked`: Port of embarkation (C = Cherbourg; Q = Queenstown; S = Southampton).
* `boat`: Lifeboat number.

You can read the dataset file `titanic.csv` into the dataset `dat` using

```{r}
dat <- read_csv("titanic.csv")
```

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (data frames) and answer the following questions.



## `r ctrQ(reset=T)`

Create a new column named `male` which is true if the person is a male. 

```{r, solution=TRUE}
dat <- dat %>% 
  mutate(male = if_else(sex == "male", TRUE, FALSE))
dat
```


<div class="solution">
We use `if_else` to set the new column.</div>


## `r ctrQ()` 

How many persons are we considering, how many men (in percentage) and how many survived? 


```{r, solution=TRUE}
skim(dat)
```


<div class="solution">
From the output we can consider 1309 persons of which approx 64% are males. Approx. 38% of the passangers survived. </div>


## `r ctrQ()`

How many of the females survived in percent (and how many males)?

```{r, solution=TRUE}
dat %>% 
  group_by(sex, survived) %>% 
  count() %>% 
  group_by(sex) %>% 
  mutate(rate = n/sum(n))
# Alternatively
res <- dat %>% 
  group_by(male) %>% 
  summarise(survived = sum(survived)/n()) %>% 
  print()
```


<div class="solution">
The survival rate for women and men are `r str_c(round(100 * res %>% pull(survived)), collapse = " and ")` percent, respectively.</div>




## `r ctrQ()`

Define children as people with age below 19. How many children survived?

```{r solution=TRUE}
res <- dat %>% 
  filter(age < 19, survived == TRUE) %>% 
  nrow() %>% 
  print()
```


<div class="solution">
`r res` childern survived. </div>


## `r ctrQ()`

Did relatively more people survive at first class compared to third class?

```{r, solution=TRUE}
res <- dat %>% 
  group_by(pclass) %>% 
  summarise(rate = sum(survived)/n()) %>% 
  print()
```


<div class="solution">
There seems to be a big difference in survival rate between first class (`r round(100 * res %>% filter(pclass == 1) %>% pull(rate))`%) and third class (`r round(100 * res %>% filter(pclass == 3) %>% pull(rate))`%).</div>




## `r ctrQ()`

How many persons that entered a lifeboat did die in percent? 

```{r, solution=TRUE}
res <- dat %>% 
  filter(!is.na(boat)) %>% 
  summarise(rate = 1-sum(survived)/n()) %>% 
  print()
```


<div class="solution">
The survival rate when entered the lifeboat was high. Only `r round(100 * res %>% pull(), 2)`% died.</div>


## `r ctrQ()`

How many persons with `Hansen` in their name survived?

```{r, solution=TRUE}
dat %>% 
  filter(str_detect(name, "Hansen"))
```


<div class="solution">
Only a single person survived.</div>





# R - `r ctrA()` (20%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you might need dplyr for preparing the datasets you want to plot). We work with the dataset `titanic` from Assignment 2 which can be read using:

```{r}
dat <- read_csv("titanic.csv")
```


## `r ctrQ(reset=T)`

Create a visualization showing the number of persons embarking the different harbors with the following features:

* Number of persons is represented using bars.
* Informative figure title and axis titles are given.
* The labels on the x-axis are the harbor names (not abbreviations).

What harbor was the main departure harbor?

```{r, solution=TRUE}
dat <- dat %>% mutate(harbor = case_when(
  embarked == "C" ~ "Cherbourg",
  embarked == "Q" ~ "Queenstown",
  embarked == "S" ~ "Southampton",
  TRUE ~ NA_character_
))
dat %>% 
  ggplot(aes(x = harbor)) +
  geom_bar() +
  labs(title = "Number of persons embarking each departure harbor",
       x = "Harbor",
       y = "Persons") 
```


<div class="solution">
The main harbor was Southampton.</div>

## `r ctrQ()`

Create a visualization showing the age distribution of the persons with the following features:

* Number of persons is represented using a histogram.
* Informative figure title and axis titles are given.

Where there most people on board in their twenties or thirties?



```{r, solution=TRUE}
dat %>% 
  ggplot(aes(x = age)) +
  geom_histogram(binwidth = 2) +
  labs(title = "Age distribution",
       x = "Age",
       y = "Persons") 
```


<div class="solution">
There where most people in their twenties. </div>


## `r ctrQ()`

Create a visualization showing the fare as a function of age with the following features:

* Different colors are used for each sex. 
* Informative figure title and axis titles are given.
* A trend line for each sex is added using `geom_smooth`.

Based on the trend lines, do females in general pay more for a ticket?

```{r, solution=TRUE}
dat %>% ggplot(aes(x = age, y = fare, color = sex)) +
  geom_point() + 
  geom_smooth() +
  labs(title = "Fare prices given age",
       x = "Age",
       y = "Price",
       color = "Sex") + 
  theme(legend.position = "bottom")
```


<div class="solution">
Based on the trend lines females in general pays more than males.</div>

## `r ctrQ()`

Create a visualization showing the survival rate for each passenger class with the following features:

* Bars are used for each passenger class. 
* All bars have same height (100 %).
* Colors are used to identify who survived and did not survived.
* Informative figure title and axis titles are given.

Is the survival rate different on first and third class?

```{r, solution=TRUE}
dat %>% 
  ggplot(aes(x = pclass, fill = as.logical(survived))) +
  geom_bar(position="fill") +
  labs(title = "Survival rate given passenger class",
       x = "Class",
       y = "",
       fill = "Survived") + 
  theme(legend.position = "bottom")
```


<div class="solution">
Based on the plot most people survived at first classs.</div>



## `r ctrQ()`

Let column `level` denote the the first letter in `cabin`. Create a visualization showing the variance in fare prices with the following features:

* Ignore rows with missing `level`.
* Variation is shown using a boxplot. 
* Informative figure title and axis titles are given.

Is the fare price different at the B and C level?

```{r, solution=TRUE}
dat <- dat %>% 
  mutate(level = str_sub(cabin, 1, 1)) 

  dat %>% filter(!is.na(level)) %>% 
    ggplot(aes(x = level, y = fare)) +
    geom_boxplot() 
```


<div class="solution">
Based on the plot the fare prices at B and C level are more or less the same.</div>









