---
title: "`r params$title`"
author: "`r params$author`"
date: "`r params$date`"
output: 
   html_document:
     toc: false
   pdf_document:
      includes:
         in_header: "preamble.tex"
      # extra_dependencies:
      #    a4wide: null
      #    lmodern: null
      #    newtxtext: ["scaled=1", "helvratio=0.95"]
      #    newtxmath: null
fontsize: 11pt
classoption: a4paper
# bibliography: litt.bib  
params:
  doc_output: solution
  title: "Main file TFA reexam (R and VBA part)"
  author: "Lars Relund Nielsen"
  date: ""
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```

```{r main, include=FALSE}
doc_output <- params$doc_output

if (doc_output == "solution") {
  knitr::opts_hooks$set(solution = function(options) {
    if (isTRUE(options$solution)) {
      options$eval = TRUE
    }
    options
  })
  knitr::knit_hooks$set(solution = function(before, options, envir) {
    if (before) {
      paste0('\n<div class="solution">\nGuiding solution:\n\n')
    } else {
      paste0('\n</div>\n')
    }
  })
}

if (doc_output == "template") {
  knitr::opts_hooks$set(output = function(options) {
    if (!is.na(options$output)) {
      if (options$output == "solution") {
        options$code = "# Your code here"
      }
    }
    options
  })
}

if (doc_output == "exam") {
  knitr::opts_hooks$set(solution = function(options) {
    if (isTRUE(options$solution)) {
      options$eval = FALSE
      options$include = FALSE
    }
    options
  })
}
```

\includepdf[fitpaper=true, pagecommand={\pagestyle{fancy}}]{front.pdf}




# R - `r ctrA()` (15%)
 
```{r gen-data, eval=FALSE, include=FALSE, exam=TRUE}
dat <- tibble(sex = "male", 
   income = rnorm(1845, 43000, 12000) + rnorm(1845, 0, 1200)) %>% 
   bind_rows(
      tibble(sex = "female", 
   income = rnorm(967, 36000, 10000) + rnorm(967, 0, 1200))
   )
write_csv(dat, "income.csv")
```

Consider the dataset in the file `income.csv` with the monthly income from a population. You can read the dataset file using

```{r}
library(tidyverse)
dat <- read_csv("income.csv")
head(dat)
```

## `r ctrQ()` 

Fit probability distributions to data for both men and women. 

<div class="solution">
Guiding solution:

The data are continuous and we try to fit a continuous distribution.

**Male:**
</div>

```{r, solution=TRUE}
library(fitdistrplus)
datM <- dat %>% 
   filter(sex == "male") %>% 
   pull(income)
plotdist(datM, histo = TRUE, demp = TRUE)
descdist(datM)
```

<div class="solution">
A normal distribution seem as a good choice.
</div>

```{r, solution=TRUE}
fitM <- fitdist(datM, distr = "norm")
summary(fitM)
par(mfrow = c(2, 2))  # 4 plots in one figure
pLegend <- c("normal")
lst <- list(fitM)
denscomp(lst, legendtext = pLegend)
qqcomp(lst, legendtext = pLegend)
cdfcomp(lst, legendtext = pLegend)
ppcomp(lst, legendtext = pLegend)
```

<div class="solution">
A normal distribution seem to fit very well.

**Female**:
</div>

```{r, solution=TRUE}
datF <- dat %>% 
   filter(sex == "female") %>% 
   pull(income)
plotdist(datF, histo = TRUE, demp = TRUE)
descdist(datF)
```

<div class="solution">
A normal distribution seem as a good choice.
</div>

```{r, solution=TRUE}
fitF <- fitdist(datF, distr = "norm")
summary(fitF)
par(mfrow = c(2, 2))  # 4 plots in one figure
pLegend <- c("normal")
lst <- list(fitF)
denscomp(lst, legendtext = pLegend)
qqcomp(lst, legendtext = pLegend)
cdfcomp(lst, legendtext = pLegend)
ppcomp(lst, legendtext = pLegend)
```

<div class="solution">
A normal distribution seem to fit very well.
</div>

## `r ctrQ()` 

Make a plot using `ggplot` of the two empirical densities with the following features:

* Different fill colours are used for each sex.
* Informative figure title and axis titles are given.

Based on the plot and Question 1, is there any difference in the income for men and women? If yes on average how much?

```{r, solution=TRUE}
dat %>% 
   ggplot(aes(x = income, fill = sex)) +
   geom_density() +
   labs(title = "Density for each sex",
       x = "Income",
       y = "",
       fill = "Sex"
    ) +
   theme(legend.position = "bottom") 
```

<div class="solution">
Based on the plot there seems to be a difference in income for men and women. We may use the mean values estimated to find the average difference:
</div>

```{r, solution = TRUE}
fitM$estimate[1] - fitF$estimate[1]
```







# R - `r ctrA()` (18%)

```{r get-data, eval=FALSE, include=FALSE, exam=TRUE}
library(tidyverse)
library(skimr)
library(lubridate)
dat <- read_csv("owid-covid-data.csv") %>%
   select(
      date,
      country = location,
      cases = new_cases,
      deaths = new_deaths,
      icu_patients,
      hosp_patients,
      tests = total_tests,
      positive_rate,
      vac = people_vaccinated,
      fully_vac = people_fully_vaccinated,
      population
   ) %>%
   filter(country %in% c("Denmark", "Germany", "Norway", "Sweeden", "United Kingdom"), cases >= 0 | is.na(cases)) %>%
   mutate(month = month(date), year = year(date)) %>% 
   write_csv("covid.csv")

dat <- read.csv("covid.csv") %>% as_tibble()
dat
```

Consider COVID-19 data obtained from Our World in Data in the file `covid.csv`. The dataset contains data from different countries. Some of the columns/variables are:

  * `cases`: New confirmed cases of COVID-19.
  * `deaths`:  New deaths attributed to COVID-19.
  * `icu_patients`: Number of COVID-19 patients in intensive care units (ICUs) on a given day.
  * `hosp_patients`: Number of COVID-19 patients in hospital on a given day.
  * `tests`: Total tests for COVID-19.
  * `positive_rate`: The share of COVID-19 tests that are positive, given as a rolling 7-day average.
  * `vac`: Total number of people who received at least one vaccine dose.
  * `fully_vac`: Total number of people who received all doses prescribed by the vaccination protocol.
  * `population`: Country population.

Other columns are `date`, `country`, `month` and `year`. You can read the dataset file using

```{r}
library(tidyverse)
dat <- read_csv("covid.csv")
```

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (tibbles/data frames) and answer the following questions.


## `r ctrQ()`

Which countries are considered and what is the timespan of the data?

```{r, solution=TRUE}
res1 <- dat %>%
   distinct(country) %>% 
  print()

res2 <- dat %>% 
   ungroup() %>% 
   summarise(start = min(date), end = max(date)) %>% 
   print()
```

<div class="solution">
We have a total of `r res1 %>% nrow()` countries with data from `r res2$start` to `r res2$end`.
</div>



## `r ctrQ()`

What is the number of new confirmed cases November 22nd, 2021 in Denmark?



```{r, solution=TRUE}
res <- dat %>%
   filter(country == "Denmark", date == "2021-11-22") %>% 
  print()
```

<div class="solution">
The number of confirmed cases was `r res$cases`.
</div>


## `r ctrQ()` 

Calculate the total number of confirmed cases and deaths. Hint: you may use the `cumsum` function to add all cases up until a given date. You may here consider `NA` values in the `cases` and `deaths` columns as equal to zero (e.g. using `replace_na(cases, 0)`). 

What is the total number of deaths in Norway up to October 10th, 2021?

```{r, solution=TRUE}
res1 <- dat %>%
   group_by(country) %>%
   mutate(total_cases = cumsum(replace_na(cases, 0)),
          total_deaths = cumsum(replace_na(deaths, 0))) %>%
   print()

res2 <- res1 %>% 
   filter(country == "Norway", date == "2021-10-10") %>% 
   select(contains("total")) %>% 
   print()
```

<div class="solution">
The total number of deaths in Norway up to 10th October 2021 is `r res2$total_deaths`.
</div>


## `r ctrQ()`

For each country calculate the number of tests done in each month in a given year. Which country had the highest number of tests per capita in March 2021?

```{r, solution=TRUE}
res1 <- dat %>%
   group_by(country, month, year, population) %>% 
   summarize(tests = max(tests, na.rm = TRUE) - min(tests, na.rm = TRUE)) %>% 
   ungroup() %>% 
   mutate(testsCap = tests/population) %>% 
   arrange(desc(year), month) %>% 
   print()

res2 <- res1 %>% 
   filter(year == 2021, month == 3) %>% 
   arrange(desc(testsCap)) %>% 
   print()
```

<div class="solution">
The highest number of tests per capita in March 2021 was in `r res2$country[1]`.
</div>


## `r ctrQ()`

Consider United Kingdom. Which month had the highest number of ICU patients on a given day?

```{r, solution=TRUE}
res <- dat %>%
   filter(country == "United Kingdom") %>% 
   group_by(country, year, month) %>% 
   summarize(icu = max(icu_patients, na.rm = TRUE)) %>% 
   arrange(desc(icu)) %>% 
   print()
```

<div class="solution">
The highest number of ICU patients on a given day was in January (`r res$icu[1]` patients).
</div>









# R - `r ctrA()` (17%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you may need *dplyr* for preparing the datasets you want to plot). We work with the dataset from Assignment 2 which can be read using:

```{r}
library(tidyverse)
dat <- read_csv("covid.csv")
```

## `r ctrQ()`

Create a visualization showing the number of cases for each date in Denmark with the following features:

* A blue line is used to visualize the data.
* Informative figure title and axis titles are given.

Is the number of cases low or high in July 2020 in the plot?

```{r, solution=TRUE}
dat %>%
   filter(country == "Denmark", !is.na(cases)) %>% 
   ggplot(aes(y = cases, x = date)) +
   geom_line(color = "blue") +
   labs(title = "Number of cases in Denmark",
       x = "Date",
       y = "Cases"
    )
```

<div class="solution">
The number of cases in July 2020 are low.
</div>


## `r ctrQ()`

Create a visualization showing the total number of deaths per 100000 capita as a function of date with the following features:

* Different colours are used for each country.
* Lines are used to visualize the data.
* Legends are put at the bottom of the plot.
* Informative figure title and axis titles are given.

Hint: you may use the `cumsum` function to add all deaths up until a given date. You may here consider `NA` values in the `deaths` column as equal to zero (e.g. using `replace_na(deaths, 0)`). 

Which country has the highest relative number of deaths in general? 

```{r, solution=TRUE}
dat %>%
   group_by(country) %>%
   mutate(total_deaths = cumsum(replace_na(deaths, 0))) %>%
   mutate(total_deaths_cap = total_deaths/(population/100000)) %>% 
   ggplot(aes(x = date, y = total_deaths_cap, color = country)) +
   geom_line() +
   labs(title = "Total number of deaths per 100000",
       x = "Date",
       y = "Deaths per 100000 capita",
       color = "Country"
    ) +
   theme(legend.position = "bottom") 
```

<div class="solution">
The highest relative number of deaths are for the United Kingdom.
</div>


## `r ctrQ()`

Create a visualization showing the percentage of persons vaccinated as a function of date with the following features:

* We consider 2021. 
* Different colours are used to differ between vaccinated and fully vaccinated.
* The plot is divided using `country` (facet).
* Lines are used to visualize the data.
* Informative figure title and axis titles are given.

Hint: If you calculated the two percentages in two new columns `partly` and `full`, then the values can be joined to one column using

```{r, eval=FALSE}
dat %>% 
   pivot_longer(cols = c(partly, full))
```

Which country has the highest percentage of vaccinated people and the lowest gap between partly and fully vaccinated?

```{r, solution=TRUE}
dat %>% 
   filter(year == 2021) %>% 
   mutate(partly = vac/population, 
          full = fully_vac/population) %>%
   pivot_longer(cols = c(partly, full)) %>% 
   ggplot() + 
   geom_line(aes(x = date, y = value, color = name)) + 
   facet_wrap(vars(country)) +
   labs(title = "Percentage of vaccinated people",
       x = "Date",
       y = "%",
       color = "Vac. type") +
   theme(legend.position = "bottom") 
```

<div class="solution">
The plot showes that Denmark has the highest percentage of vaccinated people and the lowest gap between partly and full vaccinated.
</div>


## `r ctrQ()`

Consider Germany. Create a visualization showing the variation in ICU patients with the following features:

* A sub-plot is given for each month and year (facet).
* Informative figure title and axis titles are given.

In which year did the ICU have the most patients when considering October?

```{r, solution=TRUE}
dat %>%
   filter(country == "Germany") %>% 
  ggplot(aes(x = icu_patients)) +
  geom_boxplot() +
  facet_grid(month ~ year, scales = "free_y") +
  labs(title = "ICU patients given year and month",
       x = "",
       y = "Patients") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), 
        legend.position = "bottom")
```

<div class="solution">
In 2021 the mean value of the ICU patients was highest when considering October.
</div>



## `r ctrQ()`

Create a visualization showing the total number of deaths per 100000 capita for each country and year with the following features:

* The numbers are shown using columns for each country
* Different fill colours are used for year. Hint: columns for each year can be shown beside each other using `position = position_dodge()`.
* Informative figure title and axis titles are given.

Which country had the lowest number of deaths in 2021?

```{r, solution=TRUE}
dat %>%
   group_by(country, year) %>% 
   summarise(deaths = sum(deaths/(population/100000) , na.rm = TRUE)) %>% 
   ggplot(aes(y = deaths, x = country, fill = factor(year))) +
   geom_col(position = position_dodge()) +
   labs(title = "Total number of deaths",
    x = "",
    y = "Deaths",
    fill = "Year") +
   theme(legend.position = "bottom")
```

<div class="solution">
Norway had the lowest number of deaths in 2021.
</div>



