---
title: "`r params$title`"
author: "`r params$author`"
date: "`r params$date`"
output: html_document
params:
  doc_output: solution
  title: "Main file TFA exam (R part)"
  author: "Lars Relund Nielsen"
  date: ""
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)

a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}
```


```{r main, include=FALSE}
doc_output <- params$doc_output

knitr::knit_engines$set(
  templatetext = function(options) {
    if (doc_output == "template") return(options$code)
    return("")
  },
  solutiontext = function(options) {
    if (doc_output == "solution") 
      return(paste0('\n<div class="solution">\n', options$code, '</div>'))
    return("")
  },
  privatetext = function(options) {
    return("")
  })

if (doc_output == "solution") {
  knitr::opts_hooks$set(solution = function(options) {
    if (isTRUE(options$solution)) {
      options$eval = TRUE
    }
    options
  })
  knitr::knit_hooks$set(solution = function(before, options, envir) {
    if (before) {
      paste0('\n<div class="solution">\nGuiding solution:\n\n')
    } else {
      paste0('\n</div>\n')
    }
  })
}

if (doc_output == "template") {
  knitr::opts_hooks$set(output = function(options) {
    if (!is.na(options$output)) {
      if (options$output == "solution") {
        options$code = "# Your code here"
      }
    }
    options
  })
}

if (doc_output == "exam") {
  knitr::opts_hooks$set(solution = function(options) {
    if (isTRUE(options$solution)) {
      options$eval = FALSE
      options$include = FALSE
    }
    options
  })
}
```

In the following, use R to answer the questions/assignments. Your solutions must be written up in a R Markdown (Rmd) file. This file must include your code and a (brief) comment for each question. For example, "The three companies with smalles profit are ..." or "The plot shows that ...".

An R markdown template file is given in the appendix that you may use as a starting point. 

You may load and use the following packages:

```{r}
library(tidyverse)
library(skimr)
```


# R - `r ctrA()` (12%)




The euclidean distance between two points $p = (p_1,p_2)$ and $q = (q_1,q_2)$ can be calculated using formula $$ d(p,q) = \sqrt{(p_1-q_1)^2 + (p_2-q_2)^2}.$$


## `r ctrQ(reset=T)`  

Calculate the distance between points $p = (10,10)$ and $q = (4,3)$ using the formula.

```{r, solution=TRUE}
p <- c(10,10)
q <- c(4,3)
sqrt((p[1] - q[1])^2 + (p[2] - q[2])^2)
```


<div class="solution">
The distance is `r sqrt((p[1] - q[1])^2 + (p[2] - q[2])^2)`.</div>

## `r ctrQ()`

Consider 4 points in a matrix (one in each row):

```{r}
p_mat <- matrix(c(0, 7, 8, 2, 10, 16, 8, 12), nrow = 4)
p_mat
```

The distance matrix of `p_mat` is a 4 times 4 matrix where entry (i,j) contains the distance from the point in row `i` to the point in row `j`. 

Calculate the distance matrix of `p_mat`.

```{r, solution=TRUE}
d_mat = matrix(NA, nrow = nrow(p_mat), ncol = nrow(p_mat))
for (i in 1:nrow(d_mat)) {
  for (j in 1:ncol(d_mat)) {
    # if (i>j) {d_mat[i,j] <- d_mat[j,i]; next}   # assume symmetric distances
    # if (!is.na(d_mat[i,j])) next          # value already calculated 
    if (i==j) {d_mat[i,j] <- 0; next}
    d_mat[i,j] <- sqrt((p_mat[i,1] - p_mat[j,1])^2 + (p_mat[i,2] - p_mat[j,2])^2)
  }
}
d_mat
```


<div class="solution">
The distance matrix is given above.</div>

## `r ctrQ()`

Create a function `calc_distances` with the following features (implement as many as you can):

* Takes a matrix `p_mat` with a point in each row as input argument.
* Takes two additional input arguments `from` and `to` with default values `1:nrow(p_mat)`
* Return the distance matrix with values calculated for rows in the `from` input argument and columns in the `to` input argument. The other entries equals `NA`.
* The function should work for different `p_mat` (you may assume that the matrix always have two columns).

You may test your code using:

```{r, eval=FALSE}
p_mat <- matrix(c(10, 9, 15, 15, 11, 19, 12, 11, 7, 15), nrow = 5)
calc_distances(p_mat)
calc_distances(p_mat, to = 3:4)
calc_distances(p_mat, from = c(1, nrow(p_mat)), to = 3:4)
```

```{r, solution=TRUE}
calc_distances <- function(p_mat, from = 1:nrow(p_mat), to = 1:nrow(p_mat)) {
  d_mat <- matrix(NA, nrow = nrow(p_mat), ncol = nrow(p_mat))
  ite <- expand_grid(from = from, to = to)
  for (r in 1:nrow(ite)) {
    i <- ite$from[r]
    j <- ite$to[r]
    if (!is.na(d_mat[i,j])) next          # value already calculated 
    if (i==j) {d_mat[i,j] <- 0; next}
    d_mat[i,j] <- sqrt((p_mat[i,1] - p_mat[j,1])^2 + (p_mat[i,2] - p_mat[j,2])^2)
  }
  return(d_mat)
}
p_mat <- matrix(c(10, 9, 15, 15, 11, 19, 12, 11, 7, 15), nrow = 5)
calc_distances(p_mat)
calc_distances(p_mat, to = 3:4)
calc_distances(p_mat, from = c(1, nrow(p_mat)), to = 3:4)
```


<div class="solution">
The function with test are given above.</div>


# R - `r ctrA()` (22%)



The dataset `companies`, given in the appendix, lists approx. 1000 of the world's biggest companies, measured by sales, profits, assets and market value. The column/variables are:

* `name`: the name of the company.
* `country`: the country the company is situated in.
* `category`: the products the company produces.
* `sales`: the amount of sales of the company in billion USD.
* `profits`: the profit of the company in billion USD.
* `assets`: the assets of the company in billion USD.
* `marketvalue`: the market value of the company in billion USD.

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (data frames) and answer the following questions.


## `r ctrQ(reset=T)`

Read the dataset file `companies.csv` into the dataset `companies`.

```{r, solution=TRUE}
companies <- read_csv("companies.csv")
```


<div class="solution">
The file is read using the code above.</div>

## `r ctrQ()`

How many different companies are we considering, how many different product categories and how many different countries? Hint: the *skimr* package might be useful. 

```{r, solution=TRUE}
skim(companies)
```


<div class="solution">
From the output we can see that there are 1002 different companies (one for each row) 27 different product categories and 47 different countries. </div>



## `r ctrQ()`

What are the 3 biggest companies with respect to market value?

```{r, solution=TRUE}
dat <- companies %>% 
  arrange(desc(marketvalue)) %>% 
  head(n = 3) %>% 
  print()
# or
# companies %>% 
#   slice_max(marketvalue, n = 3)
```


<div class="solution">
The 3 biggest companies are `r dat %>% pull(name)`.</div>




## `r ctrQ()`

For each country find the company with highest profit. What company has the highest profit in Denmark?

```{r, solution=TRUE}
dat <- companies %>% 
  group_by(country) %>% 
  slice_max(profits, n = 1) %>% 
  print()
```


<div class="solution">
The company with highest profit for each country is listed above. In Denmark the company is `r dat %>% filter(country == "Denmark") %>% pull(name)`.</div>



## `r ctrQ()`

Which 4 product categories have the highest total market value?

```{r, solution=TRUE}
dat <- companies %>% 
  group_by(category) %>% 
  summarise(marketvalue = sum(marketvalue)) %>% 
  slice_max(marketvalue, n = 4) %>% 
  print()
```


<div class="solution">
The 4 product categories that have the highest total market value are `r dat %>% pull(category)`.</div>



## `r ctrQ()`

Create a new data frame only containing rows from Denmark and with columns `name`, `category` and a column `value` which equals the sum of columns `profits`, `assets` and `marketvalue`. Which company have the lowest value?

```{r, solution=TRUE}
dat <- companies %>% 
  filter(country == "Denmark") %>% 
  mutate(value = profits + assets + marketvalue) %>% 
  select(name, category, value) %>% 
  print()
```


<div class="solution">
The companies can be seen above. The company with lowest value is `r dat %>% slice_min(value) %>% pull(name)`.</div>





# R - `r ctrA()` (16%)





Answer this assignment using the *ggplot2* package in *tidyverse* (you might need dplyr for preparing the datasets you want to plot). We work with the dataset `companies` from Assignment 2 which can be read using:

```{r}
companies <- read_csv("companies.csv")
```


## `r ctrQ(reset=T)`

Create a visualization showing the number of companies for each product category with the following features:

* Number of companies is represented using bars and sorted increasingly or decreasingly.
* Informative figure title and axis titles are given.
* The labels on the x-axis are rotated 90 degrees.

What product category has the lowest number of companies?

```{r, solution=TRUE}
companies %>% 
  count(category) %>% 
  ggplot(aes(x = reorder(category, -n), y = n)) +
  geom_col() +
  labs(title = "Number of companies in each product category",
       x = "Product category",
       y = "") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


<div class="solution">
Based on the plot, the lowest number of companies are in the _Trading companies_ category.</div>



## `r ctrQ()`

Consider product categories `Drugs & biotechnology` and `Media`. Create a visualization showing the profit given sales of each company with the following features:

* Different colors are used for each product category. 
* Informative figure title and axis titles are given.
* A trend line for each category is added using `geom_smooth`.

Based on the trend lines which product category gives the best profit? 

```{r, solution=TRUE}
companies %>% 
  filter(category %in% c("Drugs & biotechnology", "Media")) %>% 
  ggplot(aes(x = sales, y = profits, color = category)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Profit given sales",
       x = "Sales (billion USD)",
       y = "Profit (billion USD)",
       color = "Product Category") +
  theme(legend.position = "bottom")
```


<div class="solution">
Based on the plot the _Drugs & biotechnology_ product category gives the best profit.</div>


## `r ctrQ()`

Consider product categories `Banking` and `Aerospace & defense`. Let `ratio` denote a variable/column that equals profit divided by sales. Create a visualization showing the variation in `ratio` with the following features:

* Different colors are used for each product category. 
* Informative figure title and axis titles are given.

Based on the visualization comment on the variation and median. Which product category gives the highest ratio? 

```{r, solution=TRUE}
companies <- companies %>% 
  mutate(ratio = profits/sales) 

# Option 1 - Use a boxplot
companies %>% 
  filter(category %in% c("Banking", "Aerospace & defense")) %>% 
  ggplot(aes(y = ratio, color = category)) +
  geom_boxplot() +
  labs(title = "Ratio of profit/sales",
       x = "",
       y = "Profit/sales ratio",
       color = "Product Category") +
  theme(legend.position = "bottom")

# Option 2 - Use a density plot
companies %>% 
  filter(category %in% c("Banking", "Aerospace & defense")) %>% 
  ggplot(aes(x = ratio, fill = category)) +
  geom_density(alpha = 0.5) +
  labs(title = "Ratio of profit/sales",
       y = "Density",
       x = "Profit/sales ratio",
       fill = "Product Category") +
  theme(legend.position = "bottom")
```


<div class="solution">
Given the plot above, it can be seen that the highest ratio with respect to the mean and the median is for _Banking_. However, the variation is also highest here.  </div>




## `r ctrQ()`

The `continents` dataset given in the appendix matches countries to continents and contains two columns:

* `country`: the country.
* `continent`: the corresponding continent.

You can load the dataset using:

```{r}
continents <- read_csv("continents.csv")
```

Consider product categories `Banking`, `Aerospace & defense`, `Telecommunications services` and `Semiconductors`. Create a visualization showing assets given market value for each company with the following features (hint: you may need to do a mutating join):

* Two continents `Americas` and `Europe` are considered.
* Different colors are used for each continent. 
* A plot is given for each product category (facet).
* Informative figure title and axis titles are given.
* A trend line for each category is added using `geom_smooth(method = lm)`.

Based on the visualization consider the trend lines for _Banking_ and comment.

```{r, solution=TRUE, warning=FALSE}
companies %>% 
  left_join(continents) %>% 
  filter(continent %in% c("Americas", "Europe")) %>% 
  filter(category %in% c("Banking", "Aerospace & defense", 
                         "Telecommunications services", "Semiconductors")) %>% 
  ggplot(aes(x = marketvalue, y = assets, color = continent)) +
  geom_point() +
  geom_smooth(method = lm) +
  facet_wrap(vars(category), scales = "free") +
  labs(title = "Assets given market value",
       x = "Market value (billion USD)",
       y = "Assets (billion USD)",
       color = "Continent") +
  theme(legend.position = "bottom")
```


<div class="solution">
By considering the trend lines for _Banking_, it seems that if compare companies with the same marketvalue, then companies in Europe have more assets. </div>

