---
title: "TFA exam - Guiding solution"
author: "Lars Relund Nielsen"
---

<style>
.solution {
  background-color: aliceblue;
  line-height: 1.5;
  margin-top: 5px;
  margin-bottom: 15px;
  margin-right: 0px;
  margin-left: 0px;
  padding-top: 10px;
  padding-right: 0px;
  padding-bottom: 1px;
  padding-left: 5px;
  /*border: 3px solid green;*/
}

h2 {
  font-size: large;
}
</style>



```{r setup, include=FALSE, eval=T}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  eval = TRUE,
  message=FALSE)
options(width = 90, scipen = 10, digits = 2)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```
















# R - `r ctrA()` (55%)
 
Consider the dataset in the file `catch.csv` containing catches from rivers in Denmark. A row is given for each catch with information about weight, length, date, angler, place, method etc.

Some of the columns/variables are:

  - `Date`: Date of catch.
  - `Length`: Length of fish (cm).
  - `Weight`: Weight of fish (kilograms).
  - `Place`: Place of catch (Lower, Middle, Upper).
  - `Method`: Method used (Spin, Fly, Worm).
  - `Killed`: If the fish was killed.
  - `River`: River of catch.
  - `Species`: Species of fish.
  - `Name`: Name of angler.

The dataset can be read using:

```{r, warning=FALSE}
library(tidyverse)
dat <- read_csv("catch.csv") 
```

Moreover, consider the dataset in the file `weight.csv` containing weight estimates for a given length. A row is given for each length with information about weight, month, river, species etc.

Some of the columns/variables are:

  - `Length`: Length of fish (cm).
  - `Month`: Month considered.
  - `Avg`: Average weight of fish (kilograms).
  - `Lower`: Lower confidence interval weight (kilograms).
  - `Upper`: Uppper confidence interval weight (kilograms).

The dataset can be read using:

```{r, warning=FALSE}
library(tidyverse)
dat_weight <- read_csv("weight.csv")
res <- dat_weight |> 
   filter(River == "Karup", 
          Species == "Seatrout", 
          Length == 50, 
          Month == 5) |> 
   print()
```

That is, for instance, the average weight of a 50 cm seatrout in Karup River in May is `r res$Avg` kg with a 95% confidence interval of [`r res$Lower`, `r res$Upper`] kg.

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (tibbles/data frames), and answer/complete the following questions/tasks:


## `r ctrQ()`

Which method resulted in most fish caught in Skjern River?

```{r, solution=TRUE}
res <- dat |>
   filter(River == "Skjern") |> 
   group_by(Method) |> 
   summarize(
      n_caught = n()
   ) |> 
   arrange(desc(n_caught)) |>
   print()
```

<div class="solution">
The method that caught most fish in Skjern River is `r res$Method[1]` with `r res$n_caught[1]` fish caught.
</div>


## `r ctrQ()`

Which angler has caught most fish in total and in kilograms over all the years considered?

```{r, solution=TRUE}
res <- dat |>
   group_by(Name) |> 
   summarize(
      total_weight = sum(Weight, na.rm = TRUE),
      n_caught = n()
   ) |> 
   arrange(desc(total_weight)) |>
   print() 
res2 <- res |> 
   arrange(desc(n_caught)) |> 
   print()
```

<div class="solution">
The angler `r res$Name[1]` have caught and killed the highest total weight of `r round(res$total_weight[1], 2)` kilograms over all the years considered. The angler `r res2$Name[1]` have caught the highest number of fish of `r res2$n_caught[1]` over all the years considered.
</div>


## `r ctrQ()`

Calculate the condition factor for each fish using the formula `Weight*100000/Length^3`. 

```{r, solution=TRUE}
res <- dat |> 
   mutate(Condition = Weight*100000/Length^3) |> 
   print()
```

Is the average condition factor higher for salmon or seatrout?

```{r, solution=TRUE}
res1 <- res |> 
   group_by(Species) |>
   summarize(
      avg_Condition = mean(Condition, na.rm = TRUE)
   ) |>
   arrange(desc(avg_Condition)) |>
   print()
```

<div class="solution">
The average condition factor is highest for `r res1$Species[1]` with an average of `r round(res1$avg_Condition[1], 2)`.
</div>

Consider the months after April and fish with a weight above 50 cm. Do seatrout in Karup River have a higher average condition factor than seatrout in Skjern River?

```{r, solution=TRUE}
res2 <- res |> 
   filter(Species == "Seatrout", Month > 4, Length > 50) |> 
   group_by(River) |>
   summarize(
      n = n(),
      avg_Condition = mean(Condition, na.rm = TRUE)
   ) |>
   arrange(desc(avg_Condition)) |>
   print()
```

<div class="solution">
The condition factor seems to be the same.
</div>


## `r ctrQ()`

Are there more catches per day on average on weekdays or during weekends in Skjern River? 

```{r, solution=TRUE}
res <- dat |>
   filter(River == "Skjern") |> 
   mutate(day_type = if_else(Weekday %in% c("Saturday", "Sunday"), "Weekend", "Weekday")) |> 
   group_by(day_type) |> 
   summarize(
      n_caught = n(),
      n_days = n_distinct(Date),
      avg_catches_per_day = n_caught / n_days
   ) |> 
   ungroup() |> 
   print()
```

<div class="solution">
There are more catches per day on average on `r ifelse(res$avg_catches_per_day[res$day_type == "Weekend"] > res$avg_catches_per_day[res$day_type == "Weekday"], "weekends", "weekdays")` in Skjern River.
</div>


## `r ctrQ()`

Calculate the catch and release rate (number released / total number caught) for each river over all years. Have the catch and release rates increased over the years in Karup River?

```{r, solution=TRUE}
res <- dat |>
   group_by(River, Year) |> 
   summarize(
      n_caught = n(),
      n_released = sum(!Killed)
   ) |> 
   mutate(catch_release_rate = n_released / n_caught) |> 
   ungroup() |> 
   print()
res2 <- res |> 
   filter(River == "Karup") |> 
   select(Year, catch_release_rate) |>
   arrange(Year) |>
   print()
```


## `r ctrQ()`

Use the weight dataset to update missing weights in the catch dataset. That is, for each fish, with a missing weight, use the average weight from the weight dataset for the corresponding length, river, species and month. Hint: Before joining the datasets, you need to round the length to an integer.



```{r, solution=TRUE}
dat_full <- dat |>
   mutate(Length = round(Length)) |>
   left_join(dat_weight) |> 
   mutate(
      Weight = if_else(is.na(Weight), Avg, Weight)
   ) |> 
   print()
```



# R - `r ctrA()` (45%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you may need *dplyr* for preparing the datasets you want to plot). We use the same dataset as given in Assignment 1 if not specified.


## `r ctrQ()`

Create a visualization showing the average weight of seatrout in Karup River over the months with the following features:

* Use the dataset from Assignment 1 in which missing weights have been updated if possible.
* Bars with the fill colour `darkred` are used.
* An informative figure title and axis titles are given.

Which month has the highest average weight?

```{r, solution=TRUE}
dat_full |> 
   filter(River == "Karup", Species == "Seatrout") |>
   group_by(Month) |> 
   summarize(
      avg_weight = mean(Weight, na.rm = TRUE)
   ) |>
   ggplot(aes(x = Month, y = avg_weight)) +
   geom_col(fill = "darkred") +
   labs(
      title = "Average weight of seatrout in Karup River given month",
      x = "Month",
      y = "Average weight (kg)"
   )
```

<div class="solution">
The month with the highest average weight is May.
</div>





## `r ctrQ()`

Create a visualization showing the relationship between length and weight for all fish caught with the following features:

* Different colours are used for each species.
* Points are used to show the individual catches with some transparency (alpha).
* A trend line is added for each species.
* An informative figure title and axis titles are given.
* The legend is placed at the bottom of the figure.

Are salmon generally heavier than seatrout for a given length? Can salmon reach a higher weight than seatrout?

```{r, solution=TRUE}
dat |>
   filter(Weight > 0, Length > 0) |> 
   ggplot(aes(x = Length, y = Weight, color = Species)) +
   geom_point(alpha = 0.5, size = 0.5) +
   geom_smooth(se = FALSE) +
   labs(
      title = "Relationship between Length and Weight",
      x = "Length (cm)",
      y = "Weight (kg)"
   ) + 
   theme(legend.position = "bottom")
```

<div class="solution">
Salmon are generally NOT heavier than seatrout for a given length as seen from the trend lines. Moreover, salmon can reach a bigger weight than seatrout as seen from the points.
</div>


## `r ctrQ()`

Consider the weight dataset. Create a visualization showing the weight for a given length with the following features:

* Only Skjern River is considered.
* Lines are used to show the average weight given length for each species.
* Different colours are used for each month.
* A subplot is made for each species.
* An informative figure title and axis titles are given.
* The legend is placed at the bottom of the figure.

In which month is the weight lowest for seatrout?

```{r, solution=TRUE}
dat_weight |> 
   filter(River == "Skjern" ) |> 
   ggplot(aes(x = Length, y = Avg, color = factor(Month))) +
   geom_line() +
   facet_wrap(~Species) +
   labs(
      title = "Weight given Length",
      x = "Length (cm)",
      y = "Weight (kg)",
      color = "Month"
   ) +
   theme(legend.position = "bottom") 
```

<div class="solution">
The weight is lowest for seatrout in October (month 10).
</div>


## `r ctrQ()`

Create a visualization showing the percentage of fish over 59 cm killed over the years with the following features:

* Bars are used to visualize the percentage of fish killed.
* Different fill colours are used for each river (use `position = "dodge2"` to place bars beside each other).
* An informative figure title and axis titles are given.
* The legend is placed at the bottom of the figure.

In which river do most anglers catch and release fish (i.e. the lowest percentage killed)?

```{r, solution=TRUE}
dat |>
   filter(Length >= 60) |> 
   group_by(River, Year) |> 
   summarize(
      n_caught = n(),
      n_killed = sum(Killed),
      pct_killed = n_killed / n_caught
   ) |> 
   ungroup() |>
   ggplot(aes(x = factor(Year), y = pct_killed, fill = River)) +
   geom_col(position = "dodge2") +
   labs(
      title = "Percentage of fish killed",
      x = "Year",
      y = "Percentage killed"
   ) +
   theme(legend.position = "bottom")
```

<div class="solution">
The river with the most catch and release is Skjern River.
</div>


## `r ctrQ()`

Consider seatrout. For each river, year, and month, calculate the number of catches. Next, calculate the catch percentage for each month (number caught per month / total number caught per year). 

```{r, solution=TRUE}
res <- dat |> 
   filter(Species == "Seatrout") |>
   group_by(River, Year, Month) |> 
   summarize(
      n_caught = n()
   ) |> 
   group_by(River, Year) |> 
   mutate(
      catch_pct = n_caught / sum(n_caught)
   ) |> 
   ungroup() |> 
   print()
```

Create a visualization showing the variation in these percentages for each month with the following features:

* A boxplot is used to show the variation of the catch percentages.
* A subplot is made for each River.
* An informative figure title and axis titles are given.

Comment on the results for July.

```{r, solution=TRUE}
res |> 
   ggplot(aes(x = factor(Month), y = catch_pct)) +
   geom_boxplot() +
   facet_wrap(~River) +
   labs(
      title = "Variation of catch percentages for Seatrout",
      x = "Month",
      y = "Catch percentage"
   ) +
   theme(legend.position = "bottom")
```

<div class="solution">
In July the catch percentage seems to be higher in Karup River compared to Skjern River (the median is higher). However, there is a larger variation in Skjern River (over years).
</div>


