---
title: "Teaching Assistens - Tasks"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Tasks

   - Coordinate and teach study cafés.
   - Read syllabus and calculate and check each week's exercises.
   - Provide help for project reports (week 41 and 49).
   - Report errors, typos etc. (you may use GitHub).
   - Teach a lecture (week 42). Use it as independent teaching in PhDPlanner. 
   - Check and answer questions at the [course help page](https://github.com/bss-osca/tfa/issues). Do NOT accept and give answers using mail. Instead kindly direct them to the help page.  

Expect that more help is needed in the weeks where the students do project reports. 

### Note 

- The study café is moved in Week 40 so two study cafés in Week 41. This gives you better time to help with the project report.
- You can always check your schedule at [https://timetable.au.dk/schedule](https://timetable.au.dk/schedule).
- The course schedule can be found [here](https://docs.google.com/spreadsheets/d/e/2PACX-1vRka2NKE77BvtIoJc1GW-Dr_oABzQdQhMC94KpPjW7DpSTnepN9RV3m6w1ILKe3YoQwUj0C6JAe3K1T/pubhtml?gid=1591856300)
- The learning paths are at https://bss-osca.github.io/tfa/. Use hypothes.is to annotate/notify me about errors/typos etc. Join the group using this [link](https://hypothes.is/groups/b7oEZy31/tas).

```{r copy to docs, include=FALSE}
file.copy(list.files(pattern = ".html"), "../docs/admin/", overwrite = T)
```