---
title: "TFA exam - R part"
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90, scipen=999)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```




 

# R - `r ctrA()` (14%)

## `r ctrQ()` 
 
Make functions:

  - `SecToMin` which takes an input argument `sec` in seconds and return the number converted to minutes.
  - `SecToHours` which takes an input argument `sec` in seconds and return the number converted to hours.
  - `MinToSec` which takes an input argument `min` in minutes and return the number converted to seconds.
  - `MinToHours` which takes an input argument `min` in minutes and return the number converted to hours.
  - `HoursToMin` which takes an input argument `hours` in hours and return the number converted to minutes.
  - `HoursToSec` which takes an input argument `hours` in hours and return the number converted to seconds.

All numbers may be decimal numbers, e.g. 90 seconds is 1.5 minutes and 1.5 hours is 90 minutes. 

```{r}
# your code here
```




## `r ctrQ()` 

Make a function `ConvertTime` which takes two input arguments:

  - `val` A number.
  - `unit` A string that can take values "sec", "min" and "hours".

The function should return `val` converted to seconds, minutes and hours with features:

  - works for all possible values for `unit`,
  - uses the functions in Question 1,
  - returns a vector with 3 numbers (seconds, minutes and hours) or `NA` if `unit` does not equals "sec", "min" or "hours".



```{r}
# your code here
```




# R - `r ctrA()` (18%)



The dataset in the file `election.csv` lists election votes for municipal elections in Denmark. The column/variables are:

* `area`: municipality,
* `party`: political party,
* `year`: election year,
* `validVotes`: Number of (valid) votes, 
* `personalVotes`: total number of personal votes, 
* `listedMen`: men listed,
* `listedWomen`: women listed, 
* `electedMen`: elected men,
* `electedWomen`: elected women.

You can read the dataset file into the dataset `dat` using

```{r}
library(tidyverse)
dat <- read_csv("elections.csv")
```

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (data frames) and answer the following questions.


## `r ctrQ()`

How many different municipalities were there in Denmark in 2017?

```{r}
# your code here
```

[Your comment here]



## `r ctrQ()`

In which election year was the total number of votes highest?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()` 

Consider the 2017 election. How many votes and elected candidates did the different parties get in total (sort your result descending by votes)?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Calculate the percentage of females listed for each year. Are there on average more women than men listed? Is there any tendency over the years?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Consider the 2017 election. In which municipality did the Liberal Democratic Party get the highest percentage of votes? 

```{r}
# your code here
```

[Your comment here]



























# R - `r ctrA()` (18%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you might need dplyr for preparing the datasets you want to plot). We work with the dataset from Assignment 2 which can be read using:

```{r}
library(tidyverse)
dat <- read_csv("elections.csv")
```


## `r ctrQ()`

Create a visualization showing the total number of votes for each election year with the following features:

* Number of votes is represented using columns.
* Columns are filled with colours for each party.
* Informative figure title and axis titles are given.

Which year had the lowest number of votes?

```{r}
# your code here
```

[Your comment here]

## `r ctrQ()`

Create a visualization showing the relative number of elected women in each municipality with the following features:

* The relative number of elected women is shown using columns.
* Municipalities are rotated 90 degrees on the x-axis.
* The columns are sorted increasing.
* Informative figure title and axis titles are given.

Which municipality has the lowest/highest percentage of women elected?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Create a visualization showing the elected number of candidates as a function of election year with the following features:

* We consider only municipalities Herning, Copenhagen and Aarhus.
* We consider only the Social Democratic Party, Conservative Peoples Party and Liberal Democratic Party.
* The plot is divided using `area` (facet).
* Different colours are used for each party.
* Informative figure title and axis titles are given.
* Data points are given in the plot.
* A line for each party is added.

Consider the Social Democratic Party in Copenhagen. Based on the plot do the elected number of candidates increase or decrease over the election years?

```{r}
# your code here
```

[Your comment here]

## `r ctrQ()`

Create a visualization showing the relative number of women elected compared to listed (e.g. if listed is 12 and elected is 4 then the number is 4/12) with the following features:

* Columns are used for each party.
* Colours are used to identify the party.
* A sub-plot is given for each year.
* Informative figure title and axis titles are given.

Which party seems to get the highest number of the listed women elected? 

```{r}
# your code here
```

[Your comment here]



## `r ctrQ()`

Create a visualization showing the variance in the relative number of personal votes compared to valid votes (i.e. divide the two numbers) with the following features:

* We consider only municipalities Herning, Copenhagen and Aarhus.
* Variation is shown using a box-plot for each municipality.
* Informative figure title and axis titles are given.

Which municipality has the highest median?

```{r}
# your code here
```

[Your comment here]





