---
title: "TFA exam - R part"
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)

a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}
```




In the following, use R to answer the questions/assignments. Your solutions must be written up in a R Markdown (Rmd) file. This file must include your code and a (brief) comment for each question. For example, "The three companies with smalles profit are ..." or "The plot shows that ...".

An R markdown template file is given in the appendix that you may use as a starting point. 

You may load and use the following packages:

```{r}
library(tidyverse)
library(skimr)
```


# R - `r ctrA()` (12%)




The euclidean distance between two points $p = (p_1,p_2)$ and $q = (q_1,q_2)$ can be calculated using formula $$ d(p,q) = \sqrt{(p_1-q_1)^2 + (p_2-q_2)^2}.$$


## `r ctrQ(reset=T)`  

Calculate the distance between points $p = (10,10)$ and $q = (4,3)$ using the formula.

```{r}
# your code here
```

[Your comment here]

## `r ctrQ()`

Consider 4 points in a matrix (one in each row):

```{r}
p_mat <- matrix(c(0, 7, 8, 2, 10, 16, 8, 12), nrow = 4)
p_mat
```

The distance matrix of `p_mat` is a 4 times 4 matrix where entry (i,j) contains the distance from the point in row `i` to the point in row `j`. 

Calculate the distance matrix of `p_mat`.

```{r}
# your code here
```

[Your comment here]

## `r ctrQ()`

Create a function `calc_distances` with the following features (implement as many as you can):

* Takes a matrix `p_mat` with a point in each row as input argument.
* Takes two additional input arguments `from` and `to` with default values `1:nrow(p_mat)`
* Return the distance matrix with values calculated for rows in the `from` input argument and columns in the `to` input argument. The other entries equals `NA`.
* The function should work for different `p_mat` (you may assume that the matrix always have two columns).

You may test your code using:

```{r, eval=FALSE}
p_mat <- matrix(c(10, 9, 15, 15, 11, 19, 12, 11, 7, 15), nrow = 5)
calc_distances(p_mat)
calc_distances(p_mat, to = 3:4)
calc_distances(p_mat, from = c(1, nrow(p_mat)), to = 3:4)
```

```{r}
# your code here
```

[Your comment here]


# R - `r ctrA()` (22%)



The dataset `companies`, given in the appendix, lists approx. 1000 of the world's biggest companies, measured by sales, profits, assets and market value. The column/variables are:

* `name`: the name of the company.
* `country`: the country the company is situated in.
* `category`: the products the company produces.
* `sales`: the amount of sales of the company in billion USD.
* `profits`: the profit of the company in billion USD.
* `assets`: the assets of the company in billion USD.
* `marketvalue`: the market value of the company in billion USD.

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (data frames) and answer the following questions.


## `r ctrQ(reset=T)`

Read the dataset file `companies.csv` into the dataset `companies`.

```{r}
# your code here
```

[Your comment here]

## `r ctrQ()`

How many different companies are we considering, how many different product categories and how many different countries? Hint: the *skimr* package might be useful. 

```{r}
# your code here
```

[Your comment here]



## `r ctrQ()`

What are the 3 biggest companies with respect to market value?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

For each country find the company with highest profit. What company has the highest profit in Denmark?

```{r}
# your code here
```

[Your comment here]



## `r ctrQ()`

Which 4 product categories have the highest total market value?

```{r}
# your code here
```

[Your comment here]



## `r ctrQ()`

Create a new data frame only containing rows from Denmark and with columns `name`, `category` and a column `value` which equals the sum of columns `profits`, `assets` and `marketvalue`. Which company have the lowest value?

```{r}
# your code here
```

[Your comment here]





# R - `r ctrA()` (16%)





Answer this assignment using the *ggplot2* package in *tidyverse* (you might need dplyr for preparing the datasets you want to plot). We work with the dataset `companies` from Assignment 2 which can be read using:

```{r}
companies <- read_csv("companies.csv")
```


## `r ctrQ(reset=T)`

Create a visualization showing the number of companies for each product category with the following features:

* Number of companies is represented using bars and sorted increasingly or decreasingly.
* Informative figure title and axis titles are given.
* The labels on the x-axis are rotated 90 degrees.

What product category has the lowest number of companies?

```{r}
# your code here
```

[Your comment here]



## `r ctrQ()`

Consider product categories `Drugs & biotechnology` and `Media`. Create a visualization showing the profit given sales of each company with the following features:

* Different colors are used for each product category. 
* Informative figure title and axis titles are given.
* A trend line for each category is added using `geom_smooth`.

Based on the trend lines which product category gives the best profit? 

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Consider product categories `Banking` and `Aerospace & defense`. Let `ratio` denote a variable/column that equals profit divided by sales. Create a visualization showing the variation in `ratio` with the following features:

* Different colors are used for each product category. 
* Informative figure title and axis titles are given.

Based on the visualization comment on the variation and median. Which product category gives the highest ratio? 

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

The `continents` dataset given in the appendix matches countries to continents and contains two columns:

* `country`: the country.
* `continent`: the corresponding continent.

You can load the dataset using:

```{r}
continents <- read_csv("continents.csv")
```

Consider product categories `Banking`, `Aerospace & defense`, `Telecommunications services` and `Semiconductors`. Create a visualization showing assets given market value for each company with the following features (hint: you may need to do a mutating join):

* Two continents `Americas` and `Europe` are considered.
* Different colors are used for each continent. 
* A plot is given for each product category (facet).
* Informative figure title and axis titles are given.
* A trend line for each category is added using `geom_smooth(method = lm)`.

Based on the visualization consider the trend lines for _Banking_ and comment.

```{r}
# your code here
```

[Your comment here]

