---
title: "Programming"
author: "Lars Relund Nielsen"
output:
  xaringan::moon_reader:
    css: "./libs/slides.css"
    lib_dir: libs
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r, child = "../setup.Rmd", echo=FALSE}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# use xaringan::inf_mr(cast_from = "..") to preview
library(tidyverse)
```


## Comparison/relational operators

- `<` for less than, `>` for greater than,
- `<=` for less than or equal to, `>=` for greater than or equal to,
- `==` for equal to each other (and not `=` which is typically used for assignment!),
- `!=` not equal to each other.

```{r l1, eval=FALSE}
x <- TRUE   # you may use x <- T 
y <- FALSE  # you may use y <- F
x == y
```
--
```{r, ref.label="l1", echo=FALSE}
```
--
```{r l2, eval=FALSE}
z <- c(1, -1, 3)
z > 0   # vector comparison
```
--
```{r, ref.label="l2", echo=FALSE}
```

---

## A vector of booleans  

```{r l3, eval=FALSE}
z <- c(1, -1, 3)
all(z > 0)
any(z > 0)
```
--
```{r, ref.label="l3", echo=FALSE}
```
--
.your-turn[
Your turn - Try to guess the output
```{r, eval=FALSE}
x <- c(1, 2, 45, NA, -1)
x > 0
x > -100
any(x > 0)
all(x > -100)
```
]

```{r echo=FALSE}
countdown(minutes = 1)
```

---

## Logical operators 

- `&` and,
- `|` or,
- `!` not.
- If you use `&&` and `||` only the first element in vectors are compared. 

```{r}
x <- 4
y <- "Lars"
z <- NA
x == 4 & y == "lars"
```
--
```{r l4, eval=FALSE}
x > 4 | y != "lars"
```
--
```{r, ref.label="l4", echo=FALSE}
```

---
Remember parenthesis 
```{r l5, eval=FALSE}
(x == 4 | y != "Lars") & !is.na(z)
x == 4 | (y != "Lars" & !is.na(z))
```
--
```{r, ref.label="l5", echo=FALSE}
```
--

Logical operators applied to vectors
```{r}
v1 <- 1:4
v2 <- c(2,-1,7,9)
v1 > 1 & v2 < 3
```
--
```{r l6, eval=FALSE}
v1 > 1 && v2 < 3
```
--
```{r, ref.label="l6", echo=FALSE}
```

---

## Conditional Statements (a single boolean)

Use of `if` and `while` (a single boolean)

```{r}
x <- 5
if (x < 3) {
  cat("first")
} else if (x > 4) {
  cat("next")
} else {
  cat("last")
}
```

```{r}
x <- 1
while(x < 5) {
  x <- x + 1
}
```

---

## Conditional Statements (vector of booleans)

Use of `if_else` (a vectorized alternative)

```{r}
library(tidyverse)
## using if and for
res <- rep("", length(x))
for (i in seq_along(x)) {
  if (is.na(x[i])) res[i] <- "missing"
  else if (x[i] < 0) res[i] <- "negative"
  else res[i] <- "positive"
}

## using if_else
x <- c(-5:5, NA)
res <- if_else(x < 0, "negative (true)", "positive (false)", "missing (na)")
res
```

---

## Loops

Use `for` or `while`:

```{r}
v <- c(2,4,9,0)  # values we want to iterate over
res <- rep(NA,4) # vector to store results in
for (idx in 1:length(v)) {
  res[idx] <- 6 * v[idx] + 9
}
res

x <- 1; res <- x
while(x < 5) {
  x <- x + 1
  res <- res + x 
}
res
sum(1:5)
```

---

## Use a vectorized alternative

Loops in R may be slow. However, not if you follow some golden rules:

* Do not use a loop when a vectorized alternative exists.
* Do not grow objects (via `c`, `cbind`, etc) during the loop - R has to create a new object and copy across the information just to add a new element or row/column. Instead, allocate an object to hold the results and fill it in during the loop.

```{r}
6 * v + 9
sum(1:5)
```

---

## Nested loops

```{r, eval=FALSE}
for (i in 1:3) {
  for (j in 1:2) {
    # code here
  }
}
```

You can replace nested loops with a single loop by using `expand_grid`:

```{r, eval=FALSE}
library(tidyverse)
ite <- expand_grid(i = 1:3, j=1:2)
for (r in 1:nrow(ite)) { # iterate over rows
   i <- ite$i[r]
   j <- ite$j[r]
   # code here
}
```

This way of looping is a more flexible approach since you can nest more loops by adding more columns to `ite` and add different values in each column.

---

## Your Turn

.your-turn[
- Given a dice and a 52-card deck, use a for loop to write all the possible combinations as strings in the form `Dice = [value] and card = [value]`. Hint: `cat` can be used to concatenate the strings and start with vectors: 
  ```{r}
  dice <- 1:6
  card_suit <- c("clubs", "diamonds", "hearts", "spades")
  card_number <- c("ace", 2:10, "jack", "queen", "king")
  ```
]


```{r echo=FALSE}
countdown(minutes = 5, top = "0")
```

```{r, include=FALSE}
library(tidyverse)
dice <- 1:6
card_suit <- c("clubs", "diamonds", "hearts", "spades")
card_number <- c("ace", 2:10, "jack", "queen", "king")
dat <- expand_grid(Dice = dice, Card_suit = card_suit, Card_number = card_number)
for (r in 1:nrow(dat)) {
  cat("Dice = ", dat$Dice[r], " and ", dat$Card_number[r], " (", dat$Card_suit[r], ").\n", sep="")
} 
```

---

class: middle

> To understand computations in R, two slogans are helpful:
>
>    Everything that exists is an object.
>
>    Everything that happens is a function call.
>
>
>    .right[\- _John Chambers_]

---

## Functions

- Functions are often used to encapsulate a sequence of expressions that need to be executed numerous times, perhaps under slightly different conditions. 
<!-- - In programming, functional programming is a programming paradigm, a style how code is written. Rather than repeating code, functions and control structures allow one to build code in blocks.  -->
- Functions are (often) verbs, followed by what they will be applied to in parentheses.
- Functions can be defined using the `function()` directive. 
```{r}
loop_output <- function(dat, rows = 1:2, cols = 2:3, ...) {
  ite <- expand_grid(r = rows, c = cols)
  for (i in 1:nrow(ite)) {
    r <- ite$r[i]
    c <- ite$c[i]
    cat("Row ", r, " column ", c, " = ", dat[r,c], ".\n", ...)
  } 
  return("Finished!")
}
```
- Functions always return the last expression evaluated in the function body or when you use the `return` flow control statement (good coding practice).

---

## Functions - Arguments

```{r}
loop_output(mtcars)  # default values
loop_output(mtcars, rows = c(3,1), cols = 6)  # specify values
```

---

## Functions - The `...` argument

```{r}
loop_output(mtcars, rows = c(3,1), cols = 6)  # specify values
loop_output(mtcars, rows = c(3,1), cols = 6, sep = "")  # use the ... argument
```

---

## Functions - Document your functions 

```{r}

#' Loop through some values in a data frame
#'
#' @param dat A data frame
#' @param rows Rows to loop through.
#' @param cols Columns to loop through.
#' @param ... Arguments passed to `cat`.
#'
#' @return A string.
#' @examples
#' loop_output(mtcars, rows = c(3,1), cols = 6)
loop_output <- function(dat, rows = 1:2, cols = 2:3, ...) {
  ite <- expand_grid(r = rows, c = cols)
  for (i in 1:nrow(ite)) {
    r <- ite$r[i]
    c <- ite$c[i]
    cat("Row ", r, " column ", c, " = ", dat[r,c], ".\n", ...)
  } 
  return("Finished!")
}
```


---

## Your Turn

.your-turn[
- Create a function that given some dice numbers and some cards from a 52-card deck (e.g. 1-spade), write out all the possible combinations as strings in the form `Dice = [value] and card = [value]`. 
- Moreover, if the dice numbers contains the number 2, function must return true; otherwise, false.
]

```{r echo=FALSE}
countdown(minutes = 5)
```

```{r, include=FALSE}
get_combinations <- function(dice, card) {
  ite <- expand_grid(d = dice, c = card)
  found <- FALSE
  for (i in 1:nrow(ite)) {
    d <- ite$d[i]
    c <- ite$c[i]
    cat("Dice ", d, " and card  = ", c, ".\n", sep = "")
    if (d == 2) found = TRUE
  } 
  return(found)
}
get_combinations(dice = c(2,3), card = c("2-spade", "ace-dimond"))
get_combinations(dice = c(3,1), card = c("10-heart", "king-dimond"))
```





```{r copy to docs, include=FALSE}
file.copy(list.files(pattern = ".html"), "../../docs/slides/", overwrite = T)
file.copy("../slides.css", "./libs/", overwrite = T)
file.copy("libs", "../../docs/slides/", recursive = T)
file.copy("img", "../../docs/slides/", recursive = T)
```

