---
title: "TFA exam"
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90, scipen = 1, digits = 2)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```












# R - `r ctrA()` (40%)
 
Consider the dataset in the file `jobs.csv` with engine maintenance jobs for a group of ships. 

The dataset contains a row for each item used. The columns/variables are:

  - `ship`: The ship considered.
  - `job_id`: Maintenance job id. A job is a collection of items replaced.
  - `job_desc`: Job description.
  - `item_id`: Item id.
  - `item_name`: Item name.
  - `item_quantity`: Number of items used.
  - `item_manufaturer`: Item manufacturer.
  - `component_id`: Engine component id.
  - `component_desc`: Engine component description.
  - `done_date`: Date the job finished.
  - `year`: Year of done date.
  - `days`: Days since the item was last used for maintenance on the ship.

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (tibbles/data frames) and answer/complete the following questions/tasks.




## `r ctrQ()`

Find the delimiter used in the csv file and load the file into a tibble/data frame called `dat`.

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Provide a short overview of the data. How many rows are there in the data, what is the number of different ships, and what is the range of dates?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

What is the total number of different jobs, number of different jobs per ship, and average number of jobs per ship? Hint: The function `n_distinct` may be used to find distinct values within a group.

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

What is the minimum, average, and  maximum number of *different* items considered at each job? What is the minimum, average, and  maximum number of items used for maintenance at each job?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

It seems that some items considered in a job are not used anyway. Which ship has most jobs where an item should have been used, but has not (quantity is zero)?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Which item is the most used one for maintenance?

```{r}
# your code here
```

[Your comment here]






# R - `r ctrA()` (40%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you may need *dplyr* for preparing the datasets you want to plot). We work with the dataset from Assignment 1, which can be read using:


```{r}
library(tidyverse)
dat <- read_csv("jobs.csv")
```




## `r ctrQ()`

For each year and job id, identify the ship the job was done on. Hint: The `first` function may be used to select the first item within a group. Use this to create a visualization showing the number of maintenance jobs for each year with the following features:

   * Bars are used for each year.
   * Fill colors are used to identify the ship.
   * Legends are put at the bottom of the plot.
   * Informative figure title and axis titles are given.

Which year had the most jobs?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Create a visualization showing the number of different items used for each ship with the following features:

* The numbers are shown using columns.
* Reorder the columns so they increase along the x-axis.
* Informative figure title and axis titles are given.

Which ship uses most items? 

```{r}
# your code here
```

[Your comment here]



## `r ctrQ()`

Create a visualization showing the number of items used as a function of done date with the following features:

* The numbers are shown using columns with a fixed line width of 1 and a blue color. 
* The plot is divided using `ship` (facet). Hint: You may use `scales = "free_y"`. 
* Informative figure title and axis titles are given.

Which ship has most items used at a specific date?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Consider two items with id:

```{r}
items <- c("601.004.006", "601.026.128")
```

Create a visualization showing the variation of demand (item quantity) of each item with the following features:

* A density is plotted for each item with a fixed bandwidth (`bw`) of 1 and transparency (`alpha`) of 0.5. 
* Item name is used as fill. 
* The plot is divided using `year` (facet).
* Informative figure title and axis titles are given.

Over the years, is it consistent that on average the demand of one of the items is higher than the other?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Consider two items with id:

```{r}
items <- c("601.004.006", "601.026.128")
```

Create a visualization showing the variation of days since last used for maintenance given an item with the following features:

* A violin is used to plot days since last used given an item. 
* Item name is used as fill. 
* Informative figure title and axis titles are given.

Comment on the plot.

```{r}
# your code here
```

[Your comment here]





# R - `r ctrA()` (20%)

Create a function `process` with the following features:

   * The arguments are:
     - `v`: A vector with numbers (may contain missing values,`NA`s).
     - `find`: A number to search for. Default value is `NULL` indicating not to search for a number.
     - `replace`: A number. If `find` is a number, then replace all occurrences of `find` in the vector with `replace` (if any). Default value is `NULL` indicating not to replace anything.
   * The function returns a list with 
     - the original vector `v`, 
     - the length of `v`, 
     - the number of missing values,
     - if find is a number, the number of times the number is found, and
     - if replace is a number, the new vector with the replaced numbers.
   * Document your function. 

```{r}
# your code here
```





