---
title: "TFA reexam - R part"
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90)

a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}
```




In the following, use R to answer the questions/assignments. Your solutions must be written in a R Markdown (Rmd) file. This file must include your code and a (brief) comment for each question. For example, "The three companies with smallest profit are ..." or "The plot shows that ...".

An R markdown template file is given in the appendix that you may use as a starting point. 

You may load and use the following packages:

```{r}
library(tidyverse)
library(skimr)
```


# R - `r ctrA()` (10%)

Consider the vector:

```{r}
v <- c(9, 19, 2, 8, NA, 12, 9, 23, NA, 34)
v
```

Use R to answer the following questions.

## `r ctrQ()`

Is any of the entries in `v` below or equal to 2?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()` 

Is all of the entries in `v` above or equal to 2?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Does `v` have missing values?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()` 

Which entries in `v` are above 10? You must return the indices, e.g. the index of `v[3]` is 3.

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()` 

Create a vector `res` where `res[i]` is equal to `v[i]` if `v[i]` is less than 10 and otherwise zero (also if `v[i]` is `NA`).  

```{r}
# your code here
```

[Your comment here]




# R - `r ctrA()` (20%)

The dataset `titanic`, given in the appendix, lists approx. 1300 passengers on Titanic. The column/variables are:

* `pclass`: Passenger class (1 = 1st; 2 = 2nd; 3 = 3rd).
* `survived`: Survival (0 = No; 1 = Yes).
* `name`: Name.
* `sex`: Sex.
* `age`: Age.
* `fare`: Passenger Fare.
* `cabin`: Cabin number.
* `embarked`: Port of embarkation (C = Cherbourg; Q = Queenstown; S = Southampton).
* `boat`: Lifeboat number.

You can read the dataset file `titanic.csv` into the dataset `dat` using

```{r}
dat <- read_csv("titanic.csv")
```

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (data frames) and answer the following questions.



## `r ctrQ(reset=T)`

Create a new column named `male` which is true if the person is a male. 

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()` 

How many persons are we considering, how many men (in percentage) and how many survived? 


```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

How many of the females survived in percent (and how many males)?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

Define children as people with age below 19. How many children survived?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Did relatively more people survive at first class compared to third class?

```{r}
# your code here
```

[Your comment here]




## `r ctrQ()`

How many persons that entered a lifeboat did die in percent? 

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

How many persons with `Hansen` in their name survived?

```{r}
# your code here
```

[Your comment here]





# R - `r ctrA()` (20%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you might need dplyr for preparing the datasets you want to plot). We work with the dataset `titanic` from Assignment 2 which can be read using:

```{r}
dat <- read_csv("titanic.csv")
```


## `r ctrQ(reset=T)`

Create a visualization showing the number of persons embarking the different harbors with the following features:

* Number of persons is represented using bars.
* Informative figure title and axis titles are given.
* The labels on the x-axis are the harbor names (not abbreviations).

What harbor was the main departure harbor?

```{r}
# your code here
```

[Your comment here]

## `r ctrQ()`

Create a visualization showing the age distribution of the persons with the following features:

* Number of persons is represented using a histogram.
* Informative figure title and axis titles are given.

Where there most people on board in their twenties or thirties?



```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Create a visualization showing the fare as a function of age with the following features:

* Different colors are used for each sex. 
* Informative figure title and axis titles are given.
* A trend line for each sex is added using `geom_smooth`.

Based on the trend lines, do females in general pay more for a ticket?

```{r}
# your code here
```

[Your comment here]

## `r ctrQ()`

Create a visualization showing the survival rate for each passenger class with the following features:

* Bars are used for each passenger class. 
* All bars have same height (100 %).
* Colors are used to identify who survived and did not survived.
* Informative figure title and axis titles are given.

Is the survival rate different on first and third class?

```{r}
# your code here
```

[Your comment here]



## `r ctrQ()`

Let column `level` denote the the first letter in `cabin`. Create a visualization showing the variance in fare prices with the following features:

* Ignore rows with missing `level`.
* Variation is shown using a boxplot. 
* Informative figure title and axis titles are given.

Is the fare price different at the B and C level?

```{r}
# your code here
```

[Your comment here]









