---
title: "TFA exam - Guiding solution"
author: "Lars Relund Nielsen"
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90, scipen = 1, digits = 2)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```












# R - `r ctrA()` (40%)
 
Consider the dataset in the file `jobs.csv` with engine maintenance jobs for a group of ships. 

The dataset contains a row for each item used. The columns/variables are:

  - `ship`: The ship considered.
  - `job_id`: Maintenance job id. A job is a collection of items replaced.
  - `job_desc`: Job description.
  - `item_id`: Item id.
  - `item_name`: Item name.
  - `item_quantity`: Number of items used.
  - `item_manufaturer`: Item manufacturer.
  - `component_id`: Engine component id.
  - `component_desc`: Engine component description.
  - `done_date`: Date the job finished.
  - `year`: Year of done date.
  - `days`: Days since the item was last used for maintenance on the ship.

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (tibbles/data frames) and answer/complete the following questions/tasks.




## `r ctrQ()`

Find the delimiter used in the csv file and load the file into a tibble/data frame called `dat`.

```{r, solution=TRUE}
# load packages
library(tidyverse)
library(skimr)

lines <- read_lines("jobs.csv", n_max = 3)
lines
dat <- read_csv("jobs.csv")
```

<div class="solution">
It can be seen that a comma (`,`) is used a delimiter. That is, the dataset is loaded using `read_csv`.
</div>




## `r ctrQ()`

Provide a short overview of the data. How many rows are there in the data, what is the number of different ships, and what is the range of dates?

```{r, solution=TRUE, warning=FALSE}
skim(dat)
```

<div class="solution">
We have a total of `r dat %>% nrow()` rows/observations for 9 ships over a period from 2015-01-24 to 2020-04-21.
</div>




## `r ctrQ()`

What is the total number of different jobs, number of different jobs per ship, and average number of jobs per ship? Hint: The function `n_distinct` may be used to find distinct values within a group.

```{r, solution=TRUE}
res1 <- dat %>% 
   distinct(job_id) %>% 
   nrow() %>% 
   print()

res2 <- dat %>% 
   group_by(ship) %>% 
   summarise(jobs = n_distinct(job_id)) %>% 
   print()

res3 <- res2 %>% 
   summarise(avg_jobs = mean(jobs)) %>% 
   print()
```

<div class="solution">
We have a total of `r res1` jobs, the number of jobs for each ship can be seen using `res2` with and average of `r res3[1,1]` jobs per ship.
</div>




## `r ctrQ()`

What is the minimum, average, and  maximum number of *different* items considered at each job? What is the minimum, average, and  maximum number of items used for maintenance at each job?

```{r, solution=TRUE}
res1 <- dat %>% 
   distinct(job_id, item_id) %>% 
   count(job_id) %>% 
   summarise(min = min(n), avg = mean(n), max = max(n)) %>% 
   print()

res2 <- dat %>% 
   group_by(job_id) %>% 
   summarise(n = sum(item_quantity)) %>% 
   summarise(min = min(n), avg = mean(n), max = max(n)) %>% 
   print()
```

<div class="solution">
The range and average of unique items considered can be seen using `res1` and the range and average of items used can be seen using `res2`. Note some jobs don't use the allocated item.
</div>




## `r ctrQ()`

It seems that some items considered in a job are not used anyway. Which ship has most jobs where an item should have been used, but has not (quantity is zero)?

```{r, solution=TRUE}
res <- dat %>% 
   filter(item_quantity == 0) %>% 
   group_by(ship) %>%
   summarise(jobs = n_distinct(job_id)) %>% 
   arrange(desc(jobs)) %>% 
   print()
```

<div class="solution">
`r res$ship[1]` have a lot of jobs where the items scheduled in a job, have not been used.
</div>




## `r ctrQ()`

Which item is the most used one for maintenance?

```{r, solution=TRUE}
res <- dat %>% 
   group_by(item_id) %>%  
   summarize(n = sum(item_quantity), item_name = first(item_name)) %>% 
   arrange(desc(n)) %>% 
   print()
```

<div class="solution">
Item `r res$item_id[1]` a `r res$item_name[1]` is used most.
</div>






# R - `r ctrA()` (40%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you may need *dplyr* for preparing the datasets you want to plot). We work with the dataset from Assignment 1, which can be read using:


```{r}
library(tidyverse)
dat <- read_csv("jobs.csv")
```




## `r ctrQ()`

For each year and job id, identify the ship the job was done on. Hint: The `first` function may be used to select the first item within a group. Use this to create a visualization showing the number of maintenance jobs for each year with the following features:

   * Bars are used for each year.
   * Fill colors are used to identify the ship.
   * Legends are put at the bottom of the plot.
   * Informative figure title and axis titles are given.

Which year had the most jobs?

```{r, solution=TRUE}
dat %>% 
   group_by(year, job_id) %>% 
   summarise(ship = first(ship)) %>% 
   ggplot(aes(x = year, fill = ship)) +
   geom_bar() +
   labs(title = "Total number of jobs",
    x = "Year",
    y = "Jobs",
    fill = "Ship") +
   theme(legend.position = "bottom")
```

<div class="solution">
Most jobs where done in 2018.
</div>




## `r ctrQ()`

Create a visualization showing the number of different items used for each ship with the following features:

* The numbers are shown using columns.
* Reorder the columns so they increase along the x-axis.
* Informative figure title and axis titles are given.

Which ship uses most items? 

```{r, solution=TRUE}
res <- dat %>% 
   group_by(ship) %>% 
   summarize(n = n_distinct(item_id)) %>% 
   arrange(desc(n)) 
res %>% 
   ggplot(aes(x = reorder(ship, n), y = n)) +
   geom_col() +
   labs(title = "Number of unique items used",
    x = "Ship",
    y = "Items") +
   theme(legend.position = "bottom")
```

<div class="solution">
Most items are used for `r res$ship[1]`.
</div>



## `r ctrQ()`

Create a visualization showing the number of items used as a function of done date with the following features:

* The numbers are shown using columns with a fixed line width of 1 and a blue color. 
* The plot is divided using `ship` (facet). Hint: You may use `scales = "free_y"`. 
* Informative figure title and axis titles are given.

Which ship has most items used at a specific date?

```{r, solution=TRUE, warning=FALSE}
res <- dat %>% 
   group_by(ship, done_date) %>% 
   summarize(n = sum(item_quantity)) 
res %>% 
   ggplot(aes(x = done_date, y = n)) +
   geom_col(linewidth = 1, color = "blue") +
   facet_wrap(~ ship, scales = "free_y") +
   labs(title = "Number of items used",
    x = "Date",
    y = "Items used") 
```

<div class="solution">
On a specific date must items are used for Ship-12.
</div>




## `r ctrQ()`

Consider two items with id:

```{r}
items <- c("601.004.006", "601.026.128")
```

Create a visualization showing the variation of demand (item quantity) of each item with the following features:

* A density is plotted for each item with a fixed bandwidth (`bw`) of 1 and transparency (`alpha`) of 0.5. 
* Item name is used as fill. 
* The plot is divided using `year` (facet).
* Informative figure title and axis titles are given.

Over the years, is it consistent that on average the demand of one of the items is higher than the other?

```{r, solution=TRUE}
res <- dat %>% 
   filter(item_id %in% items)
res %>% 
   ggplot(aes(item_quantity, fill = item_name)) + 
   geom_density(bw = 1, alpha = 0.5) +
   facet_wrap(~ year) +
   labs(title = "Demand density",
    x = "Demand",
    y = "",
    fill = "Item") +
   theme(legend.position = "bottom")
```

<div class="solution">
It can be seen that over the years, the demand peak of the Piston Ring is higher that for the O-Ring, i.e. it seems to hold. 
</div>




## `r ctrQ()`

Consider two items with id:

```{r}
items <- c("601.004.006", "601.026.128")
```

Create a visualization showing the variation of days since last used for maintenance given an item with the following features:

* A violin is used to plot days since last used given an item. 
* Item name is used as fill. 
* Informative figure title and axis titles are given.

Comment on the plot.

```{r, solution=TRUE}
dat %>% 
    filter(item_id %in% items) %>% 
   ggplot(aes(y = days, x = item_id, fill = item_name)) + 
   geom_violin() + 
   labs(title = "Violin density",
    x = "Item id",
    y = "Days",
    fill = "Item") +
   theme(legend.position = "bottom")
```

<div class="solution">
Most often there are less than 100 days between a repeated usage of the item, but there can be over 600 days. The Piston Ring have in general a lower number of days in between than the O-Ring on average.
</div>





# R - `r ctrA()` (20%)

Create a function `process` with the following features:

   * The arguments are:
     - `v`: A vector with numbers (may contain missing values,`NA`s).
     - `find`: A number to search for. Default value is `NULL` indicating not to search for a number.
     - `replace`: A number. If `find` is a number, then replace all occurrences of `find` in the vector with `replace` (if any). Default value is `NULL` indicating not to replace anything.
   * The function returns a list with 
     - the original vector `v`, 
     - the length of `v`, 
     - the number of missing values,
     - if find is a number, the number of times the number is found, and
     - if replace is a number, the new vector with the replaced numbers.
   * Document your function. 

```{r, solution=TRUE}
#' Process a vector
#'
#' @param v The vector.
#' @param find A number to search for. Default value is `NULL` indicating not to 
#'        search for a number.
#' @param replace A number. If `find` is a number then replace all these numbers 
#'        with `replace`. Default value is `NULL` indicating not to replace anything.
#' @return A list with the processed vector and some info.
process <- function(v, find = NULL, replace = NULL) {
   lst <-list(original = v, length = length(v), missing = sum(is.na(v)))
   if (!is.null(find)) {
      idx <- which(v == find)
      lst$found <- length(idx)
      if (!is.null(replace)) {
         v[idx] <- replace
         lst$new <- v
      }
   }
   return(lst)
}
v <- c(8, 19, -2, 8, NA, 12, 9, 23, NA, -34)
process(v)
process(v, find = 1)
process(v, find = 8)
process(v, find = 8, replace = 5)
```





