---
title: "TFA exam"
---

<style>
.solution {
  background-color: aliceblue;
  line-height: 1.5;
  margin-top: 5px;
  margin-bottom: 15px;
  margin-right: 0px;
  margin-left: 0px;
  padding-top: 10px;
  padding-right: 0px;
  padding-bottom: 1px;
  padding-left: 5px;
  /*border: 3px solid green;*/
}

h2 {
  font-size: large;
}
</style>



```{r setup, include=FALSE, eval=T}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  eval = TRUE,
  message=FALSE)
options(width = 90, scipen = 10, digits = 2)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```
















# R - `r ctrA()` (55%)
 
Consider the dataset in the file `catch.csv` containing catches from rivers in Denmark. A row is given for each catch with information about weight, length, date, angler, place, method etc.

Some of the columns/variables are:

  - `Date`: Date of catch.
  - `Length`: Length of fish (cm).
  - `Weight`: Weight of fish (kilograms).
  - `Place`: Place of catch (Lower, Middle, Upper).
  - `Method`: Method used (Spin, Fly, Worm).
  - `Killed`: If the fish was killed.
  - `River`: River of catch.
  - `Species`: Species of fish.
  - `Name`: Name of angler.

The dataset can be read using:

```{r, warning=FALSE}
library(tidyverse)
dat <- read_csv("catch.csv") 
```

Moreover, consider the dataset in the file `weight.csv` containing weight estimates for a given length. A row is given for each length with information about weight, month, river, species etc.

Some of the columns/variables are:

  - `Length`: Length of fish (cm).
  - `Month`: Month considered.
  - `Avg`: Average weight of fish (kilograms).
  - `Lower`: Lower confidence interval weight (kilograms).
  - `Upper`: Uppper confidence interval weight (kilograms).

The dataset can be read using:

```{r, warning=FALSE}
library(tidyverse)
dat_weight <- read_csv("weight.csv")
res <- dat_weight |> 
   filter(River == "Karup", 
          Species == "Seatrout", 
          Length == 50, 
          Month == 5) |> 
   print()
```

That is, for instance, the average weight of a 50 cm seatrout in Karup River in May is `r res$Avg` kg with a 95% confidence interval of [`r res$Lower`, `r res$Upper`] kg.

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (tibbles/data frames), and answer/complete the following questions/tasks:


## `r ctrQ()`

Which method resulted in most fish caught in Skjern River?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Which angler has caught most fish in total and in kilograms over all the years considered?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Calculate the condition factor for each fish using the formula `Weight*100000/Length^3`. 

```{r}
# your code here
```

Is the average condition factor higher for salmon or seatrout?

```{r}
# your code here
```

[Your comment here]

Consider the months after April and fish with a weight above 50 cm. Do seatrout in Karup River have a higher average condition factor than seatrout in Skjern River?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Are there more catches per day on average on weekdays or during weekends in Skjern River? 

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Calculate the catch and release rate (number released / total number caught) for each river over all years. Have the catch and release rates increased over the years in Karup River?

```{r}
# your code here
```


## `r ctrQ()`

Use the weight dataset to update missing weights in the catch dataset. That is, for each fish, with a missing weight, use the average weight from the weight dataset for the corresponding length, river, species and month. Hint: Before joining the datasets, you need to round the length to an integer.



```{r}
# your code here
```



# R - `r ctrA()` (45%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you may need *dplyr* for preparing the datasets you want to plot). We use the same dataset as given in Assignment 1 if not specified.


## `r ctrQ()`

Create a visualization showing the average weight of seatrout in Karup River over the months with the following features:

* Use the dataset from Assignment 1 in which missing weights have been updated if possible.
* Bars with the fill colour `darkred` are used.
* An informative figure title and axis titles are given.

Which month has the highest average weight?

```{r}
# your code here
```

[Your comment here]





## `r ctrQ()`

Create a visualization showing the relationship between length and weight for all fish caught with the following features:

* Different colours are used for each species.
* Points are used to show the individual catches with some transparency (alpha).
* A trend line is added for each species.
* An informative figure title and axis titles are given.
* The legend is placed at the bottom of the figure.

Are salmon generally heavier than seatrout for a given length? Can salmon reach a higher weight than seatrout?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Consider the weight dataset. Create a visualization showing the weight for a given length with the following features:

* Only Skjern River is considered.
* Lines are used to show the average weight given length for each species.
* Different colours are used for each month.
* A subplot is made for each species.
* An informative figure title and axis titles are given.
* The legend is placed at the bottom of the figure.

In which month is the weight lowest for seatrout?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Create a visualization showing the percentage of fish over 59 cm killed over the years with the following features:

* Bars are used to visualize the percentage of fish killed.
* Different fill colours are used for each river (use `position = "dodge2"` to place bars beside each other).
* An informative figure title and axis titles are given.
* The legend is placed at the bottom of the figure.

In which river do most anglers catch and release fish (i.e. the lowest percentage killed)?

```{r}
# your code here
```

[Your comment here]


## `r ctrQ()`

Consider seatrout. For each river, year, and month, calculate the number of catches. Next, calculate the catch percentage for each month (number caught per month / total number caught per year). 

```{r}
# your code here
```

Create a visualization showing the variation in these percentages for each month with the following features:

* A boxplot is used to show the variation of the catch percentages.
* A subplot is made for each River.
* An informative figure title and axis titles are given.

Comment on the results for July.

```{r}
# your code here
```

[Your comment here]


