---
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```

# (PART) Getting started with VBA {-}

# A introduction to VBA {#sec-vba-intro}

## Learning path diagramme {-}

Click/hover the nodes to follow links and see details.

```{r, echo=FALSE, out.width="100%", fig.asp=NA}
g <- create_learning_path(
   url = "https://docs.google.com/spreadsheets/d/1bBe42LHK-bE7CsU9eNBzi_7VNjbmv-Ybr7285pE61jM/edit?usp=sharing", 
   sheet = "vba-intro", 
   x_legend = 0, y_legend = 0)
render_graph(g, height = 300)
```

## Learning outcomes {#lo-vba-intro}

By the end of this module, you are expected to have:

* Basic knowledge about what VBA is.
* Setup Excel for VBA.
* Knowledge about how the macro recorder works.
* Made your first program.

The learning outcomes relate to the [overall learning goal](#lg-course) number 2 and 4 of the course.


## What is VBA {#vba-what}

Visual Basic for Applications (VBA) is an implementation of the BASIC programming language intended to control and automate Microsoft Office applications, developed by Microsoft. For instance, you can automatically create sheets, delete objects, create user-defined functions or read/write data to a sheet. It is not a standalone program, it can only run in the host application. In this course we will focus on runing VBA from Excel. VBA is widely used in the industry (specially linked to Excel) and easy to learn. Microsoft stopped investing in VBA in 2008. It only update it for small changes. However, VBA is still a vital part of desktop Office applications, and will continue to be so in the future. 

VBA is a [compiled language](https://en.wikipedia.org/wiki/Compiled_language#:~:text=A%20compiled%20language%20is%20a,%2Druntime%20translation%20takes%20place) implemented using compilers (translators that generate machine code from source code). That is, code need to be compiled first before running it. You can only run VBA using the desktop version of Excel. That is, you can't create, run, or edit VBA in Excel for the web. 


## Your first program {#vba-hello}

We will use the book _VBA Programming in Business Economics_ by @vba-book. Read Chapter 1 in @vba-book.


## The macro recorder {#vba-macro}

Read Chapter 2 in @vba-book.












```{r links, child="links.md", include=FALSE}
```