```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```

# (PART) Programming in VBA {-}

# Data types

This module gives a short introduction to VBA. In general syllabus will point to chapters in @vba-book; however, there is a lot for videos about VBA online such as courses [30 for 30 Excel VBA Absolute Beginner Course][vba-yt-course1] and [14-Hour VBA Course][vba-yt-course2]. You may have a look at these videos instead if you prefer a different learning style. In the learning path diagram links to alternative online content will be pointed out. Note this is alternatives to the standard learning path that you may use instead. The learning path may also have extra content that is not a part of syllabus you can have a look at.

Since Excel change as new versions arrive and it work on different operating systems, the look in the screenshots and videos might be a bit different from your version. However, you still should be able to understand it. 

__Learning path diagram__

Click/hover the nodes to follow links and see details.

```{r, echo=FALSE, out.width="100%", fig.asp=NA}
g <- create_learning_path(
   url = "https://docs.google.com/spreadsheets/d/1bBe42LHK-bE7CsU9eNBzi_7VNjbmv-Ybr7285pE61jM/edit?usp=sharing", 
   sheet = "vba-datatypes", 
   x_legend = 4.5, y_legend = 6.1)
render_graph(g, height = 500)
```

## Learning outcomes {#lo-vba-datatypes}

By the end of this module, you are expected to:

* Describe what VBA is.
* Setup Excel for VBA.
* Know how the macro recorder works.
* Make your first program.
* Have an overview over what VBA can do.
* Recorded you first macro using the macro recorder

The learning outcomes relate to the [overall learning goals](#lg-course) number 1, 2 and 8 of the course.

<!-- SOLO increasing: identify · memorise · name · do simple procedure · collect data · -->
<!-- enumerate · describe · interpret · formulate · list · paraphrase · combine · do -->
<!-- algorithms · compare · contrast · explain causes · analyse · relate · derive · -->
<!-- evaluate · apply · argue · theorise · generalise · hypothesise · solve · reflect -->


By the end of this module, you are expected to have:



```{r links, child="links.md", include=FALSE}
```