```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```

# Object model {#mod-vba-object-model}


## Exercises {#ex-vba-object model}

`r strExercises`

### Exercise - Worksheets {#ex-vba-worksheets}

The [file](https://github.com/bss-osca/tfa/blob/master/vba/vba-worksheets-template.xlsm) should be used as a starting point for this exercise.

Use the procedure `Main` to test the following questions.

   (1) Have a look at the function `WorksheetExists` and check if a worksheet named _Main_ exists.
   (2) Make a function `WorksheetCreate` that creates a worksheet given an input name (if does not exist in advance) and return true if created and false otherwise. Use the function to create a worksheet named _Testing_.
   (3) Make a function `WorksheetDelete` that deletes a worksheet given an input name (if it exists) and return true if deleted and false otherwise. Use the function to delete the worksheet named _Testing_.
   (4) Make a function `WorksheetRename` that renames a worksheet given input arguments name and a new name (if it exists) and return true if renamed and false otherwise. Use the function to create a worksheet named _Test_ and rename it to _Testing_.
   
You may use the functions you have created in other exercises.


### Exercise - Ranges {#ex-vba-ranges}

The [file](https://github.com/bss-osca/tfa/blob/master/vba/vba-ranges-template.xlsm) should be used as a starting point for this exercise.





### Exercise - Collections {#ex-vba-ranges}



```{r links, child="links.md", include=FALSE}
```