```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```

# (PART) Extra topics in VBA {-}

# Generating random numbers {#mod-vba-random-numbers}


## Generating random numbers {#vba-random-numbers}

Often we want to model a system where some of the elements are uncertain. That is, we want to generate some random numbers following different distributions. This can be done using the built-in VBA and Excel functions.

Read Sections 18.1 and 18.2 of Chapter 18 in @vba-book. You may also have a look at the extra videos as listed in the learning path diagram.

## Simulation {#vba-simulation}





## Exercises {#ex-vba-random-numbers}

`r strExercises`

### Exercise - Generating random numbers {#ex-vba-generate}

This [file](https://github.com/bss-osca/tfa/blob/master/vba/vba-generate-template.xlsm) should be used as a starting point for this exercise.

   (1) Create a procedure `GenNormal` with arguments `intSize`, `dblMean`, `dblSD`  and `aryResult`. The procedure must generate `intSize` random numbers from a normal distribution with mean `dblMean` and standard deviation `dblSD`. The procedure must store the values in `aryResult`.
   (2) Create a procedure `GenUniform` with arguments `intSize`, `dblMin`, `dblMax` and `aryResult`. The procedure must generate `intSize` random numbers from a uniform distribution in the range `dblMin` to `dblMax`. The procedure must store the values in `aryResult`.
   (3) Create a procedure `GenBinomial` with arguments `intSize`, `intTrials`, `dblPr` and `aryResult`. The procedure must generate `intSize` random numbers from a binomial distribution using `intTrials` trials with `dblPr` probability of success. The procedure must store the values in `aryResult`.
   (4) Create a procedure `GenPoisson` with arguments `intSize`, `dblLambda` and `aryResult`. The procedure must generate `intSize` random numbers from a poisson distribution with mean `dblLambda`. The procedure must store the values in `aryResult`. You may use the `PoissonInv` function. 
   (5) Create a procedure `Main` that generate 5000 random numbers of 
       - A normal distribution with mean 100 and standard deviation 20.
       - An uniform distribution with range 10 to 500.
       - A binomial distribution with 100 trials and a 0.2 probability of success.
       - A poisson distribution with mean 5.
       - Plot the results for each distribution using `CreatePlot`.

### Exercise - Traveling Salesman {#ex-vba-tsp}

This [file](https://github.com/bss-osca/tfa/blob/master/vba/vba-tsp-template.xlsm) should be used as a starting point for this exercise.

The [travelling salesman problem](https://en.wikipedia.org/wiki/Travelling_salesman_problem) (TSP) asks the following question: 

> Given a list of cities and the distances between each pair of the cities, 
> what is the shortest possible route that visits each city exactly once 
> and returns to the origin city?

The problem is an NP-hard problem (worst case solution time grows exponential with the number of cities) in combinatorial optimization, important in theoretical computer science and operations research. The problem was first formulated in 1930 and is one of the most intensively studied problems in optimization. Even though the problem is computationally difficult, many heuristics and exact algorithms are known. 

The goal with this exercise is to test different heuristics on a set of problem instances. To see which one works best.

   (1) Create a procedure `GenTSPData` satisfying:
       - Arguments are `ByRef dblCoord() As Double, Optional intCities As Integer = 0, Optional blnPrint As Boolean = False`.
       - Each city has an id number and a  x and y-coordinate.
       - The generated cities are stored in `dblCoord` with `intCities` rows and three columns (column 1 store the city id, column 2 the x-coordinate and column 3 the y-coordinate) 
       - The x and y-coordinates are random numbers from an uniform distribution between 0 and 10. 
       - If `blnPrint` is true then coordinates are printed to the main sheet starting in cell A2. 
       - If `intCities` is zero then use an input box to ask for the number of cities.  
   (2) Create a procedure `BtnGenTSPData` that calls `GenTSPData`, ask for the number of cities and print the result to cell A2. A button linking to the procedure has already been created. 
   (3) Create a procedure `CalcDistArray` satisfying:
       - Arguments are `ByRef dblDist() As Double, dblCoord() As Double` where `dblCoord` are the coordinates found using `GenTSPData`.
       - The distance between two cities are stored in `dblDist`, i.e. `dblDist(i, j)` store the distance between city `i` and `j`.
       - Symmetric distances are assumed, i.e. `dblDist(i, j) = dblDist(j, i)`.
   (4) Create a procedure `SolveTSPIncX` satisfying:
       - Arguments are `dblCoord() As Double, dblDist() As Double, ByRef dblCost As Double, ByRef intSeq() As Integer`
       - Sort the array `dblCoord` increasing in the x-coordinate (you may use the procedure `ArySort` in module `ModuleAry` to do this).
       - Visit the cities in the order of the sorted array and return to the starting city.
       - The visiting city sequence are stored in `intSeq`.
       - The total cost are stored in `dblCost`. You may use the `CalcCost` function to find the total cost.
   (5) Create a procedure `BtnSolveTSPIncX` that calls `SolveTSPIncX` using the TSP data in the Main sheet and print out the total cost in cell H4. A button linking to the procedure has already been created.
   (6) Create a procedure `SolveTSPIncY` satisfying:
       - Arguments are `dblCoord() As Double, dblDist() As Double, ByRef dblCost As Double, ByRef intSeq() As Integer`
       - Sort the array `dblCoord` increasing in the y-coordinate (you may use the procedure `ArySort` in module `ModuleAry` to do this).
       - Visit the cities in the order of the sorted array and return to the starting city.
       - The visiting city sequence are stored in `intSeq`.
       - The total cost are stored in `dblCost`. You may use the `CalcCost` function to find the total cost.
   (7) Create a procedure `BtnSolveTSPIncY` that calls `SolveTSPIncY` using the TSP data in the Main sheet and print out the total cost in cell H8. A button linking to the procedure has already been created.
   (8) Create a procedure `SolveTSPRand` satisfying:
       - Arguments are `dblCoord() As Double, dblDist() As Double, ByRef dblCost As Double, ByRef intSeq() As Integer`
       - Visit the cities in the order of the `dblCoord` array and return to the starting city. This may seen as we visit the cities in random order since generate x and y-coordinate random.
       - The visiting city sequence are stored in `intSeq`.
       - The total cost are stored in `dblCost`. You may use the `CalcCost` function to find the total cost.
   (9) Create a procedure `BtnSolveTSPRand` that calls `SolveTSPRand` using the TSP data in the Main sheet and print out the total cost in cell H11. A button linking to the procedure has already been created.
  (10) Create a procedure `SolveTSPNN` satisfying:
       - Arguments are `dblCoord() As Double, dblDist() As Double, ByRef dblCost As Double, ByRef intSeq() As Integer`
       - Start in city 1.
       - Given the current city find the next city (not already visited) as the city with shortest distance.
       - The visiting city sequence are stored in `intSeq`.
       - The total cost are stored in `dblCost`. You may use the `CalcCost` function to find the total cost.
  (11) Create a procedure `BtnSolveTSPNN` that calls `SolveTSPNN` using the TSP data in the Main sheet and print out the total cost in cell H14. A button linking to the procedure has already been created.
  (12) Create a procedure `Simulation` satisfying:
       - Generate an instance of TSP data with `intCities` cities which are chosen random between 10 and 500.
       - Calculate the distance matrix.
       - Solve the instance using all of the above solutions algorithms.
       - Store the result in a row in the _Simulation_ sheet. 
       - Repeat 100 times and calculate min, mean and max values for each solution algorithm.
       
       Have a look at sheet _Simulation example_ for an example. Note you do not have to print the coordinates out to the _Main_ sheet.
  (13) Which solution algorithm seems to give the shortest total travel distance?
```{r links, child="links.md", include=FALSE}
```