```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```

# (PART) Extra topics in VBA {-}

# Generating random numbers  {#mod-vba-random-numbers}




## Exercises {#ex-vba-random-numbers}

`r strExercises`

### Exercise - Generating numbers {#ex-vba-generate}


The [file](https://github.com/bss-osca/tfa/blob/master/vba/vba-random-template.xlsm) may be used as a starting point for this exercise.

   (1) Create a procedure `GenNormal` with arguments `intSize`, `dblMean`, `dblSD`  and `aryResult`. The procedure must generate `intSize` random numbers from a normal distribution with mean `dblMean` and standard deviation `dblSD`. The procedure must store the values in `aryResult`.
   (2) Create a procedure `GenUniform` with arguments `intSize`, `dblMin`, `dblMax` and `aryResult`. The procedure must generate `intSize` random numbers from a uniform distribution in the range `dblMin` to `dblMax`. The procedure must store the values in `aryResult`.
   (3) Create a procedure `GenBinomial` with arguments `intSize`, `intTrials`, `dblPr` and `aryResult`. The procedure must generate `intSize` random numbers from a binomial distribution using `intTrials` trials with `dblPr` probability of success. The procedure must store the values in `aryResult`.
   (4) Create a procedure `GenPoisson` with arguments `intSize`, `dblLambda` and `aryResult`. The procedure must generate `intSize` random numbers from a poisson distribution with mean `dblLambda`. The procedure must store the values in `aryResult`. You may use the `PoissonInv` function. 
   (5) Create a procedure `Main` that generate 5000 random numbers of 
       - A normal distribution with mean 100 and standard deviation 20.
       - An uniform distribution with range 10 to 500.
       - A binomial distribution with 100 trials and a 0.2 probability of success.
       - A poisson distribution with mean 5.
       - Plot the results for each distribution using `CreatePlot`.

```{r links, child="links.md", include=FALSE}
```