---
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```


<!-- Exercises from: https://rafalab.github.io/dsbook/programming-basics.html -->

# (PART) Programming in R {-} 


# Loops and conditionals {#mod-r-loops-cond}

This module considers programming with loops and conditional statements in R.

We are all different and you may like different learning styles compared to others. In the learning path diagram, there may be links to alternative online content. Note this is an alternative to the standard learning path that you may use instead. The learning path may also have extra content, that is not a part of syllabus, you can have a look at. 

```{r, echo=FALSE, out.width="100%", fig.asp=NA, include=TRUE}
g <- create_learning_path(
   url = "https://docs.google.com/spreadsheets/d/1bBe42LHK-bE7CsU9eNBzi_7VNjbmv-Ybr7285pE61jM/edit?usp=sharing", 
   sheet = "r-loops-cond", 
   x_legend = 3.5, y_legend = 4.25)
render_graph(g)
```


## Learning outcomes {#lo-loops}

By the end of this module, you are expected to be able to:

* Formulate conditional statements.
* Use functions `any` and `all`.
* Formulate loops in R using for and while statements.
* Use function `if_else`.

The learning outcomes relate to the [overall learning goals](#lg-course) number 2, 4 and 10 of the course.

<!-- SOLO increasing: identify · memorise · name · do simple procedure · collect data · -->
<!-- enumerate · describe · interpret · formulate · list · paraphrase · combine · do -->
<!-- algorithms · compare · contrast · explain causes · analyse · relate · derive · -->
<!-- evaluate · apply · argue · theorise · generalise · hypothesise · solve · reflect -->


## Conditionals and control flow {#r-dc-loops-cond}

An excellent introduction to conditionals and if statements is given in [Chapter 1][datacamp-r-intermediate-chap1] of the interactive DataCamp course [Intermediate R][datacamp-r-intermediate]. Please complete the chapter before continuing. 

Some functions are also useful for comparing logical data types. Consider this example: 

```{r}
x <- c(1, 3, 5, 10, 2, 17, 11, NA, 4)
x > 10  # are the elements greater that 10
any(x > 10)  # are any of the elements greater that 10
all(x > 10)  # are all of the elements greater that 10
all(x < 20)  # are all of the elements greater that 20
all(x < 20, na.rm = TRUE)  # are all of the elements greater that 20
```
That is, functions `any` and `all` can be used to join logical values in vectors. 

Some if statements can be written alternatively using function `if_else`:

```r
if_else(condition, true, false, missing = NULL)
```

For example:

```{r}
x <- c(-5:5, NA)
x

## using if and for
res <- rep("", length(x))
for (i in seq_along(x)) {
  if (is.na(x[i])) res[i] <- "missing"
  else if (x[i] < 0) res[i] <- "negative"
  else res[i] <- "positive"
}
res

## implicit if statement
res <- rep("", length(x))
res
res[x < 0] <- "negative"
res[x >= 0] <- "positive"
res[is.na(x)] <- "missing"
res

## using if_else
res <- if_else(x < 0, "negative", "positive", "missing")
res
```




## Loops {#r-loops}

An excellent introduction to conditionals and if statements is given in [Chapter 2][datacamp-r-intermediate-chap2] of the interactive DataCamp course [Intermediate R][datacamp-r-intermediate]. Please complete the chapter before continuing (stop when Chapter 2 finishes). 

Loops in R may be slow. However, not if you follow some golden rules:

* Do not use a loop when a vectorized alternative exists.
* Do not grow objects (via `c`, `cbind`, etc) during the loop - R has to create a new object and copy across the information just to add a new element or row/column. Instead, allocate an object to hold the results and fill it in during the loop.

As an example, consider the for loop with 4 iterations:

```{r}
i_val <- c(1,2,6,9)
res <- rep(NA,4)
res
for (idx in 1:length(i_val)) {
  res[idx] <- 6 * i_val[idx] + 9
}
res
```

Note we allocate memory for the result vector before the loop so we do not have to grow the result object. Next, we calculate results $6i+9$ using a loop. Be careful here! This is not the same:

```{r}
res <- rep(NA,4)
for (i in i_val) {
  res[i] <- 6 * i + 9
}
res
```

In this example, however, we can use a vectorized alternative:

```{r}
res <- 6 * i_val + 9
res
```

where the operation is applied to each element in the vector.

Nested for loops is also possible. A simple example of a nested loop:

```{r}
for (i in 1:3) {
  for (j in 1:2) {
    cat(str_c("i =", i, " j = ",j, "\n"))
  }
}
```

We here use the function `cat` to print out a string (`\n` indicates new line). Note how the nested loops are executed:

* Set `i = 1` (outer loop)
   - Set `j = 1` (inner loop), `i` stays 1
   - Set `j = 2` (inner loop), `i` stays 1
   - Inner loop finishes, proceed with outer loop.
* Increase `i = 2` (outer loop)
   - Set `j = 1` (inner loop), `i` stays 2
   - Set `j = 2` (inner loop), `i` stays 2
   - Inner loop finishes, proceed with outer loop.
* Increase `i = 3` (outer loop)
   - Set `j = 1` (inner loop), `i` stays 3
   - Set `j = 2` (inner loop), `i` stays 3
   - Inner loop finishes, proceed with outer loop.
* Outer loop finishes as well (we looped over `i in 1:3`). Job done.

Nested loops can be used to iterate over matrices or data frames:

```{r}
mat <- matrix(NA, nrow = 2, ncol = 3)
mat
for (i in 1:nrow(mat)) {
  for (j in 1:ncol(mat)) {
     mat[i,j] <- (i-1)*ncol(mat) + j
     cat(str_c("Entry (", i, ", ", j, ") = ", mat[i,j], "\n"))
  }
}
mat
```

Often you can replace nested loops with a single loop by using `expand_grid`:

```{r}
library(tidyverse)  # load function expand_grid
mat <- matrix(NA, nrow = 2, ncol = 3)
ite <- expand_grid(i = 1:2, j=1:3)
ite
for (r in 1:nrow(ite)) { # iterate over rows
   i <- ite$i[r]
   j <- ite$j[r]
   mat[i,j] <- (i-1)*ncol(mat) + j
   cat(str_c("Entry (", i, ", ", j, ") = ", mat[i,j], "\n"))
}
mat
```

Note `expand_grid` creates a data frame with all combinations. This way of looping is a more flexible approach since you can

* nest more loops by adding more columns to `ite`,
* add different values in each column.

For instance, if you only want to calculate values for row 2 and columns 1 and 3 the code becomes:

```{r}
mat <- matrix(NA, nrow = 2, ncol = 3)
ite <- expand_grid(i = 2, j = c(1,3))
ite
for (r in 1:nrow(ite)) { # iterate over rows
   i <- ite$i[r]
   j <- ite$j[r]
   mat[i,j] <- (i-1)*ncol(mat) + j
   cat(str_c("Entry (", i, ", ", j, ") = ", mat[i,j], "\n"))
}
mat
```


## Recap {#rc-r-loops-cond}

*Comparison/relational operators* known to R are:

- `<` for less than,
- `>` for greater than,
- `<=` for less than or equal to,
- `>=` for greater than or equal to,
- `==` for equal to each other (and not `=` which is typically used for assignment!),
- `!=` not equal to each other.
   
*Logical operators* known to R are:

- `&` and,
- `|` or,
- `!` not.

If you use `&&` and `||` only the first element in vectors are compared. In general this is used rarely. 

Useful functions that return a logical are `any` and `all` which can be used to join logical values in vectors.

Conditional Statements can be constructed using for instance `if` and `while` statements. Moreover, function `if_else` is a vectorized alternative.

*Loops* can be created using `for` and `while` statements. 

You can break out of a loop using `break` and jump to the next iteration (skipping the remainder of the code in the loop) using `next`.

Do not use a loop when a vectorized alternative exists.

Do not grow objects during the loop. Instead, allocate an object to hold the results and fill it in during the loop.

Nested loops are possible in R. However, often they can be converted into a single loop by defining a data frame having the values of the nested loops in each row. Here function `expand_grid` may be useful to create the data frame. 

You may also have a look at the [slides for this module](https://bss-osca.github.io/tfa/slides/06-01_r-loops-cond-slides.html).


## Exercises {#ex-r-loops-cond}

`r strExercises`

[Link to the project on RStudio Cloud for this module][r-cloud-mod13]. 

### Exercise (conditional expressions)

*Solve this exercise using a script file*

1) Consider object `x`:

   ```{r}
   x <- c(1,2,-3,4)
   ```
  
   What will this conditional expression return?

   ```{r, eval=FALSE}
   if(all(x>0)){
    print("All Postives")
   } else {
    print("Not all positives")
   }
   ```

2) What will the following expressions return?

   ```{r, eval=FALSE}
   x <- c(TRUE, FALSE, TRUE, TRUE)
   all(x)
   any(x)
   any(!x)
   all(!x)
   ```

3) Which of the expressions above is always `FALSE` when at least one entry of a logical vector `x` is TRUE?

Consider vector:

```{r}
library(tidyverse)
x <- 1:15
x
```

```{r, solution=TRUE}
if_else(x < 7, as.integer(0), x)
```
```{r, hint=TRUE, eval=FALSE}
if_else(x < 7, as.integer(___), ___)
```

4) Use the `if_else` function to set elements with value below 7 to 0.



```{r, solution=TRUE}
if_else(x < 7 | x > 10, NA_integer_, x)
```
```{r, hint=TRUE, eval=FALSE}
if_else(___, NA_integer_, ___)
```

5) Use the `if_else` function to set elements with value below 7 or above 10 to `NA_integer_` (which is the NA/missing value of an integer).



```{r, solution=TRUE}
x <- sample(c(1:10,NA,5.5), 1)
if (is.na(x)) {
  y <- "missing"
} else if (x %% 2 == 0) {
  y <- "even"
} else if (x %% 2 == 1) {
  y <- "odd"
} else if (x %% 1 > 0) {
  y <- "decimal"
} 
x
y
```
```{r, hint=TRUE, eval=FALSE}
x <- sample(c(1:10,NA,5.5), 1)
if (is.na(x)) {
  y <- ___
} else if (x %% 2 == 0) {
  ___
} else if (___) {
  ___
} else if (___) {
  y <- "decimal"
} 
x
y
```

6) Consider code

   ```{r}
   x <- sample(c(1:10,NA,5.5), 1)
   x
   ```

   which generates a number from the vector `c(1:10,NA,5.5)`. 

   Write code which set object `y` equal to "even" if `x` is even, "odd" if `x` is odd, "decimal" if `x` has a decimal not zero and "missing" if `x` is `NA`. Hint: have a look at `?'%%'` (the modulo operator).





### Exercise (loops)

```{r, solution=TRUE}
x <- rep(NA,4)
for (i in 1:4) {
  x[i] <- 2 * i + 4
}
x
```
```{r, hint=TRUE, eval=FALSE}
x <- rep(NA,4)
for (i in 1:4) {
  x[i] <- ___
}
x
```

1) Using a `for` loop, create a vector having values $2i + 4$ given $i=1\ldots 4$.



```{r, solution=TRUE}
i_val <- c(2, 5, 6, 12)
x <- rep(NA, length(i_val))
for (idx in 1:length(i_val)) {
  x[idx] <- 2 * i_val[idx] + 4
}
```
```{r, hint=TRUE, eval=FALSE}
i_val <- c(2, 5, 6, 12)
x <- rep(NA, length(i_val))
for (idx in 1:length(i_val)) {
  x[idx] <- ___
}
```

2) Using a `for` loop, create a vector having values $2i + 4$ given $i=2,5,6,12$.



```{r, solution=TRUE}
i_val <- c(2, 5, 6, 12)
x <- rep(NA, length(i_val))
idx <- 1
while (idx < 5) {
  x[idx] <- 2 * i_val[idx] + 4
  idx <- idx + 1
}
```
```{r, hint=TRUE, eval=FALSE}
i_val <- c(2, 5, 6, 12)
x <- rep(NA, length(i_val))
idx <- 1
while (idx < 5) {
  x[idx] <- ___
  idx <- ___
}
```

3) Solve Question 2 using a `while` loop.



```{r, solution=TRUE}
2 * 1:4 + 4             # Q1
2* c(2, 5, 6, 12) + 4   # Q2
```
```{r, hint=TRUE, eval=FALSE}
2 * ___ + 4             # Q1
___   # Q2
```

4) Solve Questions 1 and 2 using a vectorized alternative. 




### Exercise (calculating distances) {#ex-dist}

Consider zip codes in Jutland:

```{r}
# remotes::install_github("bss-osca/tfa/tfa-package", upgrade = FALSE)  # run to upgrade
library(tidyverse)
library(tfa)   
zips
```

We want to calculate distances between a subset of zip areas:

```{r}
idx <- 1:5
dat <- zips[idx,]
dat
distanceMat <- matrix(NA, nrow = length(idx), ncol = length(idx))
colnames(distanceMat) <- str_c(dat$Zip[idx], dat$Area[idx], sep = " ") 
rownames(distanceMat) <- colnames(distanceMat)
distanceMat
```

We can find average distances between two zip codes (here rows 1 and 2 in `dat`) using Bing maps:

```{r}
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
             dat$Zip[1], ",Denmark",
             "&wp.1=",
             dat$Zip[2], ",Denmark",
             "&avoid=minimizeTolls&key=", key)
library(jsonlite)
lst <- jsonlite::fromJSON(url)
dist <- lst$resourceSets$resources[[1]]$travelDistance
dist
lst$statusCode
lst$statusDescription
```

Note we call the Bing maps API with the two zip codes. A [json file](https://en.wikipedia.org/wiki/JSON) is returned and stored in a list. To get the average travel distance we access `travelDistance`. The status code should be 200 if the calculation returned is okay.

```{r, solution=TRUE}
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
for(i in 1:nrow(distanceMat)) {
   for(j in 1:ncol(distanceMat)) {
      if (i>j) {distanceMat[i,j] <- distanceMat[j,i]; next}   # assume symmetric distances
      if (!is.na(distanceMat[i,j])) next   # value already calculated 
      if (i==j) {distanceMat[i,j] <- 0; next}
      
      url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
                   dat$Zip[i], ",Denmark",
                   "&wp.1=",
                   dat$Zip[j], ",Denmark",
                   "&avoid=minimizeTolls&key=", key)
      lst <- jsonlite::fromJSON(url)
      if (lst$statusCode == 200) {
         distanceMat[i,j] <- lst$resourceSets$resources[[1]]$travelDistance
      }
   }
}
distanceMat
```
```{r, hint=TRUE, eval=FALSE}
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
for(i in 1:nrow(distanceMat)) {
   for(j in 1:ncol(___)) {
      if (i>j) {distanceMat[i,j] <- ___; next}   # assume symmetric distances
      if (!is.na(distanceMat[i,j])) next   # value already calculated 
      if (i==j) {distanceMat[i,j] <- ___; next}
      
      url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
                   dat$Zip[i], ",Denmark",
                   "&wp.1=",
                   dat$Zip[j], ",Denmark",
                   "&avoid=minimizeTolls&key=", key)
      lst <- jsonlite::fromJSON(url)
      if (lst$statusCode == ___) {
         distanceMat[i,j] <- ___
      }
   }
}
distanceMat
```

Use nested for loops to fill `distanceMat` with distances. Assume that the distance from a to b is the same as from b to a. That is, you only have to call the API once for two zip codes. Use an if statement to check if the status code is okay.


### Exercise (`expand_grid`)

```{r, solution=TRUE}
ite <- expand_grid(i = c(1,5), j = 2:3)
ite
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
for (r in 1:nrow(ite)) { # iterate over rows
   i <- ite$i[r]
   j <- ite$j[r]
   if (i==j) {distanceMat[i,j] <- 0; next}
   url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
                dat$Zip[i], ",Denmark",
                "&wp.1=",
                dat$Zip[j], ",Denmark",
                "&avoid=minimizeTolls&key=", key)
   lst <- jsonlite::fromJSON(url)
   if (lst$statusCode == 200) {
      distanceMat[i,j] <- lst$resourceSets$resources[[1]]$travelDistance
      distanceMat[j,i] <- distanceMat[i,j]
   }
}
distanceMat
```
```{r, hint=TRUE, eval=FALSE}
ite <- expand_grid(i = ___, j = ___)
ite
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
for (r in 1:nrow(ite)) { # iterate over rows
   i <- ite$i[r]
   j <- ite$j[r]
   if (i==j) {distanceMat[i,j] <- 0; next}
   url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
                dat$Zip[i], ",Denmark",
                "&wp.1=",
                dat$Zip[j], ",Denmark",
                "&avoid=minimizeTolls&key=", key)
   lst <- jsonlite::fromJSON(url)
   if (lst$statusCode == ___) {
      distanceMat[i,j] <- lst$resourceSets$resources[[1]]$travelDistance
      distanceMat[j,i] <- ___
   }
}
distanceMat
```

Consider the solution of Exercise \@ref(ex-dist) and assume that you only want to calculate the distance from rows 1 and 5 to rows 2 and 3 in `dat`. Modify the solution using `expand_grid` so only one loop is used.

```{r links, child="links.md"}
```














<!-- <div class = "boxExercise"> -->
<!-- `r addIcon("laptop-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`   -->

<!-- **Exercise 1** -->
<!-- asdf asdlfkj sdfklj fsdjka æ -->

<!--    1. sdf -->
<!--    2. edasf -->

<!-- **Exercise 2** -->
<!-- asdfav sdaf sdhj sdjkahdf jkhasdkjæsdfh asjf -->

<!--    1. sdf -->
<!--    2. sdf -->
<!-- </div> -->


<!-- `r addIcon("file-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("external-link-alt", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->




