---
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```

```{r links, child="links.md"}
```

# (PART) Visualization {-} 

# Data visualization using ggplot {#mod-r-plot}

This module considers visualization of your data using the [ggplot2][tidyverse-ggplot2] package which is a part of [tidyverse][tidyverse-main-page]. R has several systems for making plots, but ggplot2 is one of the most elegant and most versatile. Using ggplot2 you can make plots faster by learning one system and applying it in many different plot types.

We are all different and you may like different learning styles compared to others. In the learning path diagram, there may be links to alternative online content. Note this is an alternative to the standard learning path that you may use instead. The learning path may also have extra content, that is not a part of syllabus, you can have a look at. 

```{r, echo=FALSE, out.width="100%", fig.asp=NA, include=TRUE}
g <- create_learning_path(
   url = "https://docs.google.com/spreadsheets/d/1bBe42LHK-bE7CsU9eNBzi_7VNjbmv-Ybr7285pE61jM/edit?usp=sharing", 
   sheet = "r-plot", 
   x_legend = 1.75, y_legend = 0.55)
render_graph(g)
```

<!-- * A detailed introduction to visualization using ggplot2 is given in Chapters 22-29 in @stat545.  -->
<!-- * A short introduction is given in [Chapter 3](https://r4ds.had.co.nz/data-visualisation.html) in @r4ds. -->
<!-- * Further advanded possibilities for ggplot2 are given in the interactive DataCamp course [Intermediate Data Visualization with ggplot2][datacamp-r-ggplot2-intermediate]. -->


## Learning outcomes {#lo-plot}

By the end of this module, you are expected to:

* Know how to create basic plots using ggplot.
* Formulate the ideas behind the grammar of graphics.
* Explain the idea behind aesthetics such as color, fill, and line type.
* Add geometries to a plot such as a histogram, a boxplot, a barplot, a scatter plot, and a line.
* Understand how themes can be used to modify the overall look of a plot.
* Combine multiple plots into a single graphic.
* Save plots as variables and different image files.

The learning outcomes relate to the [overall learning goals](#lg-course) number 7, 11-14 and 18 of the course.

<!-- SOLO increasing: identify · memorise · name · do simple procedure · collect data · -->
<!-- enumerate · describe · interpret · formulate · list · paraphrase · combine · do -->
<!-- algorithms · compare · contrast · explain causes · analyse · relate · derive · -->
<!-- evaluate · apply · argue · theorise · generalise · hypothesise · solve · reflect -->

## Introduction to data visualization {#r-plot-intro}

The package [ggplot2][tidyverse-ggplot2] is a plotting package that makes it simple to create complex plots from data in a data frame. It provides an interface for specifying which variables to plot, how they are displayed, and general visual properties. Hence, only minimal changes are needed, if the underlying data change or if we decide to change from a bar plot to a scatterplot. 

The package implements the grammar of graphics, a coherent system for describing and building layered plots. A plot is built step by step by adding new layers. Adding layers in this fashion allows for extensive flexibility and customization of plots.

An excellent introduction to data visualization using [ggplot2][tidyverse-ggplot2] is given in the interactive DataCamp course [Introduction to data visualization with ggplot2][datacamp-r-ggplot2-intro]. Please complete the course before continuing. 

Note that there is a difference between using the pipe `%>%` operator which passes the output of the previous line of code as the first input of the next line of code and the `+` operator used between **ggplot2** functions for "layering". That is, you create the plot in layers, separated by `+`.


## Combining plots into one using _patchwork_  {#r-plot-multi}

You can combine separate ggplots into the same graphic using the _patchwork_ package. You can install _patchwork_ from CRAN using `install.packages('patchwork')`. 

The usage is simple. Plots in two rows: 

```{r, include=FALSE}
dev.off()
```

```{r}
library(ggplot2)
library(patchwork)
p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p1 + p2
```

The package provides rich support for arbitrarily complex layouts. Code for nesting three 
plots on top of a third:

```{r, message=FALSE}
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))
(p1 | p2 | p3) /
      p4
```

For further examples see the [documentation pages](https://patchwork.data-imaginist.com/).


## Saving graphics {#r-plot-save}

In general, when you do analytics using R Markdown, there is no need to save your graphics. This is done automatically. However, in a few cases you may need to save you graphics in different formats. Let us consider a simple plot:

```{r}
library(tidyverse)
p <- ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
p  # print it out
```

To save the plot as a bitmap image (png, jpeg etc) have a look at the documentation (`?png`). Let us try to save the plot as a png file.

```{r}
png("test1.png")  # open png device for writing 
p
dev.off()        # close device
png("test2.png", width = 1200, height = 600)   # use other output width and height in px
p
dev.off()
png("test3.png", width = 1200, height = 900)   # save a patchwork plot
(p1 | p2 | p3) /
      p4
dev.off()
# browseURL("test1.png")  # to have a look at the file
# browseURL("test3.png")  # to have a look at the file
```

To save the plot as a pdf use

```{r}
pdf("test1.pdf")  # open pdf device for writing 
p
dev.off()        # close device
# browseURL("test1.pdf")  # to have a look at the file
```

If you use LaTeX you may use the _tikzDevice_ package to save plots as TikZ.

```{r, include=FALSE}
unlink(c("test1.png", "test2.png", "test3.png", "test1.pdf"))
```


## Recap {#rc-r-plot}

* The tidyverse package [ggplot2][tidyverse-ggplot2] is an R package for producing data visualizations. It is based on the Grammar of Graphics by @gramma. 
* The grammar of graphics is a coherent system for describing and building layered plots.
  - Graphics are made by grammatical elements such as data, aesthetics, geometries, scales, facets, and themes.
  - Plots are made though aesthetic mappings. That is, variables are mapped to x or y position using aesthetics attributes such as color, shape, or size.
  - A plot is built step by step by adding new layers. Adding layers in this fashion allows for extensive flexibility and customization of plots.
  - Together, the data, aesthetic mappings, and geometric object form a layer.
  - A plot may have multiple layers, for example, when we overlay a scatterplot with a smoothed line.
* Aesthetics are add in `ggplot` using the `aes` function or alternatively in `geom_` functions.
* Geometries (e.g. a boxplot or line) are added to a plot using the `geom_` functions.
* Themes can be applied to the plot using the `theme_` functions and control all the non-data ink used to modify the overall look of a plot.
* Separate ggplots can be combined into the same graphic using the _patchwork_ package. 
* Save plots as variables and different image files using the device functions such as `png` and `pdf`.
* The pipe `%>%` operator is used to "pipe" the output of the previous line of code as the first input of the next line of code.
* The `+` operator in **ggplot2** functions is used for "layering". This means you create the plot in layers, separated by `+`.
* The 'Data visualization with ggplot2' cheatsheet is very useful. Find the newest version in RStudio **Help > Cheatsheets**.
* A good place to see examples are on the [main reference page](https://ggplot2.tidyverse.org/reference/index.html). Follow the link to the function of interest and have a look at the examples.

You may also have a look at the [slides for this module](https://bss-osca.github.io/tfa/slides/08-01_r-plot-slides.html).


## Exercises {#ex-r-plot}

`r strExercises` 

[Link to the project on RStudio Cloud for this module][r-cloud-mod18]. 



### Exercise (gapminder) {#ex-gapminder2}

Use the *exercise R Markdown template* to solve this exercise (**File > New File > R Markdown...**, select **From template** and then **TFA Exercise**).

In this exercise, we will demonstrate how relatively simple ggplot2 code can create insightful and aesthetically pleasing plots. As motivation we will create plots that help us better understand trends in world health and economics. 

[Hans Rosling](https://en.wikipedia.org/wiki/Hans_Rosling) was the co-founder of the [Gapminder Foundation](http://www.gapminder.org/), an organization dedicated to educating the public by using data to dispel common myths about the so-called developing world. Hans Rosling conveyed actual data-based trends in a dramatic way of his own, using effective data visualization. Here we will try to answer two questions:

* Is it a fair characterization of today's world to say it is divided into Western rich nations and the developing world in Africa, Asia, and Latin America? 
* Has income inequality across countries worsened during the last 40 years? 

To answer these questions, we will be using the `gapminder` dataset provided in the dslabs package. This dataset was created using a number of spreadsheets available from the Gapminder Foundation. You can access the table like this:

```{r load libraries, message=FALSE, echo=TRUE}
library(tidyverse)
library(dslabs)
data(gapminder)
gapminder %>% as_tibble()
```

We start by testing our knowledge regarding differences in child mortality across different countries. For each of the six pairs of countries below, which country do you think had the highest child mortality rates in 2015? Which pairs do you think are most similar?

* Sri Lanka or Turkey
* Poland or South Korea
* Malaysia or Russia
* Pakistan or Vietnam
* Thailand or South Africa

When answering these questions without data, the non-European countries are typically picked as having higher child mortality rates: Sri Lanka over Turkey, South Korea over Poland, and Malaysia over Russia. It is also common to assume that countries considered to be part of the developing world: Pakistan, Vietnam, Thailand, and South Africa, have similarly high mortality rates. 

To answer these questions __with data__, we can use __dplyr__. For example, for the first comparison we see that:

```{r, message=FALSE}
gapminder %>%
  filter(year == 2015 & country %in% c("Sri Lanka","Turkey")) %>%
  select(country, infant_mortality)
```
Turkey has the higher infant mortality rate.

We can use this code on all comparisons and find the following:
```{r, echo = FALSE}
comp_table <- tibble(comparison = rep(1:5, each = 2),
           country = c("Sri Lanka", "Turkey", "Poland", "South Korea", "Malaysia", "Russia", "Pakistan","Vietnam","Thailand","South Africa"))

tmp <- gapminder %>%
  filter(year == 2015) %>%
  select(country, infant_mortality) %>%
  mutate(country = as.character(country)) ##to match characters to characters

tab <- inner_join(comp_table, tmp, by = "country") %>% select(-comparison)

tmp <- bind_cols(slice(tab,seq(1,9,2)), slice(tab,seq(2,10,2)))
names(tmp) <- c("country", "infant mortality", "country", "infant mortality")

knitr::kable(tmp, "html") %>%
    kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE)

```

We see that the European countries on this list have higher child mortality rates: Poland has a higher rate than South Korea, and Russia has a higher rate than Malaysia. We also see that Pakistan has a much higher rate than Vietnam, and South Africa has a much higher rate than Thailand. It turns out that when Hans Rosling gave this quiz to educated groups of people, the average score was less than 2.5 out of 5, worse than what they would have obtained had they guessed randomly. This implies that we are misinformed. We will try to use visualization to help us being more informed.



#### The west vs. the developing world {-}

There is a preconceived notion that the world is divided into two groups: the Western world (Western Europe and North America), characterized by long life spans and small families, versus the developing world (Africa, Asia, and Latin America) characterized by short life spans and large families. But do the data support this dichotomous view?


```{r, solution=TRUE, text = "Most points fall into two distinct categories:\n\n1. Life expectancy around 70 years and 3 or fewer children per family.\n2. Life expectancy lower than 65 years and more than 5 children per family.\n3. Countries are from the regions we expect."}
filter(gapminder, year == 1962) %>%
  ggplot( aes(fertility, life_expectancy, color = continent)) +
  geom_point() 
``` 

```{r, hint=TRUE, eval=FALSE}
filter(gapminder, year == ___) %>%
  ggplot( aes(___, ___, color = ___)) +
  geom_point() 
``` 

   1) Make a scatterplot of life expectancy versus fertility rates (average number of children per woman) in 1962. Use continent as color aesthetic. 
   
   

```{r, solution=TRUE, text = "This plot clearly shows that the majority of countries have moved from the _developing world_ cluster to the _western world_ one. In 2012, the western versus developing world view no longer makes sense. This is particularly clear when comparing Europe to Asia, the latter of which includes several countries that have made great improvements."}
filter(gapminder, year %in% c(1962, 2012)) %>%
  ggplot(aes(fertility, life_expectancy, col = continent)) +
  geom_point() +
  facet_grid(cols = vars(year))
``` 

```{r, hint=TRUE, eval=FALSE}
filter(gapminder, ___ %in% c(1962, 2012)) %>%
  ggplot(aes(___, ___, col = ___)) +
  geom_point() +
  facet_grid(cols = vars(___))
``` 

2) In 1962, "the West versus developing world" view was grounded in some reality. Is this still the case 50 years later? We could easily plot the 2012 data in the same way we did for 1962. To make comparisons, side by side plots are preferable. In __ggplot2__, we can achieve this by _faceting_ variables and making a plot for each year. That is, you must filter by years 1962 and 2012 and add the layer `facet_grid`, which automatically separates the plots.


   
```{r, solution=TRUE, text = "The plot clearly shows how most Asian countries have improved at a much faster rate than European ones."}
years <- c(1962, 1970, 1980, 1990, 2000, 2012)
continents <- c("Europe", "Asia")
gapminder %>% 
  filter(year %in% years & continent %in% continents) %>%
  ggplot( aes(fertility, life_expectancy, col = continent)) +
  geom_point() +
  facet_wrap(vars(year)) 
``` 

```{r, hint=TRUE, eval=FALSE}
gapminder %>% 
  filter(year %in% ___ & continent %in% ___) %>%
  ggplot(aes(___)) +
  geom_point() +
  facet_wrap(___) 
``` 
   
   3) To explore the transformation through the years, make a plot for the years 1962, 1970, 1980, 1990, 2000, and 2012 considering Europe and Asia. How has Asia transformed through the years compared to Europe? Since we consider many years, we will not want all the plots on the same row. Instead, we will want to use multiple rows and columns. The function `facet_wrap` permits us to do this by automatically wrapping the series of plots.
      
  
<div class = "box">

#### Infobox - Scales {-}

The default choice of the range of the axes is important. When not using `facet`, this range is determined by the data shown in the plot. When using `facet`, this range is determined by the data shown in all plots and therefore kept fixed across plots. This makes comparisons across plots much easier. For example, in the above plot, we can see that life expectancy has increased and the fertility has decreased across most countries. We see this because the cloud of points moves. This is not the case if we adjust the scales:

```{r, echo=FALSE, warning=FALSE}
filter(gapminder, year%in%c(1962, 2012)) %>%
  ggplot(aes(fertility, life_expectancy, col = continent)) +
  geom_point() +
  facet_wrap(. ~ year, scales = "free")
```

In the plot above, we have to pay special attention to the range to notice that the plot on the right has a larger life expectancy. 
</div>


```{r, solution=TRUE}
gapminder %>% 
  filter(continent == "Asia") %>%
  ggplot(aes(fertility, life_expectancy, col = year)) +
  geom_point()  
``` 
```{r, hint=TRUE, eval=FALSE}
gapminder %>% 
  filter(___) %>%
  ggplot(aes(___)) +
  geom_point() 
``` 

  4) Illustrate the transformation for Asia using a single plot where year is used as color aesthetic. 


#### Time series plots {-}

The visualizations above effectively illustrate that data no longer supports the Western versus developing world view. Once we see these plots, new questions emerge. For example, which countries are improving more and which ones less? Was the improvement constant during the last 50 years or was it more accelerated during certain periods? For a closer look that may help answer these questions, we introduce _time series plots_.

Time series plots have time in the x-axis and an outcome or measurement of interest on the y-axis. For example, here is a trend plot of United States fertility rates:

```{r fertility-time-series-plot-points, warning=FALSE}
gapminder %>% 
  filter(country == "United States") %>% 
  ggplot(aes(year, fertility)) +
  geom_point()
```

We see that the trend is not linear at all. Instead there is sharp drop during the 1960s and 1970s to below 2. Then the trend comes back to 2 and stabilizes during the 1990s.

When the points are regularly and densely spaced, as they are here, we create curves by joining the points with lines, to convey that these data are from a single series, here a country. To do this, we use the `geom_line` function instead of `geom_point`. 


```{r, solution=TRUE}
gapminder %>% 
  filter(country == "United States") %>% 
  ggplot(aes(year, fertility)) +
  geom_line()
``` 

  5) Make a lineplot showing the time series of fertility versus year for United States.
  
  
```{r, solution=TRUE, text = "The plot clearly shows how South Korea's fertility rate dropped drastically during the 1960s and 1970s, and by 1990 had a similar rate to that of Germany."}
countries <- c("South Korea", "Germany")
gapminder %>% filter(country %in% countries) %>% 
  ggplot(aes(year, fertility, col = country)) +
  geom_line() 
``` 

```{r, hint=TRUE, eval=FALSE}
gapminder %>% filter(country %in% ___) %>% 
  ggplot(aes(year, fertility, col = ___)) +
  geom_line()
``` 

  6) Lineplots is particularly helpful when we look at more countries. Make a lineplot showing the time series of fertility versus year for South Korea and Germany. Use country as color aesthetic. 
  
  
```{r, solution=TRUE, text = "The plot clearly shows how an improvement in life expectancy followed the drops in fertility rates. In 1960, Germans lived 15 years longer than South Koreans, although by 2010 the gap is completely closed. It exemplifies the improvement that many non-western countries have achieved in the last 40 years."}
gapminder %>% 
  filter(country %in% countries) %>% 
  ggplot(aes(year, life_expectancy, col = country)) +
  geom_line() 
``` 

  7) Make a lineplot showing the time series of life expectancy versus year for South Korea and Germany. Use country as color aesthetic. 
  
  
  
#### Data transformations {-}

We now shift our attention to the second question related to the commonly held notion that wealth distribution across the world has become worse during the last decades. When general audiences are asked if poor countries have become poorer and rich countries become richer, the majority answers yes. By using stratification, histograms, smooth densities, and boxplots, we will be able to understand if this is in fact the case. First we learn how transformations can sometimes help provide more informative summaries and plots.

The `gapminder` data table includes a column with the countries' gross domestic product (GDP). GDP measures the market value of goods and services produced by a country in a year. The GDP per person is often used as a rough summary of a country's wealth. Here we divide this quantity by 365 to obtain the more interpretable measure _dollars per day_.  Using current U.S. dollars as a unit, a person surviving on an income of less than $2 a day, is defined to be living in _absolute poverty_. We add this variable to the data table:

```{r}
gapminder <- gapminder %>%  
  mutate(dollars_per_day = gdp/population/365)
```

The GDP values are adjusted for inflation and represent current U.S. dollar, so these values are meant to be comparable across the years. Of course, these are country averages and within each country there is much variability. All the graphs and insights described below relate to country averages and not to individuals.

Here is a histogram of per day incomes from 1970:

```{r dollars-per-day-distribution}
past_year <- 1970
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(dollars_per_day)) + 
  geom_histogram(binwidth = 1, color = "black")
```

We use the `color = "black"` argument to draw a boundary and clearly distinguish the bins.

In this plot, we see that for the majority of countries, averages are below \$10 a day. However, the majority of the x-axis is dedicated to the `r filter(gapminder, year == past_year & !is.na(gdp)) %>% summarise(x = sum(dollars_per_day>10)) %>% pull(x)` countries with averages above \$10. So the plot is not very informative about countries with values below \$10 a day.

It might be more informative to quickly be able to see how many countries have average daily incomes of about $1 (extremely poor), \$2 (very poor), \$4 (poor), \$8 (middle), \$16 (well off), \$32 (rich), \$64 (very rich) per day. These changes are multiplicative and log transformations convert multiplicative changes into additive ones: when using base 2, a doubling of a value turns into an increase by 1. 


```{r, solution=TRUE, text = "This provides a _close-up_ of the mid to lower income countries."}
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(log2(dollars_per_day))) + 
  geom_histogram(binwidth = 1, color = "black")
``` 

  8) Make a histogram of `log2(dollars_per_day)` from 1970. 

<div class = "box">

#### Infobox - Which base? {-}

In the case above, we used base 2 in the log transformations. Other common choices are base $\mathrm{e}$ (the natural log) and base 10.

In general, we do not recommend using the natural log for data exploration and visualization. This is because while $2^2, 2^3, 2^4, \dots$ or $10^2, 10^3, \dots$ are easy to compute in our heads, the same is not true for  $\mathrm{e}^2, \mathrm{e}^3, \dots$, so the scale is not intuitive or easy to interpret.

In the dollars per day example, we used base 2 instead of base 10 because the resulting range is easier to interpret. The range of the values being plotted is `r with(filter(gapminder, year==past_year), range(dollars_per_day, na.rm=TRUE))`. 

In base 10, this turns into a range that includes very few integers: just 0 and 1. 
With base two, our range includes -2, -1, 0, 1, 2, 3, 4, and 5. It is easier to compute $2^x$ and $10^x$ when $x$ is an integer and between -10 and 10, so we prefer to have smaller integers in the scale. Another consequence of a limited range is that choosing the binwidth is more challenging. With log base 2, we know that a binwidth of 1 will translate to a bin with range $x$ to $2x$.

For an example in which base 10 makes more sense, consider population sizes. A log base 10 is preferable since the range for these is:

```{r}
filter(gapminder, year == past_year) %>%
  summarize(min = min(population), max = max(population))
```

Here is the histogram of the transformed values:

```{r population-histogram-log10}
gapminder %>% 
  filter(year == past_year) %>%
  ggplot(aes(log10(population))) +
  geom_histogram(binwidth = 0.5, color = "black")
```

In the above, we quickly see that country populations range between ten thousand and ten billion.
</div>

There are two ways we can use log transformations in plots. We can log the values before plotting them or use log scales on the axes. Both approaches are useful and have different strengths. If we log the data, we can more easily interpret intermediate values in the scale. For example, if we see:  

`----1----x----2--------3----`

for log transformed data, we know that the value of $x$ is about 1.5. If the scales are logged:

`----1----x----10------100---`

then, to determine `x`, we need to compute $10^{1.5}$, which is not easy to do in our heads. The advantage of using logged scales is that we see the original values on the axes. However, the advantage of showing logged scales is that the original values are displayed in the plot, which are easier to interpret. For example, we would see "32 dollars a day" instead of "5 log base 2 dollars a day".


```{r, solution=TRUE, text = "The plot from Q8 is the same except the values on the x-axis."}
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(dollars_per_day)) + 
  geom_histogram(binwidth = 1, color = "black") +
  scale_x_continuous(trans = "log2")
``` 

  9) Make a histogram of `dollars_per_day` from 1970 using a log2 scale on the x-axis. Compare it to the plot from Question 8. _Hint: you can use the `scale_x_continuous` function with `trans = "log2"`._

The histograms in Questions 8 and 9 have two _bumps_: one at about 4 and another at about 32. In statistics these bumps are sometimes referred to as _modes_. The mode of a distribution is the value with the highest frequency. The mode of the normal distribution is the average. When a distribution, like the one above, does not monotonically decrease from the mode, we call the locations where it goes up and down again _local modes_ and say that the distribution has _multiple modes_ indicating different distributions for different groups.

The histogram above suggests that the 1970 country income distribution has two modes: one at about 2 dollars per day (1 in the log 2 scale) and another at about 32 dollars per day (5 in the log 2 scale). However, the histogram does not show us if the two groups of countries are _west_ versus the _rest_. Let us create the `group` column:

```{r}
gapminder <- gapminder %>% 
  mutate(group = case_when(
    region %in% c("Western Europe", "Northern Europe","Southern Europe", 
                  "Northern America", "Australia and New Zealand") ~ "West",
    TRUE ~ "Rest")) %>% 
  as_tibble()
```



```{r, solution=TRUE, text = "The plot confirms the west vs the rest dichotomy."}
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(dollars_per_day)) + 
  geom_histogram(binwidth = 1, color = "black") +
  facet_grid(cols = vars(group)) +
  scale_x_continuous(trans = "log2")
``` 

```{r, hint=TRUE, eval=FALSE}
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(dollars_per_day)) + 
  geom_histogram(binwidth = 1, color = "black") +
  facet_grid(cols = ___) +
  scale_x_continuous(trans = "log2")
``` 

  10) Make a histogram of `dollars_per_day` from 1970 using a log2 scale and _facet_ it by group. Is there a _west_ versus the _rest_ dichotomy?
  
The exploratory data analysis above has revealed two characteristics about average income distribution in 1970. Using a histogram, we found a bimodal distribution with the modes relating to poor and rich countries. We will try to visualize these summaries in one plot. 


```{r, solution=TRUE}
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(group, dollars_per_day)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log2") + 
  geom_point()
``` 

```{r, hint=TRUE, eval=FALSE}
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(group, dollars_per_day)) +
  geom____() +
  scale_y_continuous(trans = "log2") + 
  geom____()
``` 

  11) Make a boxplot (`geom_boxplot`) of `dollars_per_day` (y-axis) versus `group` (x-axis) from 1970 using a log2 scale. Also add a the data using `geom_point()`.
  

Data exploration clearly shows that in 1970 there was a "west versus the rest" dichotomy. But does this dichotomy persist? We first have to be a little careful here since there are more countries represented in 2010 than in 1970: the total counts are larger. One reason for this is that several countries were founded after 1970. For example, the Soviet Union divided into several countries during the 1990s. Another reason is that data was available for more countries in 2010. Hence we only have to consider the countries with data available for both years:

```{r}
past_year <- 1970
present_year <- 2010
years <- c(past_year, present_year)
country_list_1 <- gapminder %>% 
  filter(year == past_year & !is.na(dollars_per_day)) %>% 
  pull(country)

country_list_2 <- gapminder %>% 
  filter(year == present_year & !is.na(dollars_per_day)) %>% 
  pull(country)
      
country_list <- intersect(country_list_1, country_list_2)
```

We can now filter the rows by `years` and `country_list`.



```{r, solution=TRUE, text = "The income gap between rich and poor countries has narrowed considerably during the last 40 years"}
gapminder %>% 
  filter(year %in% years & country %in% country_list) %>%
  ggplot(aes(dollars_per_day)) +
  geom_histogram(binwidth = 1, color = "black") +
  scale_x_continuous(trans = "log2") + 
  facet_grid(year ~ group)
``` 

```{r, hint=TRUE, eval=FALSE}
gapminder %>% 
  filter(year %in% ___ & country %in% ___) %>%
  ggplot(aes(dollars_per_day)) +
  geom_histogram(binwidth = 1, color = "black") +
  scale_x_continuous(trans = "log2") + 
  facet_grid(___)
``` 

  12) Make a histogram of `dollars_per_day` from 1970 and 2010 using a log2 scale and _facet_ it by group and year. Does the dichotomy persist? 
  

```{r, solution=TRUE, text = "We now see that the rich countries have become a bit richer, but percentage-wise, the poor countries appear to have improved more. In particular, we see that the proportion of _developing_ countries earning more than $16 a day increased substantially."}
gapminder %>% 
  filter(year %in% years & country %in% country_list) %>%
  mutate(year = factor(year)) %>%
  ggplot(aes(group, dollars_per_day, fill = year)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log2") 
``` 

```{r, hint=TRUE, eval=FALSE}
gapminder %>% 
  filter(year %in% years & country %in% country_list) %>%
  mutate(year = factor(___)) %>%
  ggplot(aes(group, dollars_per_day, fill = ___)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log2") 
``` 

  13) Make a boxplot of `dollars_per_day` versus `group` from 1970 and 2010 using a log2 scale. Use year as fill aesthetic. 
  
      _Hint: you must convert year to a factor using `mutate(year = factor(year))`._



The previous data exploration suggested that the income gap between rich and poor countries has narrowed considerably during the last 40 years. 
We used a series of histograms and boxplots to see this. Let us now shift to density plots. Let us start by noting that density plots for income distribution in `r past_year` and `r present_year` deliver the message that the gap is closing:

```{r income-smooth-density-by-year}
gapminder %>% 
  filter(year %in% years & country %in% country_list) %>%
  ggplot(aes(dollars_per_day)) +
  geom_density(fill = "grey") + 
  scale_x_continuous(trans = "log2") + 
  facet_grid(cols = vars(year))
```

In the `r past_year` plot, we see two clear modes: poor and rich countries. In `r present_year`, it appears that some of the poor countries have shifted towards the right, closing the gap. 

```{r, echo=FALSE, include=FALSE}
tmp <- gapminder %>% 
  filter(year == past_year & country %in% country_list) %>%
  mutate(group = ifelse(group == "West", "West", "Developing")) %>% 
  group_by(group) %>% 
  summarize(n=n()) %>%
  spread(group, n)
```


The next message we need to convey is that the reason for this change in distribution is that several poor countries became richer, rather than some rich countries becoming poorer. To do this, we can assign a color to the groups we identified during data exploration. 

```{r income-smooth-density-by-year-west-v-developing}
gapminder %>% 
  filter(year %in% years & country %in% country_list) %>%
  ggplot(aes(dollars_per_day, fill = group)) +
  scale_x_continuous(trans = "log2") +
  geom_density(alpha = 0.2) + 
  facet_grid(cols = vars(year))
```

Note the default is to have the area represented by each distribution add up to 1, regardless of the size of each group: the number of countries in the 'west' group is `r tmp[1,2]` and in the 'rest' group is `r tmp[1,1]`. We may use count on the y-axis instead:


```{r income-smooth-density-counts}
p <- gapminder %>% 
  filter(year %in% years & country %in% country_list) %>%
  ggplot(aes(dollars_per_day, y = ..count.., fill = group)) +
  scale_x_continuous(trans = "log2", limit = c(0.125, 300)) +
  facet_grid(cols = vars(year))
p + geom_density(alpha = 0.2) 
```


```{r, solution=TRUE, text = "This plot now shows that the developing world distribution is changing."}
p + geom_density(alpha = 0.2, bw = 0.75)
``` 

  14) To get densities smoother, use `bw = 0.75` argument so that the same bandwidth is used in each density. Comment on the plot.


As a final point, we note that in these distributions the weight of every country is the same. So if most of the population is improving, but living in a very large country, such as China, we might not appreciate this. We can actually weight the smooth densities using the `weight` mapping argument. We modify the dataset:

```{r}
gapminder <- gapminder %>% 
  filter(year %in% years & country %in% country_list) %>%
  group_by(year) %>%
  mutate(weight = population/sum(population)*2) %>%
  ungroup() 
```



```{r, solution=TRUE, text = "We now see that the rich countries have become a bit richer, but percentage-wise, the poor countries appear to have improved more. In particular, we see that the proportion of _developing_ countries earning more than $16 a day increased substantially."}
gapminder %>% 
  ggplot(aes(dollars_per_day, fill = group, weight = weight)) +
  scale_x_continuous(trans = "log2", limit = c(0.125, 300)) + 
  geom_density(alpha = 0.2, bw = 0.75) + 
  facet_grid(cols = vars(year))
``` 

```{r, hint=TRUE, eval=FALSE}
gapminder %>% 
  ggplot(aes(dollars_per_day, fill = group, weight = ___)) +
  scale_x_continuous(trans = "log2", limit = c(0.125, 300)) + 
  geom_density(alpha = 0.2, bw = 0.75) + 
  facet_grid(cols = vars(year))
``` 

  15) Modify the ggplot function with a weight argument and plot the density (with area equal 1). 




### Exercise (profit) {#ex-profit2}

Use the *exercise R Markdown template* to solve this exercise (**File > New File > R Markdown...**, select **From template** and then **TFA Exercise**).

Consider the dataset profit (provided by the tfa package) containing quarterly financial records for each costumer, product, etc.:

```{r}
# remotes::install_github("bss-osca/tfa/tfa-package", upgrade = FALSE)   # upgrade first
library(tfa)
library(skimr)
glimpse(profit)
skim(profit)
```

  1) Make a barplot that shows the total profitability of the product lines. Use the following steps:

```{r, solution=TRUE}
profit <- profit %>%
  mutate(across(where(is.character), as.factor))
``` 
```{r, hint=TRUE, eval=FALSE}
profit <- profit %>%
  mutate(across(where(is.___), as.___))
``` 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
a) Convert all character columns to factor columns.


```{r, solution=TRUE}
profit %>% 
  group_by(`Product Line`) %>% 
  summarise(Profit = sum(Profit)) %>% 
  ggplot(aes(x = `Product Line`, y = Profit)) +
  geom_col()
``` 
```{r, hint=TRUE, eval=FALSE}
profit %>% 
  group_by(___) %>% 
  summarise(Profit = sum(___)) %>% 
  ggplot(aes(x = ___, y = ___)) +
  geom_col()
``` 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
b) Group by product line, calculate the total profit and plot profit for each product line.


```{r, solution=TRUE}
profit %>% 
  group_by(`Product Line`) %>% 
  summarise(Profit = sum(Profit)) %>% 
  ggplot(aes(x = reorder(`Product Line`, Profit), y = Profit)) +
  geom_col()

# Alternatively you can reorder the data frame before calling ggplot
profit %>% 
  group_by(`Product Line`) %>% 
  summarise(Profit = sum(Profit)) %>% 
  arrange(Profit) %>% 
  mutate(`Product Line` = factor(`Product Line`, levels = `Product Line`, ordered = TRUE)) %>% 
  ggplot(aes(x = reorder(`Product Line`, Profit), y = Profit)) +
  geom_col() + 
  labs(title = "Total profit for each product line") 
``` 
```{r, hint=TRUE, eval=FALSE, title="Hint 2"}
profit %>% 
  group_by(`Product Line`) %>% 
  summarise(Profit = sum(Profit)) %>% 
  ggplot(aes(x = reorder(___), y = Profit)) +
  geom_col()
``` 
```{r, hint=TRUE, eval=FALSE, title="Hint 1", text="See the last section on this [webpage](https://sebastiansauer.github.io/ordering-bars/)."}
``` 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
c) Plot profit for each product line where product line is reordered based on total profit.


```{r, solution=TRUE}
profit %>% 
  group_by(`Product Line`) %>% 
  summarise(Profit = sum(Profit)) %>% 
  ggplot(aes(x = reorder(`Product Line`, Profit), y = Profit)) +
  geom_col() + 
  labs(title = "Total profit for each product line") 
``` 
```{r, hint=TRUE, eval=FALSE, text="Try to google 'ggplot add title'."}
```
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
d) Add a title to the plot using `labs`.


```{r, solution=TRUE}
profit %>% 
  group_by(`Product Line`) %>% 
  summarise(Profit = sum(Profit)) %>% 
  ggplot(aes(x = reorder(`Product Line`, Profit), y = Profit)) +
  geom_col() +
  labs(title = "Total profit for each product line") + 
  xlab("Product line") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
```{r, hint=TRUE, eval=FALSE, text="Try to google 'ggplot2 rotate axis labels'."}
```
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
e) Rotate the x-axis labels 90 degrees.


```{r, solution=TRUE}
dat <- profit %>% 
  group_by(`Product Line`) %>% 
  summarise(Profit = sum(Profit)) 
dat %>% slice_min(Profit)
dat %>% slice_max(Profit)
```
```{r, hint=TRUE, eval=FALSE}
dat <- profit %>% 
  group_by(`Product Line`) %>% 
  summarise(Profit = sum(Profit)) 
dat %>% slice_min(___)
dat %>% ___
```
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
f) Which product line is best and worst?
 


```{r, solution=TRUE, text="Some product lines have quite different earnings in different quarters."}
profit %>% 
  group_by(`Product Line`, Quarter) %>% 
  summarise(Profit = sum(Profit)) %>% 
  ggplot(aes(x = reorder(`Product Line`, Profit), y = Profit)) +
  geom_col() +
  facet_grid(cols = vars(Quarter)) +
  labs(title = "Total profit for each product line") +  
  xlab("Product Line") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
```{r, hint=TRUE, eval=FALSE}
profit %>% 
  group_by(`Product Line`, ___) %>% 
  summarise(Profit = sum(Profit)) %>% 
  ggplot(aes(x = reorder(`Product Line`, Profit), y = Profit)) +
  geom_col() +
  facet_grid(cols = vars(___)) +
  labs(title = "Total profit for each product line") +  
  xlab("Product Line") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
  2) Make a barplot that shows the total profitability of the product lines in each quarter. Are there details we have missed in Question 1?



```{r, solution=TRUE, text="The profit varies more for three of the product lines."}
profit %>% 
  ggplot(aes(y = Profit, x = `Product Line`)) +
  geom_boxplot() +
  labs(title = "Profit for each product line") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
```{r, hint=TRUE, eval=FALSE}
profit %>% 
  ggplot(aes(y = ___, x = ___)) +
  geom_boxplot() +
  labs(title = "Profit for each product line") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
  3) Make a boxplot of profitability of the product lines. Any insight?



```{r, solution=TRUE, text="Lowest and highest total profit is for PBI and WEM, respectively."}
profit %>% 
  group_by(`Customer ID`) %>% 
  summarise(Profit = sum(Profit)) %>% 
  ggplot(aes(x = reorder(`Customer ID`, Profit), y = Profit)) +
  geom_col() +
  labs(title = "Total profit for each customer") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
```{r, hint=TRUE, eval=FALSE}
profit %>% 
  group_by(___) %>% 
  summarise(Profit = sum(___)) %>% 
  ggplot(aes(x = reorder(___, ___), y = Profit)) +
  geom_col() +
  labs(title = "Total profit for each customer") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
  4) Make a barplot that shows the total profitability of the customers. Which customer is best and worst?



```{r, solution=TRUE, text="Since the number of transactions are not the same, the order of customers will not be the same."}
profit %>% 
  group_by(`Customer ID`) %>% 
  summarise(Profit = mean(Profit)) %>% 
  ggplot(aes(x = reorder(`Customer ID`, Profit), y = Profit)) +
  geom_col() +
  labs(title = "Mean profit for each customer") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
```{r, hint=TRUE, eval=FALSE}
profit %>% 
  group_by(`Customer ID`) %>% 
  summarise(Profit = ___) %>% 
  ggplot(aes(x = reorder(`Customer ID`, Profit), y = Profit)) +
  geom_col() +
  labs(title = "Mean profit for each customer") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
  5) Make a barplot that shows the mean profitability of the customers. Which customer is best and worst? Compare against Question 4 and discuss.

  
  
```{r, solution=TRUE, eval=FALSE}
profit %>% 
  group_by(`Customer ID`) %>% 
  summarise(ctr = n(), `Total Profit` = sum(Profit)) %>% 
  ggplot(aes(x = reorder(`Customer ID`, `Total Profit`), y = ctr, fill = `Total Profit`)) +
  geom_col() +
  labs(title = "Number of transactions (rows) for each customer") + 
  xlab("Customer") + ylab("Transactions") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
```{r, hint=TRUE, eval=FALSE}
profit %>% 
  group_by(`Customer ID`) %>% 
  summarise(ctr = ___, `Total Profit` = sum(___)) %>% 
  ggplot(aes(x = reorder(___, `Total Profit`), y = ctr, fill = ___)) +
  geom_col() +
  labs(title = "Number of transactions (rows) for each customer") + 
  xlab("Customer") + ylab("Transactions") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
``` 
  6) Make a plot illustrating number of transactions for each customer. Use total profit as fill atheistic. 

  
  
```{r, solution=TRUE}
profit %>% 
  ggplot(aes(y = Profit, x = `Customer ID`)) +
  geom_boxplot() +
  labs(title = "Profit for each customer") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
```{r, hint=TRUE, eval=FALSE}
profit %>% 
  ggplot(aes(y = ___, x = ___)) +
  geom_boxplot() +
  labs(title = "Profit for each customer") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
  7) Make a boxplot illustrating the profit for each customer. 

  
  




### Exercise (COVID-19) {#ex-covid19}

Use the *exercise R Markdown template* to solve this exercise (**File > New File > R Markdown...**, select **From template** and then **TFA Exercise**).

Countries around the world are responding to an outbreak of respiratory illness caused by a novel corona virus, COVID-19. The outbreak first started in Wuhan, China, but cases have been identified in a growing number of other locations internationally, including the United States. In this report we explore how the trajectory of the cumulative deaths in a number of countries.

The data come from the [**coronavirus** package](https://github.com/RamiKrispin/coronavirus), which pulls data from the Johns Hopkins University Center for Systems Science and Engineering (JHU CCSE) Corona virus repository. The corona virus package provides a tidy format dataset of the 2019 Novel Corona virus COVID-19 (2019-nCoV) epidemic. The package is available on GitHub [here](https://github.com/RamiKrispin/coronavirus) and is updated daily.

First load the following packages:

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)   # package for handling dates
```

The data frame called `coronavirus` in the coronavirus package provides a daily summary of the Corona virus (COVID-19) cases by country. Each row in the data frame represents a country (or, where relevant, state/province). Note that the data provided in this package provides daily number of deaths, confirmed cases, and recovered cases. Since we just need the dataset we load it using `read_csv`:

```{r}
coronavirus <- read_csv(
  "https://raw.githubusercontent.com/RamiKrispin/coronavirus/master/csv/coronavirus.csv", 
  col_types = cols(
    date = col_date(format = ""),
    province = col_character(),
    country = col_character(),
    lat = col_double(),
    long = col_double(),
    type = col_character(),
    cases = col_double()
  )
)

```

We calculate the total number of cases per day, cumulative numbers and days since first record: 

```{r}
dat <- coronavirus %>%
  group_by(country, date, type) %>% 
  summarise(tot_cases = sum(cases)) %>%
  group_by(country, type) %>% 
  arrange(date) %>%
  mutate(cumulative_cases = cumsum(tot_cases)) %>%
  ungroup() %>%
  mutate(
    days_elapsed = as.numeric(date - min(date)),
    year = year(date)
  ) %>% print()
```



```{r, solution=TRUE}
dat %>% 
  group_by(date, type) %>% 
  summarise(tot_cases = sum(tot_cases)) %>% print() %>% 
  ggplot(aes(x = date, y = tot_cases)) +
  geom_col() + 
  facet_grid(rows = vars(type), scales = "free") +
  labs(
    title = "Number of Covid 19 cases per day",
    y = "cases"
  )
```
```{r, hint=TRUE, eval=FALSE, title="Hint 2"}
dat %>% 
  group_by(date, type) %>% 
  summarise(tot_cases = sum(tot_cases)) %>% print() %>% 
  ggplot(aes(x = ___, y = ___)) +
  geom_col() + 
  facet_grid(rows = vars(___), scales = "___") +
  labs(
    title = "___",
    y = "___"
  )
```
```{r, hint=TRUE, eval=FALSE, title="Hint 1", text="Note you must aggegrate the numbers for the countries."}
dat %>% 
  group_by(date, ___) %>% 
  summarise(tot_cases = sum(___)) 
```
  1) Calculate and plot the number of confirmed, death and recovered cases per day given date using `facet_grid` and `geom_col`.
  

Consider the following set of countries:

```{r}
countries <- c(
  "China",
  "France",
  "Denmark",
  "US",
  "Italy"
)
```




```{r, solution=TRUE}
dat %>% 
  filter(type == "death", country %in% countries) %>% 
  ggplot(aes(x = days_elapsed, 
             y = cumulative_cases, 
             color = country)) +
  geom_line() +
  theme(legend.position = "bottom") +
  labs(
    x = str_c("Days since ", min(dat$date)),
    y = "Cumulative number of deaths",
    title = "Cumulative deaths from COVID-19, selected countries"
  )
```
```{r, hint=TRUE, eval=FALSE, title="Hint 2"}
dat %>% 
  filter(type == "death", country %in% countries) %>% 
  ggplot(aes(x = ___, 
             y = ___, 
             color = ___)) +
  geom_line() +
  theme(legend.position = "bottom") +
  labs(
    x = str_c("Days since ", min(dat$date)),
    y = "___",
    title = "Cumulative deaths from COVID-19, selected countries"
  )
```
```{r, hint=TRUE, eval=FALSE, title="Hint 1", text="First you have to filter type and country"}
dat %>% 
  filter(type == "___", country %in% ___) %>% 
```
2) Plot a lineplot of the cumulative number of deaths as a function of days elapsed for the selected countries. Use country as color aesthetic. 


Since the countries have different population sizes, we would like to calculate some numbers relative to the population size. First we need population sizes for each country. They are given in the dataset `world_pop` in the _tfa_ package:

```{r}
world_pop <- tfa::world_pop %>%
  filter(country %in% countries) %>% 
  print()
```


We can join the datasets using:

```{r}
dat <- dat %>% 
  filter(country %in% countries) %>% 
  left_join(world_pop) %>% 
  print()
any(is.na(dat))  # check if any missing values
```



```{r, solution=TRUE}
dat <- dat %>% 
  mutate(tot_cases_pop = 100000 * tot_cases/pop)
```
3) Calculate `tot_cases_pop` as number of cases per 100000 inhabitants. That is, total cases divided by population and multiplied by 100000.



```{r, solution=TRUE, warning=FALSE}
dat %>% 
  filter(date >= today() - days(21), type == "confirmed") %>% 
  ggplot(aes(x = date, y = tot_cases_pop, fill = country)) +
  geom_col(position = position_dodge2())

```
```{r, hint=TRUE, eval=FALSE, title="Hint 2"}
dat %>% 
  filter(date >= today() - days(21), type == ___) %>% 
  ggplot(aes(x = date, y = ___, fill = ___)) +
  geom_col(position = position_dodge2())
```
```{r, solution=TRUE, title="Hint 1", warning=FALSE}
# Use this to find date 21 days ago
today() - days(21)
```
4) Plot the number of confirmed cases per 100000 inhabitants for the last 21 days. Use country as fill aesthetic.



```{r, solution=TRUE, warning=FALSE}
dat %>% 
  filter(date >= today() - days(14), country == "Denmark", type == "confirmed") %>% 
  ggplot(aes(x = date, y = tot_cases_pop)) +
  geom_col()
```
```{r, hint=TRUE, eval=FALSE}
dat %>% 
  filter(date >= ___, country == ___, type == ___) %>% 
  ggplot(aes(x = date, y = tot_cases_pop)) +
  geom_col()
```
5) Plot the number of confirmed cases per 100000 inhabitants in Denmark for the last 14 days.



### Exercise (Lego and sales) {#ex-lego}

Use the *exercise R Markdown template* to solve this exercise (**File > New File > R Markdown...**, select **From template** and then **TFA Exercise**).

Consider (simulated) data of Lego sales in 2018 for a sample of customers who bought Legos in the U.S. The dataset is called `lego_sales`. You can find descriptions of each of the variables in the help file for the dataset, which you can access by running `?lego_sales` in your Console.

You need the _tidyverse_ package as usual and the *dsbox* package for the data.

```{r}
library(tidyverse)
library(dsbox)   # install using devtools::install_github("rstudio-education/dsbox")
```

Answer the following questions using a table with numbers and try to visualize it. For each question, state your answer in a sentence, e.g. "The first three common names of purchasers are ...".

1) What are the three most common first names of purchasers?

2) What are the three most common themes of Lego sets purchased?

3) Among the most common theme of Lego sets purchased, what is the most common subtheme?




```{r, hint=TRUE, eval=FALSE, text="Use the `case_when()` function."}
```

4) Create a new variable called `age_group` and group the ages into the 
following categories: "18 and under", "19 - 25", "26 - 35", "36 - 50", "51 and over".



```{r, hint=TRUE, eval=FALSE, text="You may need to consider quantity of purchases."}
```

5) Which age group has purchased the highest number of Lego sets. 



```{r, hint=TRUE, eval=FALSE, text="*Hint:* You will need to consider quantity of purchases as well as price of Lego sets."}
```

6) Which age group has spent the most money on Legos?

7) Come up with a question you want to answer using these data, and write it down. 
Then, create a data visualization that answers the question, and explain how 
your visualization answers the question.



