---
output: html_document
editor_options: 
  chunk_output_type: console
---

# (PART) Appendix {-}

# Hints for working in groups {-}

[GitHub] is a code sharing and publishing service and may be seen as a social networking site for programmers. We are going to use GitHub for sharing R code in the R part of the course in the groups. 

<!-- https://happygitwithr.com/big-picture.html -->
Git is a version control system. Git manages the evolution of a set of files – called a repository – in a sane, highly structured way. If you have no idea what I’m talking about, think of it as the “Track Changes” features from Microsoft Word on steroids. 

GitHub provide a home for your Git-based projects on the internet. If you have no idea what I’m talking about, think of it as DropBox but much, much better. It allows other people to see your stuff, sync up with you, and perhaps even make changes. Even for private solo projects, it’s a good idea to push your work to a remote location for peace of mind.

We are going to Git and GitHub for maintaining the R code you produce in the groups.

<div class = "boxTask">
`r addIcon("tasks", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`  


* Install the desktop version of [RStudio][rstudio-download], a powerful user interface for R. 
</div>

# Course description {-}


# Using R in Excel and Excel in R {-}

https://bert-toolkit.com/

# Run a OPL model from R


# Learning goals {-}
							
The purpose of this course is to give students a knowledge about IT tools for Analytics which requires the analyst to be qualified in handling tools beyond e.g. basic Excel. 							
After having participated in the course, the student must, in addition to achieving general academic skills, demonstrate:			
							
Knowledge of	

  1.	how a computer works at a basic level.						
  2.	basic programming such as variables, arrays, loops, functions and procedures. 						
  3.	what an algorithm is. 						
  4.	how to implement an algorithm based on a description. 						
  5.	different programming languages. 						
  6.	how to manage a code in a collaborative working environment.						
							
Skills to 		

  7.	handle data such as import, tidy, transform, visualize and export. 						
  8.	develop well-structured code. 						
  9.	perform testing and debugging. 						
  10.	implement/code selected algorithms. 						
  11.	apply analytical techniques on data.						
  12.	apply relevant methods, algorithms and techniques from this course in order to solve a specific problem.						
							
Competences to		

  13.	independently handle data given a problem. 						
  14.	independently analyse data given a relevant research question.						
  15.	compare different programming languages.						
  16.	compare different algorithms solving a problem and discuss their advantages and disadvantages.						
  17.	interpret and discuss results based on a data analysis in relation to the relevant academic literature.						
  18.	communicate results from applied research in a scientific way, e.g. using literate programming.						


# Schedule {-} 

```{r, echo=FALSE, message=FALSE, cache=FALSE}
source("common.R")
library(googlesheets4)
sheets_deauth()
sheet <- read_sheet("https://docs.google.com/spreadsheets/d/1-fSG3KWip0Vm5Uwf9T6vUMuH0ReBhpITHT3rl896vwc/edit?usp=sharing",
           sheet = "Schedule (for students)")
knitr::kable(sheet)
```


# Business Analytics {- #sec:ba}

Business Analytics (BA) refers to the scientific process of transforming data into insight for making better decisions in business. BA can both be seen as the complete decision making process for solving a business problem or as a set of methodologies that enable the creation of business value. As a process it can be characterized by descriptive, predictive, and prescriptive model building using ‘big’ data sources.

Descriptive Analytics: A set of technologies and processes that use data to understand and analyze business performance. Descriptive analytics are the most commonly used and most well understood type of analytics. Descriptive analytics categorizes, characterizes, consolidates, and classifies data. Examples are standard reporting and dashboards (KPIs, what happened or is happening now?) and ad-hoc reporting (how many/often?). Descriptive analytics often serves as a first step in the successful application of predictive or prescriptive analytics.

Predictive Analytics: The use of data and statistical techniques to make predictions about future outputs/outcomes, identify patterns or opportunities for business performance. Examples of techniques are data mining (what data is correlated with other data?), pattern recognition and alerts (when should I take action to correct/adjust a spare part?), Monte-Carlo simulation (what could happen?), neural networks (which customer group are best?) and forecasting (what if these trends continue?).

Prescriptive Analytics: The use of optimization and other decision modelling techniques using the results of descriptive and predictive analytics to suggest decision options with the goal of improving business performance. Prescriptive analytics attempt to quantify the effect of future decisions in order to advise on possible outcomes before the decisions are actually made. Prescriptive analytics predicts not only what will happen, but also why it will happen and provides recommendations regarding actions that will take advantage of the predictions. Prescriptive analytics are relatively complex to administer, and most companies are not yet using it in their daily course of business. However, when implemented correctly, it can have a huge impact on business performance and how businesses make decisions. Examples on prescriptive analytics are optimization in production planning and scheduling, inventory management, the supply chain and transportation planning. 

Companies who use BA focus on fact-based management to drive decision making and treats data and information as a strategic asset that is shared within the company. This enterprise approach generates a companywide respect for applying descriptive, predictive and prescriptive analytics in areas such as supply chain, marketing and human resources. 

Related areas: In the past Business Intelligence traditionally focuses on querying, reporting, online analytical processing, i.e. descriptive analytics. However, a more modern definition of Business Intelligence is the union of descriptive and predictive analytics. Operations Research or Management Science deals with the application of advanced analytical methods to help make better decisions and can hence be seen as prescriptive analytics. However, traditionally it has been taking a more theoretical approach and focusing on problem-driven research while BA takes a more data-driven approach. Logistics is a cross-functional area focusing on the effective and efficient flows of goods and services, and the related flows of information and cash. Supply Chain Management adds a process-oriented and cross-company perspective. Both can be seen as prescriptive analytics with a more problem-driven research focus. Advanced Analytics is often used as a classification of both predictive and prescriptive analytics. Data science is an interdisciplinary field about scientific methods, processes, and systems to extract knowledge or insights from data in various forms, either structured or unstructured and can be seen as Business analytics applied to a wider range of data. 

Resources
http://analytics-magazine.org/the-analytics-journey/ 
https://en.wikipedia.org/wiki/Business_analytics 
http://connect.informs.org/analytics/home 
https://www.or-exchange.org/questions/5645/informs-analytics-definition 
https://en.wikipedia.org/wiki/Prescriptive_analytics
https://en.wikipedia.org/wiki/Predictive_analytics 

Business Analytics (BA) refers to the scientific process of transforming data into insight for making better decisions in business. BA can both be seen as the complete decision making process for solving a business problem and the creation of business value by integration of concepts, methods and data. As a process, it can be characterized by descriptive, predictive, and prescriptive model building using ‘big’ data sources. For further details, see http://analytics-magazine.org/the-analytics-journey/.
 
Operations Research and Management Science deal with the application of advanced analytical methods in order to make better decisions. Hence, they are closely related to prescriptive analytics. Companies that use BA focus on fact-based management to drive decision making and treats data and information as a strategic asset that is shared within the company. To a larger extent than before, data may be available for validation purposes.





# License {-}

This work is licensed under a [Creative Commons Attribution-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-sa/4.0/).

<center>
<i class="fab fa-creative-commons fa-2x"></i><i class="fab fa-creative-commons-by fa-2x"></i><i class="fab fa-creative-commons-sa fa-2x"></i>
</center>

*This is a human-readable summary of (and not a substitute for) the license.
Please see <https://creativecommons.org/licenses/by-sa/4.0/legalcode> for the full legal text.*

**You are free to:**

- **Share**---copy and redistribute the material in any medium or
  format

- **Remix**---remix, transform, and build upon the material for any
  purpose, even commercially.

The licensor cannot revoke these freedoms as long as you follow the
license terms.

**Under the following terms:**

- **Attribution**---You must give appropriate credit, provide a link
  to the license, and indicate if changes were made. You may do so in
  any reasonable manner, but not in any way that suggests the licensor
  endorses you or your use.
  
- **ShareAlike**---If you remix, transform, or build upon the material, you must distribute your contributions under the same license as the original. 

- **No additional restrictions**---You may not apply legal terms or
  technological measures that legally restrict others from doing
  anything the license permits.

**Notices:**

You do not have to comply with the license for elements of the
material in the public domain or where your use is permitted by an
applicable exception or limitation.

No warranties are given. The license may not give you all of the
permissions necessary for your intended use. For example, other rights
such as publicity, privacy, or moral rights may limit how you use the
material.



## Colophon 

This book was written in [bookdown](http://bookdown.org/) inside [RStudio](http://www.rstudio.com/ide/). 
<!-- The website [stat545.com](https://stat545.com) is hosted with [Netlify](https://www.netlify.com), and automatically updated after every commit by [Travis-CI](https://travis-ci.org). The complete source is available from [GitHub](https://github.com/rstudio-education/stat545). -->

<!-- The [Tools for Analytics logo](assets/logo.png) and the book style was designed by [Desirée De Leon](https://desiree.rbind.io/). -->

This version of the book was built with:

```{r message = FALSE, warning = FALSE, echo = FALSE}
# needed because new_session is set to true in _bookdown.yml
all_pkgs <- renv::dependencies(path = "DESCRIPTION") %>% 
  pull(Package) 

session <- devtools::session_info(pkgs = all_pkgs)

session$platform
```

Along with these packages:

```{r message = FALSE, warning = FALSE, echo = FALSE}
session$packages %>% 
  select(package, loadedversion, date, source) %>% 
  DT::datatable(rownames = FALSE,
                class = 'cell-border stripe',
                filter = list(position = 'top'),
                options = list(pageLength = nrow(session$packages), 
                           autoWidth = TRUE,
                           bInfo = FALSE,
                           paging = FALSE))
```



```{r include = FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

