```{r include = FALSE, cache=FALSE}
source("common.R")
```


# Functions


[Intro]

## Learning outcomes

By the end of this module, you are expected to be able to:

* Call a function.
* Formulate a function with different input arguments.
* Describe why functions are important in R.
* Set defaults for input arguments. 
* Return values from functions.
* Explain how variable scope and precedence works.
* Document functions. 


data type comparison with VBA

<!-- SOLO increasing: identify · memorise · name · do simple procedure · collect data · -->
<!-- enumerate · describe · interpret · formulate · list · paraphrase · combine · do -->
<!-- algorithms · compare · contrast · explain causes · analyse · relate · derive · -->
<!-- evaluate · apply · argue · theorise · generalise · hypothesise · solve · reflect -->


## DataCamp course

An excellent introduction to functions is given in [Chapter 3][datacamp-r-intermediate-chap3] in DataCamp course [Intermediate R][datacamp-intermediate-r]. Please complete the chapter before continuing. 


 or 





## Different learning paths

`r strLPath`

* For a more detailed interactive tutorial about functions see the DataCamp course [Introduction to Writing Functions in R][datacamp-r-functions].
* Roger Peng has a video about [Writing functions in R][video-functions] giving examples in RStudio. 
* A detailed tutorial on functions is given in Chapters [18][stat-545-functions-part1], [19][stat-545-functions-part2] and [20][stat-545-functions-part3] of @stat545.
* An introduction to functions is also given in [Chapter 14](https://bookdown.org/rdpeng/rprogdatascience/functions.html) of @rdpeng.


## Exercises

### Exercise ()

### Exercise

3\. The function `nchar` tells you how many characters long a character vector is. Write a line of code that assigns to the object `new_names` the state abbreviation when the state name is longer than 8 characters.
    
### Exercise

4\. Create a function `sum_n` that for any given value, say $n$, computes the sum of the integers from 1 to n (inclusive). Use the function to determine the sum of integers from 1 to 5,000.

### Exercise

5\. Create a function `altman_plot` that takes two arguments, `x` and `y`, and plots the difference against the sum.
  
### Exercise

6\. After running the code below, what is the value of `x`?

```{r, eval=FALSE}
x <- 3
my_func <- function(y){
  x <- 5
  y+5
}
```

### Exercise

7\. Write a function `compute_s_n` that for any given $n$ computes the sum $S_n = 1^2 + 2^2 + 3^2 + \dots n^2$. Report the value of the sum when $n=10$.

### Exercise 

8\. Define an empty numerical vector `s_n` of size 25 using `s_n <- vector("numeric", 25)` and store in the results of $S_1, S_2, \dots S_{25}$ using a for-loop.

### Exercise

9\. Repeat exercise 8, but this time use `sapply`.

### Exercise

### Exercise
10\. Repeat exercise 8, but this time use `map_dbl`.

### Exercise
11\. Plot $S_n$ versus $n$. Use points defined by $n=1,\dots,25$.

### Exercise
12\. Confirm that the formula for this sum is $S_n= n(n+1)(2n+1)/6$.

<!-- <div class = "boxExercise"> -->
<!-- `r addIcon("laptop-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`   -->

<!-- **Exercise 1** -->
<!-- asdf asdlfkj sdfklj fsdjka æ -->

<!--    1. sdf -->
<!--    2. edasf -->

<!-- **Exercise 2** -->
<!-- asdfav sdaf sdhj sdjkahdf jkhasdkjæsdfh asjf -->

<!--    1. sdf -->
<!--    2. sdf -->
<!-- </div> -->


<!-- `r addIcon("file-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("external-link-alt", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->








