
```{r, include = FALSE}
source("common.R")
```

# Install R and RStudio {#install}

<!-- 
Content from: https://stat545.com/block000_r-rstudio-install.html
-->

## Learning outcomes

By the end of this module, you are expected to have:

* Installed R.
* Installed RStudio.
<!-- * Installed Git. -->
* Tested R and RStudio. <!--  and Git. -->
* Installed some packages.

## Install R and RStudio

[R] is a programming language and free software environment. The R language is widely used among statisticians and data miners for developing statistical software and data analysis. For a further overview and description of the history of R see [Chapter 2](https://bookdown.org/rdpeng/rprogdatascience/history-and-overview-of-r.html) in @Peng2018. To run R you need to install it on your computer. Moreover, you need the IDE (integrated development environment) [RStudio], to save your work.



<div class = "boxTask">
`r addIcon("tasks", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`  

* Install R from [CRAN][cran] (Comprehensive R Archive Network). Install the latest precompiled binary distribution for your operating system (use the links up at the top of the CRAN page)!
* Install the desktop version of [RStudio][rstudio-download], a powerful user interface for R. 
* Install the latest version of [Git][git-install]. If you have problems, see [here][hg-install-git].
</div>
    
If you have a pre-existing installation of R and/or RStudio, reinstall both to the latest versions. It can be considerably harder to run old software than new.

<!-- * If you upgrade R, you will need to update any packages you have installed. The command below should get you started, though you may need to specify more arguments if, e.g., you have been using a non-default library for your packages. -->

<!--     ``` r -->
<!--     update.packages(ask = FALSE, checkBuilt = TRUE) -->
<!--     ``` -->

<!--   __Note:__ this will only look for updates on CRAN. So if you use a package that lives *only* on GitHub or if you want a development version from GitHub, you will need to  update manually, e.g. via `devtools::install_github()`. -->

## Test your installation

<div class = "boxTask">`r addIcon("tasks", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`  

* Do whatever is appropriate for your OS to launch RStudio. You should get a window similar to the screenshot you see [here][rstudio-workbench], but yours will be more boring because you haven't written any code or made any figures yet!

* Put your cursor in the pane labeled Console, which is where you interact with the live R process. Create a simple object with code like `x <- 2 * 4` (followed by enter or return). Then inspect the `x` object by typing `x` followed by enter or return. You should see the value 8 print to screen. If yes, you've succeeded in installing R and RStudio.

* Try to open a new file **File > New File > New RMarkdown...**. Use the defaults and press **OK**. Next save the file and compile it using **Knit** (Ctrl+Shift+K). You have now compiled a document with R code inside of it. 

* Close and restart R. Go to **Tools > Global Options > Git/SVN **. The 'Git executable' field should point to the path where Git is installed. On windows the path looks like `C:/Program Files/Git/bin/git.exe` and on macOS and Linux, the path usually looks something like `/usr/bin/git`. If the field is not filled with a path, have a look at [Chapter 13](https://happygitwithr.com/rstudio-see-git.html) and [14](https://happygitwithr.com/troubleshooting.html) at the [Happy Git][happy-git] site.
</div>

## Add-on packages

R is an extensible system and many people share useful code they have developed as a _package_ via CRAN and GitHub. To install a package from CRAN, for example the [dplyr][dplyr-cran] package for data manipulation, one way to do it in the R console.

```r
install.packages("dplyr", dependencies = TRUE)
```

By including `dependencies = TRUE`, we are being explicit and extra-careful to install any additional packages the target package, `dplyr` in the example above, needs to have around. 

<div class = "boxTask">
`r addIcon("tasks", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`  

* Install the package [tidyverse][tidyverse-main-page] which is in fact a bundle of packages by running (note this operation may take a long time):

```r
install.packages("tidyverse", dependencies = TRUE)
```

* Check if you have successfully installed tidyverse by loading the package:

```r
library(tidyverse)
```

* If your install was unsuccessful try to install the packages who fails one at a time.
</div>


## Different learning paths

`r strLPath`

* Suggestion 1 
* Suggestion 2


## Recap

* (Add stuff)



<!-- ## Further resources -->

<!-- The above will get your basic setup ready but here are some links if you are interested in reading a bit further. -->

<!-- * [How to Use RStudio][rstudio-support] -->
<!-- * [RStudio's leads for learning R][rstudio-R-help] -->
<!-- * [R FAQ][cran-faq] -->
<!-- * [R Installation and Administration][cran-R-admin] -->
<!-- * [More about add-on packages in the R Installation and Administration Manual][cran-add-ons] -->
    

```{r links, child="links.md"}
```
