---
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```


<!-- Exercises from: https://rafalab.github.io/dsbook/programming-basics.html -->

# (PART) Programming in R {-} 


# Loops and conditionals {#sec-loops}

This module consider programming with loops and conditional statements.


## Learning outcomes {#lo-loops}

By the end of this module, you are expected to be able to:

* Formulate conditional statements.
* Use functions `any`, and `all`.
* Formulate loops in R using for and while statements.
* Use function `if_else`.

The learning outcomes relate to the [overall learning goals](#lg-course) number 2, 4 and 10 of the course.

<!-- SOLO increasing: identify · memorise · name · do simple procedure · collect data · -->
<!-- enumerate · describe · interpret · formulate · list · paraphrase · combine · do -->
<!-- algorithms · compare · contrast · explain causes · analyse · relate · derive · -->
<!-- evaluate · apply · argue · theorise · generalise · hypothesise · solve · reflect -->


## Conditionals and control flow

An excellent introduction to conditionals and if statements is given in [Chapter 1][datacamp-r-intermediate-chap1] of the interactive DataCamp course [Intermediate R][datacamp-r-intermediate]. Please complete the chapter before continuing. 

Some functions are also useful for comparing logical data types. Consider example: 

```{r}
x <- c(1, 3, 5, 10, 2, 17, 11, NA, 4)
x > 10  # are the elements greater that 10
any(x > 10)  # are any of the elements greater that 10
all(x > 10)  # are all of the elements greater that 10
all(x < 20)  # are all of the elements greater that 20
all(x < 20, na.rm = TRUE)  # are all of the elements greater that 20
```
That is, functions `any` and `all` can be used to join logical values in vectors. 

Some if statements can be written alternatively using function `if_else`:

```r
if_else(condition, true, false, missing = NULL)
```

For example:

```{r}
x <- c(-5:5, NA)
x

## using if and for
res <- rep("", length(x))
for (i in seq_along(x)) {
  if (is.na(x[i])) res[i] <- "missing"
  else if (x[i] < 0) res[i] <- "negative"
  else res[i] <- "positive"
}
res

## implicit if statement
res <- rep("", length(x))
res
res[x < 0] <- "negative"
res[x >= 0] <- "positive"
res[is.na(x)] <- "missing"
res

## using if_else
res <- if_else(x < 0, "negative", "positive", "missing")
res
```




## Loops 

An excellent introduction to conditionals and if statements is given in [Chapter 2][datacamp-r-intermediate-chap2] of the interactive DataCamp course [Intermediate R][datacamp-r-intermediate]. Please complete the chapter before continuing. 

For loops in R may be slow. However, not if you follow some golden rules:

* Don’t use a loop when a vectorized alternative exists
* Don’t grow objects (via `c`, `cbind`, etc) during the loop - R has to create a new object and copy across the information just to add a new element or row/column. Instead, allocate an object to hold the results and fill it in during the loop.

As an example, consider the for loop with 4 iterations:

```{r}
i_val <- c(1,2,6,9)
res <- rep(NA,4)
res
for (idx in 1:length(i_val)) {
  res[idx] <- 6 * i_val[idx] + 9
}
res
```

Note we allocate memory for the result vector before the loop so we don't have to grow the result object. Next, we calculate results $6i+9$ using a loop. Be careful here! This is not the same:

```{r}
res <- rep(NA,4)
for (i in i_val) {
  res[i] <- 6 * i + 9
}
res
```

In this example we however can use a vetorized alternative:

```{r}
res <- 6 * i_val + 9
res
```

where the operation is applied to each element in the vector.

Nested for loops are also possible. A simple example of a nested loop:

```{r}
for (i in 1:3) {
  for (j in 1:2) {
    cat(str_c("i =", i, " j = ",j, "\n"))
  }
}
```

We here uses the function `cat` to print out a string (`\n` indicate new line). Note how the nested loops are executed:

* Set `i = 1` (outer loop)
   - Set `j = 1` (inner loop), `i` stays 1
   - Set `j = 2` (inner loop), `i` stays 1
   - Inner loop finishes, proceed with outer loop.
* Increase `i = 2` (outer loop)
   - Set `j = 1` (inner loop), `i` stays 2
   - Set `j = 2` (inner loop), `i` stays 2
   - Inner loop finishes, proceed with outer loop.
* Increase `i = 3` (outer loop)
   - Set `j = 1` (inner loop), `i` stays 3
   - Set `j = 2` (inner loop), `i` stays 3
   - Inner loop finishes, proceed with outer loop.
* Outer loop finishes as well (we looped over `i in 1:3`). Job done.

Nested loops can be used to iterate over matrices or data frames:

```{r}
mat <- matrix(NA, nrow = 2, ncol = 3)
mat
for (i in 1:nrow(mat)) {
  for (j in 1:ncol(mat)) {
     mat[i,j] <- (i-1)*ncol(mat) + j
     cat(str_c("Entry (", i, ", ", j, ") = ", mat[i,j], "\n"))
  }
}
mat
```

Often you can replace nested loops with a single loop by using `expand_grid`:

```{r}
mat <- matrix(NA, nrow = 2, ncol = 3)
ite <- expand_grid(i = 1:2, j=1:3)
ite
for (r in 1:nrow(ite)) { # iterate over rows
   i <- ite$i[r]
   j <- ite$j[r]
   mat[i,j] <- (i-1)*ncol(mat) + j
   cat(str_c("Entry (", i, ", ", j, ") = ", mat[i,j], "\n"))
}
mat
```

Note `expand_grid` creates a data frame with all combinations. This way of looping is a more flexible approach since you can

* nest more loops by adding more columns to `ite`,
* add different values in each column.

For instance, if you only want to calculate values for row 2 and columns 1 and 3 the code becomes:

```{r}
mat <- matrix(NA, nrow = 2, ncol = 3)
ite <- expand_grid(i = 2, j = c(1,3))
ite
for (r in 1:nrow(ite)) { # iterate over rows
   i <- ite$i[r]
   j <- ite$j[r]
   mat[i,j] <- (i-1)*ncol(mat) + j
   cat(str_c("Entry (", i, ", ", j, ") = ", mat[i,j], "\n"))
}
mat
```



## Different learning paths {#lp-loops}

`r strLPath`

* Roger Peng has a video [An introduction to control structures][video-control-structures] including `if`, `for` and `while`.
* Hefin Rhys has a video [Conditional statements and loops in R][video-conditional-loops] doing examples in RStudio.
* For a more detailed introduction to loops see [Chapter 21][r4ds-iteration] in @r4ds.
* Loops can also be created using the *apply* family of functions. An introduction is given in [Chapter 4][datacamp-r-intermediate-chap4] of the interactive DataCamp course [Intermediate R][datacamp-r-intermediate].


## Recap {#rc-loops}

*Comparison/relational operators* known to R are:

- `<` for less than,
- `>` for greater than,
- `<=` for less than or equal to,
- `>=` for greater than or equal to,
- `==` for equal to each other (and not `=` which is typically used for assignment!),
- `!=` not equal to each other.
   
*Logical operators* known to R are:

- `&` and,
- `|` or,
- `!` not.

If you use `&&` and `||` only the first element in vectors are compared. In general this is used rarely. 

Useful functions that return a logical are `any` and `all` which can be used to join logical values in vectors.

Conditional Statements can be constructed using for instance `if` and `while` statements. Moreover, function `if_else` is a vetorized alternative.

*Loops* can be created using `for` and `while` statements. 

You can break out of a look using `break` and jump to the next iteration (skipping the remainder of the code in the loop) using `next`.

Don’t use a loop when a vectorized alternative exists.

Don’t grow objects during the loop. Instead, allocate an object to hold the results and fill it in during the loop.

Nested loops are possible in R. However, often they can be converted into a single loop by defining a data frame having the values of the nested loops in each row. Here function `expand_grid` may be useful to create the data frame. 



## Exercises {#ex-loops}

`r strExercises`

### Exercise (conditional expressions)

*Solve this exercise using a script file*

1) Consider object `x`:

   ```{r}
   x <- c(1,2,-3,4)
   ```
  
   What will this conditional expression return?

   ```{r, eval=FALSE}
   if(all(x>0)){
    print("All Postives")
   } else {
    print("Not all positives")
   }
   ```

2) What will the following expressions return?

   ```{r, eval=FALSE}
   x <- c(TRUE, FALSE, TRUE, TRUE)
   all(x)
   any(x)
   any(!x)
   all(!x)
   ```

3) Which of the expressions above is always `FALSE` when at least one entry of a logical vector `x` is TRUE?

Consider vector:

```{r}
x <- 1:15
x
```

```{r, echo=FALSE}
addSolution('if_else(x < 7, as.integer(0), x)')
```

4) Use the `if_else` function set elements with value below 7 to 0.

```{r, echo=FALSE}
addSolution('if_else(x < 7 | x > 10, NA_integer_, x)')
```

5) Use the `if_else` function set elements with value below 7 or above 10 to `NA_integer_`.

6) Consider code

   ```{r}
   x <- sample(c(1:10,NA,5.5), 1)
   x
   ```

   which generate a natural number from the vector `c(1:10,NA,5.5)`. 

   ```{r, echo=FALSE}
   addSolution('
   x <- sample(c(1:10,NA,5.5), 1)
   if (is.na(x)) {
     y <- "missing"
   } else if (x %% 2 == 0) {
     y <- "even"
   } else if (x %% 2 == 1) {
     y <- "odd"
   } else if (x %% 1 > 0) {
     y <- "decimal"
   } 
   x
   y
   ')
   ```

   Write code which set object `y` equal to "even" if `x` is even, "odd" if `x` is odd, "decimal" if `x` has a decimal not zero and "missing" if `x` is `NA`. Hint: have a look at `?'%%'` (the modulo operator).





### Exercise (loops)

```{r, echo=FALSE}
addSolution('x <- rep(NA,4)
for (i in 1:4) {
  x[i] <- 2 * i + 4
}
x')
```
1) Using a `for` loop, create a vector having values $2i + 4$ given $i=1\ldots 4$.

```{r, echo=FALSE}
addSolution('
i_val <- c(2, 5, 6, 12)
x <- rep(NA, length(i_val))
for (idx in 1:length(i_val)) {
  x[idx] <- 2 * i_val[idx] + 4
}
')
```
2) Using a `for` loop, create a vector having values $2i + 4$ given $i=2,5,6,12$.

```{r, echo=FALSE}
addSolution('
i_val <- c(2, 5, 6, 12)
x <- rep(NA, length(i_val))
idx <- 1
while (idx < 5) {
  x[idx] <- 2 * i_val[idx] + 4
  idx <- idx + 1
}
')
```
3) Solve Question 2 using a `while` loop.

```{r, echo=FALSE}
addSolution('
2 * 1:4 + 4             # Q1
2* c(2, 5, 6, 12) + 4   # Q2
')
```
4) Solve Question 1 and 2 using a vectorized alternative. 

```{r, echo=FALSE}
addSolution('
2 * 1:4 + 4             # Q1
2* c(2, 5, 6, 12) + 4   # Q2
')
```
5) Solve Question 1 and 2 using a vectorized alternative. 


### Exercise (calculating distances) {#ex-dist}

Consider zip codes in Jutland:

```{r}
library(tfa)
zips
```

We want to calculate distances between between a subset of zip areas:

```{r}
idx <- 1:5
dat <- zips[idx,]
dat
distanceMat <- matrix(NA, nrow = length(idx), ncol = length(idx))
colnames(distanceMat) <- str_c(dat$Zip[idx], dat$Area[idx], sep = " ") 
rownames(distanceMat) <- colnames(distanceMat)
distanceMat
```

We can find average distances between two zip codes (here rows 1 and 2 in `dat`) using Bing maps:

```{r}
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
             dat$Zip[1], ",Denmark",
             "&wp.1=",
             dat$Zip[2], ",Denmark",
             "&avoid=minimizeTolls&key=", key)
library(jsonlite)
lst <- jsonlite::fromJSON(url)
dist <- lst$resourceSets$resources[[1]]$travelDistance
dist
lst$statusCode
lst$statusDescription
```

Note we call the Bing maps API with the two zip codes. A [json file](https://en.wikipedia.org/wiki/JSON) is returned and stored in a list. To get the average travel distance we access `travelDistance`. The status code should be 200 if the calculation returned is okay.

```{r, echo=FALSE}
addSolution('
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
for(i in 1:nrow(distanceMat)) {
   for(j in 1:ncol(distanceMat)) {
      if (i>j) {distanceMat[i,j] <- distanceMat[j,i]; next}   # assume symetric distances
      if (!is.na(distanceMat[i,j])) next   # value already calculated 
      if (i==j) {distanceMat[i,j] <- 0; next}
      
      url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
                   dat$Zip[i], ",Denmark",
                   "&wp.1=",
                   dat$Zip[j], ",Denmark",
                   "&avoid=minimizeTolls&key=", key)
      lst <- jsonlite::fromJSON(url)
      if (lst$statusCode == 200) {
         distanceMat[i,j] <- lst$resourceSets$resources[[1]]$travelDistance
      }
   }
}
distanceMat
')
```

Use nested for loops to fill `distanceMat` with distances. Assume that the distance from a to b is the same as from b to a. That is, you only have to call the API once for two zip codes. Use an if statement to check if the status code is okay.


### Exercise (`expand_grid`)

```{r, echo=FALSE}
addSolution('
ite <- expand_grid(i = c(1,5), j = 2:3)
ite
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
for (r in 1:nrow(ite)) { # iterate over rows
   i <- ite$i[r]
   j <- ite$j[r]
   if (i==j) {distanceMat[i,j] <- 0; next}
   url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
                dat$Zip[i], ",Denmark",
                "&wp.1=",
                dat$Zip[j], ",Denmark",
                "&avoid=minimizeTolls&key=", key)
   lst <- jsonlite::fromJSON(url)
   if (lst$statusCode == 200) {
      distanceMat[i,j] <- lst$resourceSets$resources[[1]]$travelDistance
      distanceMat[j,i] <- distanceMat[i,j]
   }
}
distanceMat
')
```

Consider the solution of Exercise \@ref(ex-dist) and assume that you only want to calculate the distance from rows 1 and 5 to rows 2 and 3 in `dat`. Modify the solution using `expand_grid` so only one loop is used.

```{r links, child="links.md"}
```














<!-- <div class = "boxExercise"> -->
<!-- `r addIcon("laptop-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`   -->

<!-- **Exercise 1** -->
<!-- asdf asdlfkj sdfklj fsdjka æ -->

<!--    1. sdf -->
<!--    2. edasf -->

<!-- **Exercise 2** -->
<!-- asdfav sdaf sdhj sdjkahdf jkhasdkjæsdfh asjf -->

<!--    1. sdf -->
<!--    2. sdf -->
<!-- </div> -->


<!-- `r addIcon("file-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("external-link-alt", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->




