---
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```


<!-- Exercises from: https://rafalab.github.io/dsbook/programming-basics.html -->

# (PART) Programming in R {-} 


# Loops and conditionals

This module consider programming with loops and conditional statements.


## Learning outcomes

By the end of this module, you are expected to be able to:

* Formulate conditional statements.
* Use functions `any`, and `all`.
* Formulate loops in R using for and while statements.
* Use function `if_else`.

<!-- SOLO increasing: identify · memorise · name · do simple procedure · collect data · -->
<!-- enumerate · describe · interpret · formulate · list · paraphrase · combine · do -->
<!-- algorithms · compare · contrast · explain causes · analyse · relate · derive · -->
<!-- evaluate · apply · argue · theorise · generalise · hypothesise · solve · reflect -->


## Conditionals and control flow

An excellent introduction to conditionals and if statements is given in [Chapter 1][datacamp-r-intermediate-chap1] of the interactive DataCamp course [Intermediate R][datacamp-r-intermediate]. Please complete the chapter before continuing. 

Some functions are also useful for comparing logical data types. Consider example: 

```{r}
x <- c(1, 3, 5, 10, 2, 17, 11, NA, 4)
x > 10  # are the elements greater that 10
any(x > 10)  # are any of the elements greater that 10
all(x > 10)  # are all of the elements greater that 10
all(x < 20)  # are all of the elements greater that 20
all(x < 20, na.rm = TRUE)  # are all of the elements greater that 20
```
That is, functions `any` and `all` can be used to join logical values in vectors. 

Some if statements can be written alternatively using function `if_else`:

```r
if_else(condition, true, false, missing = NULL)
```

For example:

```{r}
x <- c(-5:5, NA)
x

## using if and for
res <- rep("", length(x))
for (i in seq_along(x)) {
  if (is.na(x[i])) res[i] <- "missing"
  else if (x[i] < 0) res[i] <- "negative"
  else res[i] <- "positive"
}
res

## implicit if statement
res <- rep("", length(x))
res
res[x < 0] <- "negative"
res[x >= 0] <- "positive"
res[is.na(x)] <- "missing"
res

## using if_else
res <- if_else(x < 0, "negative", "positive", "missing")
res
```




## Loops 

An excellent introduction to conditionals and if statements is given in [Chapter 2][datacamp-r-intermediate-chap2] of the interactive DataCamp course [Intermediate R][datacamp-r-intermediate]. Please complete the chapter before continuing. 

For loops in R may be slow. However, not if you follow some golden rules:

* Don’t use a loop when a vectorized alternative exists
* Don’t grow objects (via `c`, `cbind`, etc) during the loop - R has to create a new object and copy across the information just to add a new element or row/column. Instead, allocate an object to hold the results and fill it in during the loop.

As an example, we’ll create a new version of analyze that will return the mean inflammation per day (column) of each file.


## Different learning paths

`r strLPath`

* Roger Peng has a lot of [videos](https://www.youtube.com/playlist?list=PLjTlxb-wKvXPqyY3FZDO8GqIaWuEDy-Od) on YouTube about R. Of special interest are
   
   - [An introduction to control structures][video-control-structures] including `if`, `for` and `while`.
   
* For a more detailed introduction to loops see [Chapter 21][r4ds-iteration] in @r4ds.


## Recap

*Comparison/relational operators* known to R are:

- `<` for less than,
- `>` for greater than,
- `<=` for less than or equal to,
- `>=` for greater than or equal to,
- `==` for equal to each other (and not `=` which is typically used for assignment!),
- `!=` not equal to each other.
   
*Logical operators* known to R are:

- `&` and,
- `|` or,
- `!` not.

If you use `&&` and `||` only the first element in vectors are compared. In general this is used rarely. 

Useful functions that return a logical are `any` and `all` which can be used to join logical values in vectors.

Conditional Statements can be constructed using for instance `if` and `while` statements. 




## Exercises

### Exercise (conditional expressions)

*Solve this exercise using a script file*

1) Consider object `x`:

   ```{r}
   x <- c(1,2,-3,4)
   ```
  
   What will this conditional expression return?

   ```{r, eval=FALSE}
   if(all(x>0)){
    print("All Postives")
   } else {
    print("Not all positives")
   }
   ```

2) What will the following expressions return?

   ```{r, eval=FALSE}
   x <- c(TRUE, FALSE, TRUE, TRUE)
   all(x)
   any(x)
   any(!x)
   all(!x)
   ```

3) Which of the expressions above is always `FALSE` when at least one entry of a logical vector `x` is TRUE?

Consider vector:

```{r}
x <- 1:15
x
```

```{r, echo=FALSE}
addSolution('if_else(x < 7, as.integer(0), x)')
```

4) Use the `if_else` function set elements with value below 7 to 0.

```{r, echo=FALSE}
addSolution('if_else(x < 7 | x > 10, NA_integer_, x)')
```

5) Use the `if_else` function set elements with value below 7 or above 10 to `NA_integer_`.

6) Consider code

   ```{r}
   x <- sample(c(1:10,NA,5.5), 1)
   x
   ```

   which generate a natural number from the vector `c(1:10,NA,5.5)`. 

   ```{r, echo=FALSE}
   addSolution('
   x <- sample(c(1:10,NA,5.5), 1)
   if (is.na(x)) {
     y <- "missing"
   } else if (x %% 2 == 0) {
     y <- "even"
   } else if (x %% 2 == 1) {
     y <- "odd"
   } else if (x %% 1 > 0) {
     y <- "decimal"
   } 
   x
   y
   ')
   ```

   Write code which set object `y` equal to "even" if `x` is even, "odd" if `x` is odd, "decimal" if `x` has a decimal not zero and "missing" if `x` is `NA`. Hint: have a look at `?'%%'` (the modulo operator).





### Exercise (loops)

```{r, echo=FALSE}
addSolution('x <- rep(NA,4)
for (i in 1:4) {
  x[i] <- 2 * i + 4
}
x')
```
1) Using a `for` loop, create a vector having values $2i + 4$ given $i=1\ldots 4$.

```{r, echo=FALSE}
addSolution('
i_val <- c(2, 5, 6, 12)
x <- rep(NA, length(i_val))
for (idx in 1:length(i_val)) {
  x[idx] <- 2 * i_val[idx] + 4
}
')
```
2) Using a `for` loop, create a vector having values $2i + 4$ given $i=2,5,6,12$.

```{r, echo=FALSE}
addSolution('
i_val <- c(2, 5, 6, 12)
x <- rep(NA, length(i_val))
idx <- 1
while (idx < 5) {
  x[idx] <- 2 * i_val[idx] + 4
  idx <- idx + 1
}
')
```
3) Solve Question 2 using a `while` loop.

```{r, echo=FALSE}
addSolution('
2 * 1:4 + 4             # Q1
2* c(2, 5, 6, 12) + 4   # Q2
')
```
4) Solve Question 1 and 2 using a vectorized alternative. 

```{r, echo=FALSE}
addSolution('
2 * 1:4 + 4             # Q1
2* c(2, 5, 6, 12) + 4   # Q2
')
```
5) Solve Question 1 and 2 using a vectorized alternative. 


### Exercise (calculating distances)

Consider zip codes in Jutland:

```{r}
library(tfa)
zips
```

We want to calculate distances between between a subset of zip areas:

```{r}
idx <- 1:5
dat <- zips[idx,]
dat
distanceMat <- matrix(NA, nrow = length(idx), ncol = length(idx))
colnames(distanceMat) <- str_c(dat$Zip[idx], dat$Area[idx], sep = " ") 
rownames(distanceMat) <- colnames(distanceMat)
distanceMat
```

We can find average distances between two zip codes (here rows 1 and 2 in `dat`) using Bing maps:

```{r}
key <- "AlUJdApmvPe8y2_IMrC4j4x8fzytbD2M0SvlmpemL09ae_CWS6-IuNSgrAtXoyeP"
url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
             dat$Zip[1], ",Denmark",
             "&wp.1=",
             dat$Zip[2], ",Denmark",
             "&avoid=minimizeTolls&key=", key)
library(jsonlite)
lst <- jsonlite::fromJSON(url)
dist <- lst$resourceSets$resources[[1]]$travelDistance
dist
lst$statusCode
lst$statusDescription
```

Note we call the Bing maps API with the two zip codes. A json files is returned and stored in a list. To get the average travel distance we access `travelDistance`. The status code should be 200 if the calculation returned is okay.

```{r, echo=FALSE}
addSolution('
for(i in 1:nrow(distanceMat)) {
   for(j in 1:ncol(distanceMat)) {
      if (i>j) {distanceMat[i,j] <- distanceMat[j,i]; next}   # assume symetric distances
      if (!is.na(distanceMat[i,j])) next   # value already calculated 
      if (i==j) {distanceMat[i,j] <- 0; next}
      
      url <- str_c("http://dev.virtualearth.net/REST/V1/Routes/Driving?wp.0=",
                   dat$Zip[i], ",Denmark",
                   "&wp.1=",
                   dat$Zip[j], ",Denmark",
                   "&avoid=minimizeTolls&key=", key)
      lst <- jsonlite::fromJSON(url)
      if (lst$statusCode == 200) {
         distanceMat[i,j] <- lst$resourceSets$resources[[1]]$travelDistance
      }
   }
}
distanceMat
')
```

Use nested for loops to fill `distanceMat` with distances. Assume that the distance from a to b is the same as from b to a. That is, you only have to call the API once for two zip codes. Use an if statement to check if the status code is okay.





















<!-- <div class = "boxExercise"> -->
<!-- `r addIcon("laptop-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`   -->

<!-- **Exercise 1** -->
<!-- asdf asdlfkj sdfklj fsdjka æ -->

<!--    1. sdf -->
<!--    2. edasf -->

<!-- **Exercise 2** -->
<!-- asdfav sdaf sdhj sdjkahdf jkhasdkjæsdfh asjf -->

<!--    1. sdf -->
<!--    2. sdf -->
<!-- </div> -->


<!-- `r addIcon("file-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("external-link-alt", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->




