```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```


<!-- Exercises from: https://rafalab.github.io/dsbook/programming-basics.html -->

# (PART) Programming in R {-} 


# Loops and conditionals

[Intro]


## Learning outcomes

By the end of this module, you are expected to:

* Know how to formulate conditional statements.
* Know how to formulate loops in R using for and while statements.

* Know how `any` and `all` 
* Know how to make loops in R using for and while statements.

* `if_else`
* which

<!-- SOLO increasing: identify · memorise · name · do simple procedure · collect data · -->
<!-- enumerate · describe · interpret · formulate · list · paraphrase · combine · do -->
<!-- algorithms · compare · contrast · explain causes · analyse · relate · derive · -->
<!-- evaluate · apply · argue · theorise · generalise · hypothesise · solve · reflect -->

data type comparison with VBA


## Conditionals and control flow

An excellent introduction to conditionals and if statements is given in Chapter 1 of the interactive DataCamp course [Intermediate R][intermediate-r]. Please complete the chapter before continuing. 



## Loops 

An excellent introduction to conditionals and if statements is given in Chapter 2 of the interactive DataCamp course [Intermediate R][intermediate-r]. Please complete the chapter before continuing. 



<!-- ### Tasks -->

<!-- ### Online Tasks/activitites -->

<!-- <div class = "boxTask"> -->
<!-- `r addIcon("tasks", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`   -->

<!--    * dasf -->
<!--    * fsa -->
<!--    * 324 -->
<!-- </div> -->

## Different learning paths

`r strLPath`

* Suggestion 1 
* Suggestion 2

## Exercises

### Exercise

1\. What will this conditional expression return?

```{r, eval=FALSE}
x <- c(1,2,-3,4)

if(all(x>0)){
  print("All Postives")
} else{
  print("Not all positives")
}
```

### Exercise

2\. Which of the following expressions is always `FALSE` when at least one entry of a logical vector `x` is TRUE?

a. `all(x)`
b. `any(x)`
c. `any(!x)`
d. `all(!x)`

### Exercise

3\. The function `nchar` tells you how many characters long a character vector is. Write a line of code that assigns to the object `new_names` the state abbreviation when the state name is longer than 8 characters.
    
### Exercise

4\. Create a function `sum_n` that for any given value, say $n$, computes the sum of the integers from 1 to n (inclusive). Use the function to determine the sum of integers from 1 to 5,000.

### Exercise

5\. Create a function `altman_plot` that takes two arguments, `x` and `y`, and plots the difference against the sum.
  
### Exercise

6\. After running the code below, what is the value of `x`?

```{r, eval=FALSE}
x <- 3
my_func <- function(y){
  x <- 5
  y+5
}
```

### Exercise

7\. Write a function `compute_s_n` that for any given $n$ computes the sum $S_n = 1^2 + 2^2 + 3^2 + \dots n^2$. Report the value of the sum when $n=10$.

### Exercise 

8\. Define an empty numerical vector `s_n` of size 25 using `s_n <- vector("numeric", 25)` and store in the results of $S_1, S_2, \dots S_{25}$ using a for-loop.

### Exercise

9\. Repeat exercise 8, but this time use `sapply`.

### Exercise

### Exercise
10\. Repeat exercise 8, but this time use `map_dbl`.

### Exercise
11\. Plot $S_n$ versus $n$. Use points defined by $n=1,\dots,25$.

### Exercise
12\. Confirm that the formula for this sum is $S_n= n(n+1)(2n+1)/6$.





<!-- <div class = "boxExercise"> -->
<!-- `r addIcon("laptop-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))`   -->

<!-- **Exercise 1** -->
<!-- asdf asdlfkj sdfklj fsdjka æ -->

<!--    1. sdf -->
<!--    2. edasf -->

<!-- **Exercise 2** -->
<!-- asdfav sdaf sdhj sdjkahdf jkhasdkjæsdfh asjf -->

<!--    1. sdf -->
<!--    2. sdf -->
<!-- </div> -->


<!-- `r addIcon("file-code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("code", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->
<!-- `r addIcon("external-link-alt", attrib = list(class = "fa-pull-right fa-5x iconStyle1"))` -->

## Resources



