```{r, code = readLines("common.R"), cache = FALSE, include=FALSE}
```





# Transforming data {#sec-transform}

## Learning outcomes {#lo-transform}

By the end of this module, you are expected to be able to:

* Describe what tidy and wangling is.
* Apply the most common string functions
* Apply tidy operations to data 
* Transform data
* Clean data

<!-- SOLO increasing: identify · memorise · name · do simple procedure · collect data · -->
<!-- enumerate · describe · interpret · formulate · list · paraphrase · combine · do -->
<!-- algorithms · compare · contrast · explain causes · analyse · relate · derive · -->
<!-- evaluate · apply · argue · theorise · generalise · hypothesise · solve · reflect -->




<!-- ## Tidy  -->


<!-- ## Wangling -->


<!-- https://learn.datacamp.com/projects/758 -->

<!-- https://learn.datacamp.com/courses/dealing-with-missing-data-in-r -->

<!-- https://learn.datacamp.com/courses/working-with-data-in-the-tidyverse -->


<!-- https://learn.datacamp.com/courses/introduction-to-the-tidyverse -->


## Different learning paths {#lp-transform}

`r strLPath`

* sdfad



## Recap {rc-transform}





## Exercises {#ex-transform}

`r strExercises`




```{r links, child="links.md"}
```