---
title: "`r params$title`"
author: "`r params$author`"
date: "`r params$date`"
output: 
   html_document:
      toc: false
   pdf_document:
      includes:
         in_header: "preamble.tex"
      # extra_dependencies:
      #    a4wide: null
      #    lmodern: null
      #    newtxtext: ["scaled=1", "helvratio=0.95"]
      #    newtxmath: null
fontsize: 11pt
classoption: a4paper
# bibliography: litt.bib  
params:
  doc_output: solution
  title: "TFA exam solution (R part)"
  author: "Lars Relund Nielsen"
  date: ""
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
.solution {
  background-color: aliceblue;
  padding: 10px;
}

h2 {
  font-size: large;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  out.width = "100%",
  message=FALSE)
options(width = 90, scipen=999)

library(tidyverse)
a_ctr <- 1  # assignment counter
ctrA <- function() {
  a_ctr <<- a_ctr + 1
  q_ctr <<- 1
  return(str_c("Assignment ", a_ctr - 1))
}

q_ctr <- 1  # question counter
ctrQ <- function(reset = FALSE) {
  if (reset) q_ctr <<- 1
  q_ctr <<- q_ctr + 1
  return(str_c("Question ", q_ctr - 1, ""))
}

library(conflicted)
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
conflict_prefer("lag", "dplyr", quiet = TRUE)
```

```{r main, include=FALSE}
doc_output <- params$doc_output

if (doc_output == "solution") {
  knitr::opts_hooks$set(solution = function(options) {
    if (isTRUE(options$solution)) {
      options$eval = TRUE
    }
    options
  })
  knitr::knit_hooks$set(solution = function(before, options, envir) {
    if (before) {
      paste0('\n<div class="solution">\nGuiding solution:\n\n')
    } else {
      paste0('\n</div>\n')
    }
  })
}

if (doc_output == "template") {
  knitr::opts_hooks$set(output = function(options) {
    if (!is.na(options$output)) {
      if (options$output == "solution") {
        options$code = "# Your code here"
      }
    }
    options
  })
}

if (doc_output == "exam") {
  knitr::opts_hooks$set(solution = function(options) {
    if (isTRUE(options$solution)) {
      options$eval = FALSE
      options$include = FALSE
    }
    options
  })
}
```

\includepdf[fitpaper=true, pagecommand={\pagestyle{fancy}}]{front.pdf}

 

# R - `r ctrA()` (14%)

## `r ctrQ()` 
 
Make functions:

  - `SecToMin` which takes an input argument `sec` in seconds and return the number converted to minutes.
  - `SecToHours` which takes an input argument `sec` in seconds and return the number converted to hours.
  - `MinToSec` which takes an input argument `min` in minutes and return the number converted to seconds.
  - `MinToHours` which takes an input argument `min` in minutes and return the number converted to hours.
  - `HoursToMin` which takes an input argument `hours` in hours and return the number converted to minutes.
  - `HoursToSec` which takes an input argument `hours` in hours and return the number converted to seconds.

All numbers may be decimal numbers, e.g. 90 seconds is 1.5 minutes and 1.5 hours is 90 minutes. 

```{r, solution=TRUE}
SecToMin <- function(sec) {
   return(sec / 60)
}

SecToHours <- function(sec) {
   return(sec / 60 / 60)
}

MinToSec <- function(min) {
   return(min * 60)
}

MinToHours <- function(min) {
   return(min / 60)
}

HoursToMin <- function(hours) {
   return(hours * 60)
}

HoursToSec <- function(hours) {
   return(hours * 60 * 60)
}
```




## `r ctrQ()` 

Make a function `ConvertTime` which takes two input arguments:

  - `val` A number.
  - `unit` A string that can take values "sec", "min" and "hours".

The function should return `val` converted to seconds, minutes and hours with features:

  - works for all possible values for `unit`,
  - uses the functions in Question 1,
  - returns a vector with 3 numbers (seconds, minutes and hours) or `NA` if `unit` does not equals "sec", "min" or "hours".



```{r, solution=TRUE}
ConvertTime <- function(val, unit) {
   if (unit == "sec") {
      return(c(sec = val, min = SecToMin(val), hours = SecToHours(val)))
   } 
   if (unit == "min") {
      return(c(sec = MinToSec(val), min = val, hours = MinToHours(val)))
   }
   if (unit == "hours") {
      return(c(sec = HoursToSec(val), min = HoursToMin(val), hours = val))
   }
   return(NA)
}
## We test the function:
ConvertTime(val = 1.5, unit = "min")
ConvertTime(val = 1.5, unit = "hours")
ConvertTime(val = 1.5, unit = "kr")
```




# R - `r ctrA()` (18%)

```{r get-data, eval=FALSE, include=FALSE, exam=TRUE}
library(tidyverse)
library(skimr)
library(jsonlite)

info <- function(tab_id) {
   url <- str_c("https://api.statbank.dk/v1/tableinfo/", tab_id, "?lang=en")
   lst <- read_json(url, simplifyVector = T) 
   return(list(description = lst$description, unit = lst$unit, variables = lst$variables[,1:2]))
}
info("VALGK3")

get_data <- function(tab_id, col_id = NULL) {
   url <- str_c("https://api.statbank.dk/v1/tableinfo/", tab_id, "?lang=en")
   lst <- read_json(url, simplifyVector = T) 
   cols <- lst$variables$id
   if (!is.null(col_id)) cols <- cols[col_id]
   url <- str_c("https://api.statbank.dk/v1/data/", tab_id, "/CSV?lang=en&", str_c(cols, collapse = "=*&"), "=*") %>% 
      URLencode()
   dat <- read_csv2(url)
   return(dat)
}
dat <- get_data("VALGK3")

dat <- dat %>% 
  rename(area = OMRÅDE, party = PARTI, voteType = STEMMER, year = TID, value = INDHOLD)
dat <- dat %>% 
  filter(area != "All Denmark", !str_detect(area, "Region"))
dat <- dat %>% 
  pivot_wider(names_from = voteType, values_from = value) %>% 
  rename(validVotes = `Valid votes`, personalVotes = `Personal votes (total)`, listedMen = `Listed men`, listedWomen = `Listed women`, electedMen = `Elected men`, electedWomen = `Elected women`)
write_csv(dat, "election.csv")
```

The dataset in the file `election.csv` lists election votes for municipal elections in Denmark. The column/variables are:

* `area`: municipality,
* `party`: political party,
* `year`: election year,
* `validVotes`: Number of (valid) votes, 
* `personalVotes`: total number of personal votes, 
* `listedMen`: men listed,
* `listedWomen`: women listed, 
* `electedMen`: elected men,
* `electedWomen`: elected women.

You can read the dataset file into the dataset `dat` using

```{r}
library(tidyverse)
dat <- read_csv("elections.csv")
```

Use the *dplyr* package in _tidyverse_ to calculate relevant summary tables (data frames) and answer the following questions.


## `r ctrQ()`

How many different municipalities were there in Denmark in 2017?

```{r, solution=TRUE}
res <- dat %>%
  filter(year == 2017) %>% 
  distinct(area) %>% 
  nrow() %>% 
  print()
```

<div class="solution">
We have a total of `r res` municipalities.
</div>



## `r ctrQ()`

In which election year was the total number of votes highest?

```{r, solution=TRUE}
res <- dat %>%
  group_by(year) %>% 
  summarize(votes = sum(validVotes)) %>% 
  arrange(desc(votes)) %>% 
  print()
```

<div class="solution">
The highest number of votes was in `r res$year[1]`.
</div>


## `r ctrQ()` 

Consider the 2017 election. How many votes and elected candidates did the different parties get in total (sort your result descending by votes)?

```{r, solution=TRUE}
res <- dat %>%
  filter(year == 2017) %>% 
  mutate(elected = electedMen + electedWomen) %>% 
  group_by(party) %>% 
  summarize(votes = sum(validVotes), elected = sum(elected)) %>% 
  arrange(desc(votes)) %>% 
  print()
```

<div class="solution">
The highest number of votes was given to the `r res$party[1]` (`r res$votes[1]` votes)  who got `r res$elected[1]` elected.
</div>


## `r ctrQ()`

Calculate the percentage of females listed for each year. Are there on average more women than men listed? Is there any tendency over the years?

```{r, solution=TRUE}
res <- dat %>%
  group_by(year) %>% 
  summarize(listedWomen = sum(listedWomen), listedMen = sum(listedMen)) %>% 
  mutate(listedWomenPct = listedWomen/(listedMen + listedWomen)) %>% 
  arrange(listedWomenPct) %>% 
  print()
```

<div class="solution">
The listed women is approx. `r round(mean(res$listedWomenPct)*100)`. That is, listed women is lower than listed men. This percentage seems to be almost constant over the years (but increasing very slowly).
</div>


## `r ctrQ()`

Consider the 2017 election. In which municipality did the Liberal Democratic Party get the highest percentage of votes? 

```{r, solution=TRUE}
res <- dat %>%
  filter(year == 2017) %>% 
  group_by(area) %>% 
  mutate(votesPct = validVotes/sum(validVotes)) %>% 
  filter(party == "Liberal Democratic Party") %>% 
  select(area, party, votesPct) %>% 
  arrange(desc(votesPct)) %>% 
  print()
```

<div class="solution">
The Liberal Democratic Party got must relative votes in `r res$area[1]`.
</div>





























# R - `r ctrA()` (18%)

Answer this assignment using the *ggplot2* package in *tidyverse* (you might need dplyr for preparing the datasets you want to plot). We work with the dataset from Assignment 2 which can be read using:

```{r}
library(tidyverse)
dat <- read_csv("elections.csv")
```


## `r ctrQ()`

Create a visualization showing the total number of votes for each election year with the following features:

* Number of votes is represented using columns.
* Columns are filled with colours for each party.
* Informative figure title and axis titles are given.

Which year had the lowest number of votes?

```{r, solution=TRUE}
dat %>%
  group_by(year, party) %>% 
  summarise(votes = sum(validVotes)) %>% 
  ggplot(aes(y = votes, x = factor(year), fill = party)) +
  geom_col() +
  labs(title = "Total number of votes",
       x = "Year",
       y = "Votes"
    )
```

<div class="solution">
The lowest number of votes was in 2009.
</div>

## `r ctrQ()`

Create a visualization showing the relative number of elected women in each municipality with the following features:

* The relative number of elected women is shown using columns.
* Municipalities are rotated 90 degrees on the x-axis.
* The columns are sorted increasing.
* Informative figure title and axis titles are given.

Which municipality has the lowest/highest percentage of women elected?

```{r, solution=TRUE}
dat %>%
  group_by(area) %>% 
  summarize(electedWomen = sum(electedWomen), electedMen = sum(electedMen)) %>% 
  mutate(electedWomenPct = electedWomen/(electedMen + electedWomen)) %>% 
  filter(!is.na(electedWomenPct)) %>% 
  ggplot(aes(x = reorder(area, electedWomenPct), y = electedWomenPct)) +
  geom_col() + 
  labs(title = "Relative number of elected women",
       x = "Municipality",
       y = "%"
    ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

<div class="solution">
The lowest number elected is in Ærø and the highest is in Gentofte.
</div>


## `r ctrQ()`

Create a visualization showing the elected number of candidates as a function of election year with the following features:

* We consider only municipalities Herning, Copenhagen and Aarhus.
* We consider only the Social Democratic Party, Conservative Peoples Party and Liberal Democratic Party.
* The plot is divided using `area` (facet).
* Different colours are used for each party.
* Informative figure title and axis titles are given.
* Data points are given in the plot.
* A line for each party is added.

Consider the Social Democratic Party in Copenhagen. Based on the plot do the elected number of candidates increase or decrease over the election years?

```{r, solution=TRUE}
dat %>% 
  filter(area %in% c("Herning", "Copenhagen", "Aarhus"), party %in% c("Social Democratic Party", "Conservative Peoples Party", "Liberal Democratic Party")) %>% 
  mutate(elected = electedMen + electedWomen) %>% 
  ggplot(aes(x = year, y = elected, color = party)) +
  geom_point() +
  geom_line() +
  scale_x_binned(n.breaks = 4, nice.breaks = TRUE) +
  facet_grid(. ~ area) +
  labs(title = "Elected persons",
       x = "Year",
       y = "Elected",
       color = "Party") +
  theme(legend.position = "bottom") 
```

<div class="solution">
The trend for the Social Democratic Party in Copenhagen is decreasing over the years.
</div>

## `r ctrQ()`

Create a visualization showing the relative number of women elected compared to listed (e.g. if listed is 12 and elected is 4 then the number is 4/12) with the following features:

* Columns are used for each party.
* Colours are used to identify the party.
* A sub-plot is given for each year.
* Informative figure title and axis titles are given.

Which party seems to get the highest number of the listed women elected? 

```{r, solution=TRUE}
dat %>%
  mutate(electedW = electedWomen/listedWomen) %>% 
  group_by(party, year) %>% 
  summarize(electedW = mean(electedW, na.rm = T)) %>% 
  ggplot(aes(x = reorder(party, electedW), y = electedW, fill = party)) +
  geom_col() +
  facet_grid(year ~ .) +
  labs(title = "Releative number of women elected compared to listed.",
       x = "Party",
       y = "Percentage",
       fill = "Party") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(), legend.position = "bottom")
```

<div class="solution">
For all years the Social Democratic Party has the highest number of the listed women elected.
</div>



## `r ctrQ()`

Create a visualization showing the variance in the relative number of personal votes compared to valid votes (i.e. divide the two numbers) with the following features:

* We consider only municipalities Herning, Copenhagen and Aarhus.
* Variation is shown using a box-plot for each municipality.
* Informative figure title and axis titles are given.

Which municipality has the highest median?

```{r, solution=TRUE}
dat %>%
  filter(area %in% c("Herning", "Copenhagen", "Aarhus")) %>% 
  mutate(personal = personalVotes/validVotes) %>% 
    ggplot(aes(x = area, y = personal, fill = area)) +
    geom_boxplot() +
    labs(title = "Releative number of personal votes over all the election years.",
       x = "Municipality",
       y = "Personal votes (pct)",
       fill = "Municipality") +
    theme(legend.position = "none")


```

<div class="solution">
Based on the plot Herning has the highest median of personal votes.
</div>





